### 2025-03-20 article_bihom.tex
first commit

% !TXS template
\documentclass[french]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[a4paper]{geometry}
\usepackage{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\begin{document}
Ce que j'ai compris dans l'article: Gröbner bases of bihomogeneous ideals generated by polynomials of bidegree (1, 1): Algorithms and complexity de Faugère et al.
\newline

\section{Définitions et notations}
\begin{itemize}
\item $I_{i-1}:f_i$ is spanned by $I_{i-1}$ and by the maximal minors of $\text{jac}_{\textbf{x}}(F_{i-1})$ (if $i > n_y + 1$) and $\text{jac}_{\textbf{y}}(F_{i-1})$ (if $i > n_x + 1$).
\end{itemize}

\end{document}

### 2025-03-24 article_bihom.tex
ajouts des critères, je suis à la section 3

@@
 \usepackage{amsfonts}
+\newtheorem{proposition}{Proposition}[section]
+\newcommand{\propositionTitre}[2]{
+	\begin{proposition}[#1]
+		#2
+	\end{proposition}
+}
+
+\title{Ce que j'ai compris dans l'article: Gröbner bases of bihomogeneous ideals generated by polynomials of bidegree (1, 1): Algorithms and complexity de Faugère et al.}
+
 \begin{document}
-Ce que j'ai compris dans l'article: Gröbner bases of bihomogeneous ideals generated by polynomials of bidegree (1, 1): Algorithms and complexity de Faugère et al.
-\newline
+\maketitle
 
@@
 
+\section{Théorèmes}
+\propositionTitre{Critère F5 général}{
+$$\forall j < m, \textit{si la ligne d'étiquette }(t,f_j) \textit{ dans la matrice } 
+\tilde{\mathbb{M}}_{d-d_m, m-1} \textit{ a pour terme de tête, } t'$$
+$$	\textit{alors la ligne } 
+(t', f_m) \textit{ dans la matrice } \mathbb{M}_{d, m} \textit{ est une combinaison linéaire des précédentes.}$$
+}
+
+\propositionTitre{Critère de Frobenius}{
+$$\forall j < m, \textit{si la ligne d'étiquette }(t,f_j) \textit{ dans la matrice } 
+\tilde{\mathbb{M}}_{d-d_m, m} \textit{ a pour monôme de tête, } t'$$
+$$	\textit{alors la ligne } 
+(t', f_m) \textit{ dans la matrice } \mathbb{M}_{d, m} \textit{ est une combinaison linéaire des précédentes.}$$
+}
+
 \end{document}

### 2025-03-24 article_bihom.tex
ajout de defs et thm, je suis à la section 3.3

@@
 \newtheorem{proposition}{Proposition}[section]
+\newtheorem{definition}{Définition}[section]
+\newtheorem{theoreme}{Théorème}[section]
+
 \newcommand{\propositionTitre}[2]{
@@
 \section{Définitions et notations}
+\begin{definition}
+\item $I_{i-1}:f_i$ is spanned by $I_{i-1}$ and by the maximal minors of $\text{jac}_{\textbf{x}}(F_{i-1})$ (if $i > n_y + 1$) and $\text{jac}_{\textbf{y}}(F_{i-1})$ (if $i > n_x + 1$).
+\end{definition}
+
+\begin{definition}
+Étant donné une matrice de taille $(k+1, k)$, on note minor$(\textbf{M}, j)$ le mineur obtenu en retirant la j-ième ligne de M. Considérons
+$$v = (minor(jac_\textbf{x}(F), 1), -minor(jac_\textbf{x}(F), 2),
+minor(jac_\textbf{x}(F), 3), -minor(jac_\textbf{x}(F), 4),)$$
+\end{definition}
+
 \begin{itemize}
-\item $I_{i-1}:f_i$ is spanned by $I_{i-1}$ and by the maximal minors of $\text{jac}_{\textbf{x}}(F_{i-1})$ (if $i > n_y + 1$) and $\text{jac}_{\textbf{y}}(F_{i-1})$ (if $i > n_x + 1$).
+	\item Soit \textbf{M} une matrice de taille $l \times c$. On appelle maximal 
+	minors de \textbf{M} les déterminants des sous-matrices de taille $c \times c$
+	de M.
 \end{itemize}
 
-\section{Théorèmes}
+\section{Théorèmes et propositions}
 \propositionTitre{Critère F5 général}{
@@
 
+\begin{theoreme}
+Soit $i > n_x + 1 \textit{ (resp. } i > n_y + 1\textit{) }$ et soit $s$ une 
+combinaison linéaire des maximal minors de $jac_\textbf{x}(F_{i-1}) 
+\textit{ (resp.} jac_\textbf{y}(F_{i-1}) \textit{) }$. Alors $s \in I_{i-1}:f_i$.
+\end{theoreme}
+
 \end{document}

### 2025-03-24 article_bihom.tex
dernier commit pour l'article ?

@@
 
+\begin{definition}
+	Soit $\prec$ un ordre monomial tel que sa restriction à 
+	$\mathbb{K}[x_0,...,x_{n_x}]$ (resp. $\mathbb{K}[y_0,...,y_{n_y}]$)
+	est l'ordre \textit{grevlex}. Soit $m \leq n_x + n_y$ et $f_1,...,fm$
+	des polynômes bilinéaires sur $\mathit{R}$. On dit qu'une séquence polynomiale
+	($f_1,...,f_m$) est une bi-régulière si $m = 1$ ou si ($f_1,...,f_{m-1}$)
+	est une séquence bi-régulière et
+	$$\textbf{LM}(I_{m-1}:f_m) == \langle\text{Monomials}^{x}_{m - n_y - 2}
+	(n_y + 1)\rangle + \langle\text{Monomials}^{y}_{m - n_x - 2}
+	(n_x + 1)\rangle + \langle\textbf{LM}(I_{m-1})\rangle$$
+\end{definition}
+
 \section{Théorèmes et propositions}

### 2025-03-31 article_bihom_v2.tex
ajout de ce que j'ai compris

% !TXS template
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[a4paper]{geometry}
\usepackage{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage[backend=bibtex]{biblatex}
\usepackage{csquotes}
\addbibresource{biblio.bib}
\newtheorem{proposition}{Proposition}[section]
\newtheorem{definition}{Définition}[section]
\newtheorem{theoreme}{Théorème}[section]

\newcommand{\propositionTitre}[2]{
	\begin{proposition}[#1]
		#2
	\end{proposition}
}

\author{Paul Mekhail\\ \textbf{Encadrante:} Sorina Ionica}

\title{Ce que j'ai compris jusqu'à maintenant\\Début de rapport ?}

\begin{document}
	\maketitle

	\section{Introduction}
		The MPC in the head paradigm is a new framework introduced in \cite{IKOS07} which is in vogue recently as we saw in NIST's second round competition for post-quantum signature schemes where 5 of the 14 candidates are based on this paradigm.
		The general idea is to use a NP-relation $\mathcal{R}(x, w)$ to obtain a ZK-protocol in which a prover $\mathcal{P}$ convinces a verifier $\mathcal{V}$ that she knows a valid witness $w$ for a given (public) value of $x$ without revealing any information about $x$.
		In \cite{HJ23}, the authors introduce a problem called subfield bilinear collision problem (SBC) which originates from the discrete logarithm problem and constructed a post-quantum signature scheme based on this problem using the MPCitH framework.
		The goal of this internship is to perform a cryptanalysis of this signature scheme.

		\section{Signature Scheme}
		\subsection{SBC Problem}
		The problem is the following: let $q$ be a prime power and two positive integers $k$, $n$.
		\newline
		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent overs $\mathbb{F_q}$, find two vectors $\vec{x}, \vec{y} \in (\mathbb{F_q})^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})((\vec{v} \cdot \vec{x})$$

		\subsection{NSBC}
		The authors also introduce a particular version of this problem called normalized SBC (NSBC) where $\vec{x}, \vec{y} \in (\mathbb{F_q})^n$ and
		$\vec{x} = (\vec{x'}, 1, 0), \vec{y} = (\vec{y}, 0, 1)$.

 		\subsection{Keys and parameters estimations}
		The public key of this signature scheme is $(\vec{u},\vec{v})$ and the private key is $(\vec{x}, \vec{y})$.
		The signature and verification protocol use the MPCitH paradigm, they are described in detail in \cite{HJ23} with proof of soundness, correctness and zero-knowledge.
		The authors estimate that a good parameter for NSBC problem to be computationally secure are $q = 2, n = 130, k = 257$.

		\section{Cryptanalysis of the scheme}
		One of the known attacks given by the authors is through algebraic cryptanalysis.
		One can model system by taking $g$ as the following polynomial
		$$
		g(x_1,...,x_n,...,y_1,y_n) := (\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) - (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})
		$$
		or if we consider the NSBC problem
		$$
		g(x_1,..,x_{n-2},y_1,..,y_{n-2}) := (\sum_{1}^{n-2}u_{i}x_{i} + u_{n-1})(\sum_{1}^{n-2}v_{i}y_{i} + v_{n}) - (\sum_{1}^{n-2}u_{i}y_{i} + u_{n-1})(\sum_{1}^{n-2}v_{i}x_{i} + v_{n-1})
		$$

		Then, one can perform Weil descent on the polynomial $g$ because $\mathbb{F}_q^k$ is a $\mathbb{F}_q$-vector space and obtain the polynomial system

		\begin{align}
			g_1(x_1, \dots, x_{n-2}, y_1, \dots, y_{n-2}) &= 0 \notag \\
			&\vdots \notag \\
			g_k(x_1, \dots, x_{n-2}, y_1, \dots, y_{n-2}) &= 0 \notag
		\end{align}

		The most straightforward cryptanalysis of this type of systems is to perform a Gröbner basis algorithm like Faugère's F5 \cite{F02}.

		The authors cite \cite{FSS11} to argue their parameters choice, we will dive deeper into this article's results in a later section.

		\subsection{Gröbner basis algorithms}
		Gröbner basis algorithms are used to find solutions for polynomial systems using algebraic geometry tools. For a gentle introduction to this subject \cite{CLS}.

		The major "issue" of Gröbner basis algorithms is the reductions to 0, because of the matrices sizes that are huge, multiple millions for practical examples,
		reductions to 0 can cost a lot during the linear algebra process. F5 \cite{F02} was a major improvement because it provided a criterion that avoided reduction to 0 during the linear algebra part for all regular sequences.
		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $d = H(I) -1$ where $I$ is the ideal generated by the polynomial sequence and $H(I)$ is the index of regularity of the ideal $I$.

\printbibliography

\end{document}

### 2025-03-31 article_bihom_v2.tex
fin du début de la partie sur GB

@@
 \newtheorem{proposition}{Proposition}[section]
-\newtheorem{definition}{Définition}[section]
-\newtheorem{theoreme}{Théorème}[section]
+\newtheorem{definition}{Definition}[section]
+\newtheorem{theorem}{Theorem}[section]
 
@@
 F5 \cite{F02} was a major improvement because it provided a criterion that avoided reduction to 0 during the linear algebra part for all regular sequences.
-		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $d = H(I) -1$ where $I$ is the ideal generated by the polynomial sequence and $H(I)$ is the index of regularity of the ideal $I$.
+		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $D$ which is the degree of regularity of $I$.
+
+		According to \cite{BFS15}, the complexity of computing the Gröbner basis of a system $f_1,...,f_m \in \mathbb{K}[x_1,...,x_n]$ is $O(mD\binom{n+D-1}{D}^\omega)$.
+
+		In \cite{Bardet04} Theorem 3.2.10 is the following
+
+		\begin{theorem}
+			Let $f_1,...,f_m$ homogeneous sequence, such that $<f_1,...,f_m>$ is 0-dimensional and
+			$<$ an admissible graded monomial ordering. We have,
+			\begin{itemize}
+				\item[-] If the sequence is semi-regular, then no reductions to 0 are performed during the F5-matrix algorithm until the degree of regularity $D - 1$.
+				\item[-] If there is no reductions to 0 during the F5-matrix algorithm until the degree $D - 1$, and if the matrix of degree D is full rank and is the first matrix to have more rows than columns, then the sequence is semi-regular and it's index of regularity is $H(I) = D$.
+			\end{itemize}
+		\end{theorem}
+
+		However, after some tests and after applying F5-matrix algorithm on polynomial systems generated by NSBC instances, we get reductions to 0, this means that the polynomial systems are not semi-regular.
+
+		\subsection{Bilinear systems case}
 

### 2025-04-01 article_bihom_v2.tex
manque la partie sur crossbred

@@
 \usepackage{csquotes}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{hyperref}
+
 \addbibresource{biblio.bib}
+
 \newtheorem{proposition}{Proposition}[section]
@@
 \newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Corollary}[section]
 
@@
 		In \cite{HJ23}, the authors introduce a problem called subfield bilinear collision problem (SBC) which originates from the discrete logarithm problem and constructed a post-quantum signature scheme based on this problem using the MPCitH framework.
-		The goal of this internship is to perform a cryptanalysis of this signature scheme.
+		The goal of this internship is to perform a cryptanalysis of this signature scheme and try to improve it.
 
@@
 		\newline
-		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent overs $\mathbb{F_q}$, find two vectors $\vec{x}, \vec{y} \in (\mathbb{F_q})^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})((\vec{v} \cdot \vec{x})$$
+		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent overs $\mathbb{F}_q$, find two vectors $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})((\vec{v} \cdot \vec{x})$$
 
 		\subsection{NSBC}
-		The authors also introduce a particular version of this problem called normalized SBC (NSBC) where $\vec{x}, \vec{y} \in (\mathbb{F_q})^n$ and
-		$\vec{x} = (\vec{x'}, 1, 0), \vec{y} = (\vec{y}, 0, 1)$.
+		The authors also introduce a particular version of this problem called normalized SBC (NSBC) where $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ and
+		$\vec{x} = (\vec{x'}, 1, 0), \vec{y} = (\vec{y'}, 0, 1)$.
 
@@
 		The signature and verification protocol use the MPCitH paradigm, they are described in detail in \cite{HJ23} with proof of soundness, correctness and zero-knowledge.
-		The authors estimate that a good parameter for NSBC problem to be computationally secure are $q = 2, n = 130, k = 257$.
+		The authors estimate that a good parameter for NSBC problem to be computationally secure are $q = 2, n = 130, k = 257$ using complexity analysis of what they think is the best known cryptanalysis of such problem in \cite{FSS11}.
 
@@
 		One of the known attacks given by the authors is through algebraic cryptanalysis.
-		One can model system by taking $g$ as the following polynomial
+		One can model the system by taking $g$ as the following polynomial
 		$$
-		g(x_1,...,x_n,...,y_1,y_n) := (\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) - (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})
+		g(x_1,...,x_n,y_1,...,y_n) := (\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) - (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})
 		$$
@@
 		$$
-		g(x_1,..,x_{n-2},y_1,..,y_{n-2}) := (\sum_{1}^{n-2}u_{i}x_{i} + u_{n-1})(\sum_{1}^{n-2}v_{i}y_{i} + v_{n}) - (\sum_{1}^{n-2}u_{i}y_{i} + u_{n-1})(\sum_{1}^{n-2}v_{i}x_{i} + v_{n-1})
+		g(x_1,..,x_{n-2},y_1,..,y_{n-2}) := (\sum_{i = 1}^{n-2}u_{i}x_{i} + u_{n-1})(\sum_{i = 1}^{n-2}v_{i}y_{i} + v_{n}) - (\sum_{i = 1}^{n-2}u_{i}y_{i} + u_{n})(\sum_{i = 1}^{n-2}v_{i}x_{i} + v_{n-1})
 		$$
 
-		Then, one can perform Weil descent on the polynomial $g$ because $\mathbb{F}_q^k$ is a $\mathbb{F}_q$-vector space and obtain the polynomial system
+		Then, one can perform the Weil descent on the polynomial $g$ because $\mathbb{F}_{q^k}$ is a $\mathbb{F}_q$-vector space and obtain the polynomial system
 
 		\begin{align}
-			g_1(x_1, \dots, x_{n-2}, y_1, \dots, y_{n-2}) &= 0 \notag \\
+			g_1(x_1, \dots, x_{n-2},& y_1, \dots, y_{n-2}) = 0 \notag \\
 			&\vdots \notag \\
-			g_k(x_1, \dots, x_{n-2}, y_1, \dots, y_{n-2}) &= 0 \notag
+			g_k(x_1, \dots, x_{n-2},& y_1, \dots, y_{n-2}) = 0 \notag
 		\end{align}
@@
 
-		The major "issue" of Gröbner basis algorithms is the reductions to 0, because of the matrices sizes that are huge, multiple millions for practical examples,
+		The major "issue" of Gröbner basis algorithms is the reductions to 0, because of the size of the matrices are huge, multiple millions for practical examples,
 		reductions to 0 can cost a lot during the linear algebra process.
@@
 
-		According to \cite{BFS15}, the complexity of computing the Gröbner basis of a system $f_1,...,f_m \in \mathbb{K}[x_1,...,x_n]$ is $O(mD\binom{n+D-1}{D}^\omega)$.
-
-		In \cite{Bardet04} Theorem 3.2.10 is the following
-
-		\begin{theorem}
-			Let $f_1,...,f_m$ homogeneous sequence, such that $<f_1,...,f_m>$ is 0-dimensional and
-			$<$ an admissible graded monomial ordering. We have,
+		\begin{theorem}[Theorem 3.2.10 in \cite{Bardet04}]
+			Let $f_1,...,f_m$ homogeneous sequence, such that $\langle f_1,...,f_m \rangle$ is 0-dimensional and
+			$\prec$ an admissible graded monomial ordering. We have,
 			\begin{itemize}
@@
 
+		In \cite{BFS15}, the authors show that the complexity of computing the Gröbner basis of a semi-regular system $f_1,...,f_m \in \mathbb{K}[x_1,...,x_n]$ is $O\left(mD\binom{n+D-1}{D}^\omega \right)$.
+
 		However, after some tests and after applying F5-matrix algorithm on polynomial systems generated by NSBC instances, we get reductions to 0, this means that the polynomial systems are not semi-regular.
@@
 		\subsection{Bilinear systems case}
+		Our NSBC instances are quadratic systems that are bilinear in the variables of degree 2, are not semi-regular and overdetermined.
+
+		We will note $n_x$ (resp.
+$n_y$) the number of variables in $x$ (resp.
+number of variables in y).
+In our case, $n_x = n_y = (m-1)/2$ and $m = k$.
+As we mentioned earlier, our polynomials systems are not-semi regular, so we get reductions to 0 during the F5 algorithm using the Frobenius criterion for boolean polynomial systems.
+
+		The article \cite{FSS11} and the thesis \cite{Spaen2012} gives a partial answer to this problem by providing a new criterion for bilinear determined systems and a complexity analysis of their algorithm (The proofs in the end of \cite{FSS11} are not exact, the authors recommended to refer to \cite{Spaen2012} for better proofs.).
+
+		\subsubsection{Jacobian matrices and syzygies}
+		First we will need to introduce some important notations.
+		\begin{itemize}
+			\item[-] $R = \mathbb{K}[x_1,...,x_{n_x + 1},y_1,...,y_{n_y + 1}]$
+			\item[-] Let $\mathbb{K}$ be a field and $f_1,...,f_m \in R$ be a bilinear polynomial.
+We denote by $F_i$ the polynomial sequence $(f_1,...,f_i)$ and $I_i$ the ideal spanned by this sequence $\langle F_i \rangle$.
+			\item[-] $jac_{x}(F_{i-1})$ (resp. $jac_{y}(F_{i-1})$) is the jacobian matrix with respect to the two subsets of variables $x_1,...,x_{n_x}$ (resp. $y_1,...,y_{n_y}$).
+			\item[-] Let $M$ be a $l \times c$ matrix with $l > c$. We call maximal minors of $M$ the determinants of the $c \times c$ sub-matrices of $M$. 
+			\item[-] $I_{i-1} : f_i$ is the ideal spanned by $I_{i-1}$ and by the maximal minors of $jac_{x}(F_{i-1})$ if ($i > n_y$) and $jac_{y}(F_{i-1})$ if ($i > n_x$).
+		\end{itemize}
+		We will give important results of this article before discussing them further.
+
+		\begin{theorem}[Theorem 2 in \cite{FSS11}]
+			Let $i > n_x + 1$ (resp. $i > n_y + 1$) and let $s$ be a linear combination of maximal minors of $jac_{x}(F_{i-1})$ (resp. $jac_{y}(F_{i-1})$). Then $s \in I_{i-1} : f_i$
+		\end{theorem}
+
+		The following algorithms will give us the signatures of reductions to 0 during F5-matrix algorithm.
+
+		\begin{algorithm}
+			\caption{Reduce}\label{alg:cap}
+			\begin{algorithmic}[1]
+				\Require A monomial ordering $\prec$ and $S$ a set of homogeneous polynomials and $q$ a degree.
+				\Ensure $T$ a reduced set of homogeneous polynomials of degree $q$.
+				\State $M \gets Macaulay_{\prec}(S, q)$.
+				\State $M \gets RowEchelonForm(M)$.
+				\State \textbf{Return} $T$ the set of polynomials corresponding to the rows of M.
+			\end{algorithmic}
+		\end{algorithm}
+
+		 \begin{algorithm}
+		 	\caption{BL\_Criterion}
+		 	\begin{algorithmic}[1]
+		 		\Require $m$ bilinear polynomials $f1,...f_m$ such that $m \leq n_x + n_y$.\newline $\prec$ a monomial ordering overs $R$.
+		 		\Ensure $V$ a set of pairs $(h, f_i)$ such that $h \in I_{i-1} : f_i$ and $h \notin I_{i-1}$.
+		 		\State $V \gets \emptyset$
+		 		\For{$i$ from 2 to $m$}
+					\If{$i > n_y$}
+						\State $T \gets \mathbf{Reduce}(MaxMinors(jac_y(F_{i-1})), n_y)$
+						\For{$h$ in $T$}
+							\State $V \gets V \cup \{(h, f_i)\}$
+						\EndFor
+					\EndIf
+					\If{$i > n_x$}
+						\State $T' \gets \mathbf{Reduce}(MaxMinors(jac_x(F_{i-1})), n_x)$
+						\For{$h$ in $T'$}
+							\State $V \gets V \cup \{(h, f_i)\}$
+						\EndFor
+					\EndIf
+				\EndFor
+				\State \textbf{Return} $V$
+		 	\end{algorithmic}
+		 \end{algorithm}
+
+		 \begin{algorithm}
+		 	\caption{BilinF5Critetion}
+		 	\begin{algorithmic}[1]
+		 		\Require $(t, f_i)$ the signature of a row. \newline A matrix $M$ in row echelon form.
+		 		\Ensure $True$ if the row will reduce to 0 otherwise $False$.
+		 		\If{$t$ is the leading monomial of a row of $M$ \textbf{or}
+		 		\newline $\exists (h, f_i) \in V$ such that $LM(h) = t$}
+		 		\State \textbf{Return} $True$.
+		 		\Else \textbf{ Return} $False$.
+		 		\EndIf
+		 	\end{algorithmic}
+		 \end{algorithm}
+
+		 The following theorem, ensures that there are no reductions to 0 with extended the F5 criterion.
+
+		 \begin{theorem}[Theorem 4 in \cite{FSS11}]
+		 	Let $m, n_x, n_y \in \mathbb{N}$ such that $m < n_x + n_y$.
+If Conjecture 1 of \cite{FSS11} is true, then the set of bi-regular sequences $(f_1,...,f_m)$ contains a nonempty Zariski open set.
+Moreover, if $(f_1,...,f_m)$ is a bi-regular sequence, then there are no reductions to zero with the extended F5 criterion.
+		 \end{theorem}
+
+		 For a formal definition of a \textit{bi-regular} sequence, we refer to definition 8 of \cite{FSS11}.
+
+		 Finally, the authors give the following corollary
+
+		 \begin{corollary}[Corollary 3 in \cite{FSS11}]
+		 The arithmetic complexity of computing a Gröbner basis of a generic bilinear system $f_1,...,f_{n_x + n_y}$ with the F5 algorithm is bounded by
+		 $$\mathcal{O}\left(\binom{n_x + n_y + min(n_x + 1, n_y + 1)}{min(n_x + 1, n_y + 1)}^\omega\right)$$
+		 where $2 \leq \omega \leq 3$ is the linear algebra constant.
+		 \end{corollary}
+
+		 First, we can clearly see that the complexity of algorithm \ref{alg:cap} is exponential and is exactly 
+		 $$
+		 \mathcal{O}\left(2n_x^{\omega}\left(\sum_{i = n_x}^{m} \binom{i}{n_x}\right) + 2n_y^{\omega}\left(\sum_{i = n_y}^{m} \binom{i}{n_y}\right)\right)
+		 $$
+		 It's an asymptotic complexity which doesn't take into account the constant factor in implementations that could slow down a lot the algorithm.
+		 We also recall that in our case with NSBC, $n_x = n_y$. So we get
+		 $$
+		 \mathcal{O}\left(4n_x^{\omega}\left(\sum_{i = n_x}^{m} \binom{i}{n_x}\right)\right)
+		 $$
+
+		 Since we have quadratic systems, reductions to 0 won't appear before degree 4 of the Macaulay matrix.
+Also, in practice we are using $q=2$ and there's much better algorithms in the literature for boolean systems than basic Gröbner basis algorithms like F5.
+The best algorithm used at the moment is CrossBred \cite{JV17}.
+Most of the time in practical implementations, we stop at degree 4 of the Macaulay in pre-processing step.
+So, to know what advantage we used the extended F5 criterion, we will suppose the worst, that all of our entries will reduce to 0 at degree 4 of the Macaulay matrix, which is unlikely.
+
+		 In following, we will denote $n_x + n_y$ by simply $n$.
+
+		 If this case arises, the only computations that would be done are for extended F5 criterion. The complexity for F5-Matrix algorithm with only the Frobenius criterion is 
+		 $$
+		 \mathcal{O}\left(\binom{n + 4}{4}^\omega\right)
+		 $$
+		 If we only take into consideration the big O notation and not the constants, which is not realistic at all for efficient implementations to really break a cryptographic scheme, we get
+		 \begin{equation}
+		 Adv(n, m) = 2\left(\frac{n}{2}\right)^\omega \left(\sum_{i = n/2}^{m}\binom{i}{n/2}\right) - \binom{n + 4}{4}^\omega
+		 \label{equation_complex}
+		 \end{equation}
+
+		 Using the computer algebra system SageMath \cite{sagemath} to test numerically when using the F5 extended criterion of \cite{FSS11} becomes negligible compared to the computations of reductions to 0, we see experimentally that $Adv(20, 21) = 28174524$, i.e.
+the theoretical advantage becomes negligible and it would be advantageous to simply compute reductions to 0 instead of applying the F5 extended criterion.
+This number grows exponentially with $n$ and $m$.
+For those who wonder, what it would be the estimated theoretical advantage computing reductions to 0 for the parameters given by \cite{HJ23}
+		 $$
+		 Adv(256, 257) = 753178991751680363947858801037843009326048264526089645542944265160485518735431295
+		 $$
+
+		 No need to say that it's useless to use the criterion as shown in \cite{FSS11} without adapting it to get an advantage.
+		 This can be explained by the fact that the extended criterion computes reduced polynomials of degree $n_x$ and $n_y$ and in our case, $n_x$ and $n_y$ can be too big to compute.
+Maybe there could be a way to compute reduced polynomials of only degree 2 for our use case if we stop Macaulay matrix computations at degree 4 with our quadratic systems.
+		 \newline
+		 Note pour moi-même, il faut maintenant essayer de réaliser une implémentation de ce critère en sage ou avec hpXbred et BeanPolE pour voir comment ça se comporterait avec des petits systèmes et voir quel serait l'avantage ou le désavantage sachant que la supposition que toutes les lignes se réduisent à 0 est extrêment peu probable.
+
+		 \subsection{Boolean systems and Crossbred}
 

### 2025-04-01 article_bihom_v2.tex
ajout de preuve de complexité et petites corrections

@@
 \usepackage{amsfonts}
+\usepackage{amsthm}
 \usepackage{mathtools}
@@
 		\begin{algorithm}
-			\caption{Reduce}\label{alg:cap}
+			\caption{Reduce}\label{alg:reduc}
 			\begin{algorithmic}[1]
@@
 		 \begin{algorithm}
-		 	\caption{BL\_Criterion}
+		 	\caption{BL\_Criterion}\label{alg:crit_calc}
 		 	\begin{algorithmic}[1]
@@
 					\If{$i > n_y$}
-						\State $T \gets \mathbf{Reduce}(MaxMinors(jac_y(F_{i-1})), n_y)$
+						\State $T \gets \mathbf{Reduce}(MaxMinors(jac_y(F_{i-1})), n_y + 1)$
 						\For{$h$ in $T$}
@@
 					\If{$i > n_x$}
-						\State $T' \gets \mathbf{Reduce}(MaxMinors(jac_x(F_{i-1})), n_x)$
+						\State $T' \gets \mathbf{Reduce}(MaxMinors(jac_x(F_{i-1})), n_x + 1)$
 						\For{$h$ in $T'$}
@@
 		 \begin{algorithm}
-		 	\caption{BilinF5Critetion}
+		 	\caption{BilinF5Critetion}\label{algo:crit}
 		 	\begin{algorithmic}[1]
@@
 
-		 First, we can clearly see that the complexity of algorithm \ref{alg:cap} is exponential and is exactly 
+		 First, we can clearly see that the complexity of algorithm \ref{alg:crit_calc} is exponential and is what we will try to show.
+
+		 \begin{proposition}
+		 	The complexity of algorithm \ref{alg:crit_calc} is
 		 	$$
-		 \mathcal{O}\left(2n_x^{\omega}\left(\sum_{i = n_x}^{m} \binom{i}{n_x}\right) + 2n_y^{\omega}\left(\sum_{i = n_y}^{m} \binom{i}{n_y}\right)\right)
+		 	\mathcal{O}\left(\left(\sum_{i = n_x}^{m} \binom{i}{n_x}n_x^{\omega} + \binom{2n_x + 1}{n_x}^\omega\right) + \left(\sum_{i = n_y}^{m} \binom{i}{n_y}n_y^{\omega} + \binom{2n_y + 1}{n_y}^\omega\right)\right)
 		 	$$
+		 \end{proposition}
+
+		 \begin{proof}
+		 	Fist, if $i > n_y$ (resp.
+$i > n_x$), we construct that jacobian matrix $jac_y(F_{i-1})$ (resp.
+$jac_x(F_{i-1})$) which is of size $n_y \times n_y$ (resp.
+$n_x \times n_x$)for the first one and $(m-1) \times n_y$ (resp.
+$(m-1) \times n_x$) for the last one.
+We suppose here that the construction of such matrices is free.
+Computing the $MaxMinors(jac_y(F_{i-1}))$ (resp.
+$MaxMinors(jac_x(F_{i-1}))$)means that we have to compute the determinants of all the sub-matrices of size $n_y \times n_y$ (resp.
+$n_x \times n_x$) and for a matrix $M$ of size $l \times c$ with $l > c$ there are $\binom{l}{c}$ sub-matrices of size $c \times c$, hence the term $\binom{i}{n_y}n_y^{\omega}$ in the sum if we suppose the complexity of computing the determinant of a square multivariate polynomial matrix of size $c \times c$ to be $\mathcal{O}(c^\omega)$, which is not the case but let's suppose the best case ever.
+faudrait trouver la vraie complexité de ça pour être plus juste.
+		 	For now, we have a sequence of polynomials and we apply the algorithm \textbf{Reduce} on them with degree $n_y + 1$ (resp. $n_x + 1$).
+		 	In \textbf{Reduce}, we construct the Macaulay matrix of the sequence of degree $n_y + 1$ (resp. $n_x + 1$) and compute it's row echelon form.
+		 	The size of Macauly matrix is bounded by $\binom{n + D}{D}$ where D is the degree and n the number variables.
+So the complexity of \textbf{Reduce} is $\mathcal{O}\left(\binom{n_y + n_y + 1}{n_y + 1}^\omega \right)$ (resp.
+$\mathcal{O}\left(\binom{n_x + n_x + 1}{n_x + 1}^\omega \right)$).
+Hence the second term in the sum, because we call \textbf{Reduce} at each iteration of the for loop.
+We finally that the iteration on the elements of $T$ and of $T'$ is negligible in the big O notation.
+		 \end{proof}
+
 		 It's an asymptotic complexity which doesn't take into account the constant factor in implementations that could slow down a lot the algorithm.
@@
 		 $$
-		 \mathcal{O}\left(4n_x^{\omega}\left(\sum_{i = n_x}^{m} \binom{i}{n_x}\right)\right)
+		 \mathcal{O}\left(2n_x^{\omega}\left(\sum_{i = n_x}^{m} \binom{i}{n_x} + \binom{2n_x + 1}{n_x}^\omega\right)\right)
 		 $$
@@
 		 \begin{equation}
-		 Adv(n, m) = 2\left(\frac{n}{2}\right)^\omega \left(\sum_{i = n/2}^{m}\binom{i}{n/2}\right) - \binom{n + 4}{4}^\omega
+		 Adv(n, m) = 2\left(\sum_{i = n/2}^{m}\binom{i}{n/2}\left(\frac{n}{2}\right)^\omega + \binom{n + 1}{n/2}^\omega\right) - \binom{n + 4}{4}^\omega
 		 \label{equation_complex}
@@
 
-		 Using the computer algebra system SageMath \cite{sagemath} to test numerically when using the F5 extended criterion of \cite{FSS11} becomes negligible compared to the computations of reductions to 0, we see experimentally that $Adv(20, 21) = 28174524$, i.e.
+		 Using the computer algebra system SageMath \cite{sagemath} to test numerically when using the F5 extended criterion of \cite{FSS11} becomes negligible compared to the computations of reductions to 0, we see experimentally that $Adv(10, 12) = 2498903$, i.e.
 the theoretical advantage becomes negligible and it would be advantageous to simply compute reductions to 0 instead of applying the F5 extended criterion.
 This number grows exponentially with $n$ and $m$.
-For those who wonder, what it would be the estimated theoretical advantage computing reductions to 0 for the parameters given by \cite{HJ23}
-		 $$
-		 Adv(256, 257) = 753178991751680363947858801037843009326048264526089645542944265160485518735431295
-		 $$
+For those who wonder, what it would be the estimated theoretical advantage computing reductions to 0 for the parameters given by \cite{HJ23} $Adv(256, 257)$ is a 514 bit number.
 

### 2025-04-02 article_bihom_v2.tex
c'est finalement semi-régulier

@@
 
-		However, after some tests and after applying F5-matrix algorithm on polynomial systems generated by NSBC instances, we get reductions to 0, this means that the polynomial systems are not semi-regular.
+		However, after some tests and after applying F5-matrix algorithm on polynomial systems generated by NSBC instances, we get reductions to 0, this means that the polynomial systems are not semi-regular when we only multiply by the $x$ variables or by the $y$ variables.
+If we multiply by the $x$ variables and at least one $y$ variable, we get no reductions to 0 i.e.
+the system is semi-regular
 
 		\subsection{Bilinear systems case}
-		Our NSBC instances are quadratic systems that are bilinear in the variables of degree 2, are not semi-regular and overdetermined.
+		Our NSBC instances are quadratic systems that are bilinear in the variables of degree 2, overdetermined and can be semi-regular.
 
@@
 In our case, $n_x = n_y = (m-1)/2$ and $m = k$.
-As we mentioned earlier, our polynomials systems are not-semi regular, so we get reductions to 0 during the F5 algorithm using the Frobenius criterion for boolean polynomial systems.
+As we mentioned earlier, our polynomials systems can be semi-regular, so we get no reductions to 0 during the F5 algorithm using the Frobenius criterion for boolean polynomial systems.
 
@@
 Maybe there could be a way to compute reduced polynomials of only degree 2 for our use case if we stop Macaulay matrix computations at degree 4 with our quadratic systems.
+
+		 As the authors of \cite{FSS11} noted at the end of their article, generic bilinear systems are regular.
+They didn't show that for overdetermined systems but we saw that experimentally but it's yet to be proven.
 		 \newline

### 2025-04-03 article_bihom_v2.tex
version envoyé

@@
 
-		Then, one can perform the Weil descent on the polynomial $g$ because $\mathbb{F}_{q^k}$ is a $\mathbb{F}_q$-vector space and obtain the polynomial system
+		Then, one can perform a Weil descent on the polynomial $g$ because $\mathbb{F}_{q^k}$ is a $\mathbb{F}_q$-vector space and obtain the polynomial system
 
@@
 
-		The major "issue" of Gröbner basis algorithms is the reductions to 0, because of the size of the matrices are huge, multiple millions for practical examples,
+		The major "issue" of Gröbner basis algorithms is the reductions to 0, because the size of the matrices are huge, multiple millions for practical examples,
 		reductions to 0 can cost a lot during the linear algebra process.
 F5 \cite{F02} was a major improvement because it provided a criterion that avoided reduction to 0 during the linear algebra part for all regular sequences.
-		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $D$ which is the degree of regularity of $I$.
+		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $D$ which is the degree of regularity of $I$ which is the ideal generated by the polynomials of the system.
 
 		\begin{theorem}[Theorem 3.2.10 in \cite{Bardet04}]
-			Let $f_1,...,f_m$ homogeneous sequence, such that $\langle f_1,...,f_m \rangle$ is 0-dimensional and
+			Let $f_1,...,f_m$ a homogeneous sequence, such that $\langle f_1,...,f_m \rangle$ is 0-dimensional and
 			$\prec$ an admissible graded monomial ordering. We have,
@@
 		However, after some tests and after applying F5-matrix algorithm on polynomial systems generated by NSBC instances, we get reductions to 0, this means that the polynomial systems are not semi-regular when we only multiply by the $x$ variables or by the $y$ variables.
-If we multiply by the $x$ variables and at least one $y$ variable, we get no reductions to 0 i.e.
-the system is semi-regular
+If we multiply by the $x$ variables and at least one $y$ variable, we get no reductions to 0 which means that the system could be semi-regular.
 
@@
 In our case, $n_x = n_y = (m-1)/2$ and $m = k$.
-As we mentioned earlier, our polynomials systems can be semi-regular, so we get no reductions to 0 during the F5 algorithm using the Frobenius criterion for boolean polynomial systems.
+As we mentioned earlier, our polynomials systems can be semi-regular, so we get no reductions to 0 during the F5 algorithm using the Frobenius criterion for boolean polynomial systems until the Macaulay matrix of degree $D_{reg}$.
 
-		The article \cite{FSS11} and the thesis \cite{Spaen2012} gives a partial answer to this problem by providing a new criterion for bilinear determined systems and a complexity analysis of their algorithm (The proofs in the end of \cite{FSS11} are not exact, the authors recommended to refer to \cite{Spaen2012} for better proofs.).
+		The article \cite{FSS11} and the thesis \cite{Spaen2012} give a partial answer to this problem by providing a new criterion for bilinear determined systems and a complexity analysis of their algorithm (The proofs in the end of \cite{FSS11} are not exact, the authors recommended to refer to \cite{Spaen2012} for better proofs.).
 
@@
 		\begin{itemize}
-			\item[-] $R = \mathbb{K}[x_1,...,x_{n_x + 1},y_1,...,y_{n_y + 1}]$
+			\item[-] $R = \mathbb{K}[x_1,...,x_{n_x},y_1,...,y_{n_y}]$
 			\item[-] Let $\mathbb{K}$ be a field and $f_1,...,f_m \in R$ be a bilinear polynomial.
@@
 		 \begin{proof}
-		 	Fist, if $i > n_y$ (resp.
+		 	First, if $i > n_y$ (resp.
 $i > n_x$), we construct that jacobian matrix $jac_y(F_{i-1})$ (resp.
 $jac_x(F_{i-1})$) which is of size $n_y \times n_y$ (resp.
-$n_x \times n_x$)for the first one and $(m-1) \times n_y$ (resp.
+$n_x \times n_x$) for the first one and $(m-1) \times n_y$ (resp.
 $(m-1) \times n_x$) for the last one.
@@
 Hence the second term in the sum, because we call \textbf{Reduce} at each iteration of the for loop.
-We finally that the iteration on the elements of $T$ and of $T'$ is negligible in the big O notation.
+We finally suppose that the iteration on the elements of $T$ and of $T'$ is negligible in the big O notation.
 		 \end{proof}
@@
 Most of the time in practical implementations, we stop at degree 4 of the Macaulay in pre-processing step.
-So, to know what advantage we used the extended F5 criterion, we will suppose the worst, that all of our entries will reduce to 0 at degree 4 of the Macaulay matrix, which is unlikely.
+So, to know what advantage we could have if we use the extended F5 criterion, we will suppose the worst, that all of our entries will reduce to 0 at degree 4 of the Macaulay matrix, which is unlikely.
 
@@
 They didn't show that for overdetermined systems but we saw that experimentally but it's yet to be proven.
-		 \newline
+
+		 \begin{definition}[Definition 4 of \cite{VID24}]
+		 	Let $\mathcal{F} = {f_1,...,f_m}$ be a semi-regular sequence of polynomials in $\mathbb{F}_2[x_1,...,x_n]$ and let $0 \leq \gamma \leq 1$ such that $k = (1 - \gamma)n$.
+We say that this sequence is $\gamma$-strong semi-regular if
+		 	$$
+		 	\mathcal{S}(I) = \{(a_{k+1},...,a_n) \in \mathbb{F}_{2}^{n-k} |
+		 	$$
+		 	$$
+		 	\{f_1(x_1,...,x_k,a_{k+1},...,a_n),...,f_m(x_1,...,x_k,a_{k+1},...,a_n)\} \text{ is not semi-regular}\}
+		 	$$
+		 	has cardinality $\mathcal{O}(2^{-\gamma n})$
+		 \end{definition}
+
 		 Note pour moi-même, il faut maintenant essayer de réaliser une implémentation de ce critère en sage ou avec hpXbred et BeanPolE pour voir comment ça se comporterait avec des petits systèmes et voir quel serait l'avantage ou le désavantage sachant que la supposition que toutes les lignes se réduisent à 0 est extrêment peu probable.

### 2025-04-04 article_bihom_v2.tex
v1.5

@@
 \newtheorem{corollary}{Corollary}[section]
+
+\newcommand{\Mac}[3]{{Mac_{#1, #2}(\mathcal{#3})}}
+\newcommand{\Mact}[3]{{\widetilde{Mac}_{#1, #2}(\mathcal{#3})}}
 
@@
 
-		The major "issue" of Gröbner basis algorithms is the reductions to 0, because the size of the matrices are huge, multiple millions for practical examples,
+		In \cite{Lazard83}, Lazard introduced the connection between linear algebra and Gröbner basis theory via the usage of Macaulay matrices which is defined as follows.
+
+		\begin{definition}
+			Let $\prec$ be an admissible monomial ordering on a given ring $R = \mathbb{K}[x_1,\dots,x_n]$ where $\mathbb{K}$ is a field.
+Given a sequence of homogeneous (resp.
+affine) polynomials $\mathcal{F} = (f_1,\dots,f_m) \in R$, we associate to it the Macaulay matrix of degree $D$ (resp.
+$\leq D$), denoted by $\Mac{D}{m}{\mathcal{F}}$ (resp.
+$\Mac{\leq D}{m}{\mathcal{F}}$), and defined as follows: the columns of the matrix are indexed by the monomials of degree $D$ (resp.
+$\leq D$) sorted in decreasing order from the left to right using the defined monomial ordering.
+Each row of the matrix is labeled by a tag, also called signature, ($u, f_i$) where u is a monomial in $R$ and $f_i \in \mathcal{F}$ such that $deg(uf_i) = D$ (resp.
+$deg(uf_i) \leq D$), and the polynomial $uf_i$ is written as vector of coefficients of monomials on the row.
+We denote by $\Mact{D}{m}{F}$ (resp.
+$\Mact{\leq D}{m}{F}$) the Macaulay matrix on which we apply the row echelon form without swapping columns to preserve the monomial ordering.
+		\end{definition}
+
+		The major "issue" of Gröbner basis algorithms is the reductions to 0 in the Macaulay matrix associated to a polynomial sequence $\Mac{D}{m}{F}$, because the size of the matrices are huge, multiple millions for practical examples,
 		reductions to 0 can cost a lot during the linear algebra process.
 F5 \cite{F02} was a major improvement because it provided a criterion that avoided reduction to 0 during the linear algebra part for all regular sequences.
+
+		We will introduce the F5 criterions and the notions of regularity and semi-regularity in the following.
+
+		\begin{proposition}[General criterion \cite{F02}]
+			\label{F5Crit}
+			Let $d_i$ be the total degree of $f_i$.
+			$\forall j < m$, if a row of signature ($u, f_j$) in the matrix $\Mact{D-d_j}{m-1}{F}$ has as leading term $t'$, then the row ($t', f_m$) in $\Mac{D}{m}{F}$ (resp.
+$\Mac{\leq D}{m}{F}$) will reduce to 0 i.e.
+is a linear combination of the predecessors.
+		\end{proposition}
+
+		\begin{proposition}[Frobenius criterion \cite{F02}]
+			\label{Frob}
+			If a row of signature ($t, f_m$) in $\Mact{D - d_m}{m}{F}$ has as leading term $t'$, then the row ($t', f_m$) in $\Mac{D}{m}{F}$ (resp.
+$\Mac{\leq D}{m}{F}$) will reduce to 0 i.e.
+is a linear combination of the predecessors.
+		\end{proposition}
+
 		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $D$ which is the degree of regularity of $I$ which is the ideal generated by the polynomials of the system.
@@
 				\item[-] If the sequence is semi-regular, then no reductions to 0 are performed during the F5-matrix algorithm until the degree of regularity $D - 1$.
-				\item[-] If there is no reductions to 0 during the F5-matrix algorithm until the degree $D - 1$, and if the matrix of degree D is full rank and is the first matrix to have more rows than columns, then the sequence is semi-regular and it's index of regularity is $H(I) = D$.
+				\item[-] If there are no reductions to 0 during the F5-matrix algorithm until the degree $D - 1$, and if the matrix of degree D is full rank and is the first matrix to have more rows than columns, then the sequence is semi-regular and its index of regularity is $H(I) = D$.
 			\end{itemize}
@@
 		Our NSBC instances are quadratic systems that are bilinear in the variables of degree 2, overdetermined and can be semi-regular.
+
+		\begin{definition}
+			Let $E, F, G$ be vector spaces on the same base field $\mathbb{K}$. 
+			Let $f: E \times F \leftarrow G$ an application. We say that $f$ is bilinear if
+			$$
+				\forall (x, x') \in E^2 , \forall (y, y') \in F^2 , \forall \lambda \in \mathbb{K}
+			$$
+			$$
+				f(x + x',y) = f(x, y) + f(x', y)
+			$$
+			$$
+				f(x, y + y') = f(x, y) + f(x, y')
+			$$
+			$$
+				f(\lambda x, y) = \lambda f(x, y)
+			$$
+		\end{definition}
 
@@
 		\begin{itemize}
+			\item[-] Let $\mathbb{K}$ be a field. In our case in cryptography $\mathbb{K}$ is always finite.
 			\item[-] $R = \mathbb{K}[x_1,...,x_{n_x},y_1,...,y_{n_y}]$
-			\item[-] Let $\mathbb{K}$ be a field and $f_1,...,f_m \in R$ be a bilinear polynomial.
-We denote by $F_i$ the polynomial sequence $(f_1,...,f_i)$ and $I_i$ the ideal spanned by this sequence $\langle F_i \rangle$.
+			\item[-] Let $f_1,...,f_m \in R$ be a bilinear polynomial. We denote by $F_i$ the polynomial sequence $(f_1,...,f_i)$ and $I_i$ the ideal spanned by this sequence $\langle F_i \rangle$.
 			\item[-] $jac_{x}(F_{i-1})$ (resp. $jac_{y}(F_{i-1})$) is the jacobian matrix with respect to the two subsets of variables $x_1,...,x_{n_x}$ (resp. $y_1,...,y_{n_y}$).
 			\item[-] Let $M$ be a $l \times c$ matrix with $l > c$. We call maximal minors of $M$ the determinants of the $c \times c$ sub-matrices of $M$. 
-			\item[-] $I_{i-1} : f_i$ is the ideal spanned by $I_{i-1}$ and by the maximal minors of $jac_{x}(F_{i-1})$ if ($i > n_y$) and $jac_{y}(F_{i-1})$ if ($i > n_x$).
+			\item[-] $I_{i-1} : f_i$ is the ideal spanned by $\{g \in R \mid gf_i \in I_{i-1}\}$.
 		\end{itemize}
@@
 
-		The following algorithms will give us the signatures of reductions to 0 during F5-matrix algorithm.
+		The Algorithms \ref{alg:crit_calc} will give us the signatures of reductions to 0 during F5-matrix algorithm.
 
@@
 		 \begin{proposition}
-		 	The complexity of algorithm \ref{alg:crit_calc} is
+		 	The complexity of Algorithm \ref{alg:crit_calc} is
 		 	$$
@@
 
-		 If this case arises, the only computations that would be done are for extended F5 criterion. The complexity for F5-Matrix algorithm with only the Frobenius criterion is 
+		 If this case arises, the only computations that would be done are for extended F5 criterion. The complexity for F5-Matrix algorithm with only the Frobenius criterion \ref{Frob} is 
 		 $$

### 2025-04-04 article_bihom_v2.tex
ajout de def semi-regularité et indice de regularité

@@
 \usepackage[utf8]{inputenc}
-\usepackage{lmodern}
 \usepackage[a4paper]{geometry}
@@
 \usepackage[backend=bibtex]{biblatex}
-\usepackage{csquotes}
 \usepackage{algorithm}
@@
 \usepackage{hyperref}
+\usepackage{lmodern}
+\usepackage{csquotes}
 
@@
 		\subsection{Gröbner basis algorithms}
+		In the following of the paper we will denote $R = \mathbb{K}[x_1,\dots,x_n]$ a polynomial ring on a field $\mathbb{K}$.
+
 		Gröbner basis algorithms are used to find solutions for polynomial systems using algebraic geometry tools. For a gentle introduction to this subject \cite{CLS}.
@@
 		\begin{definition}
-			Let $\prec$ be an admissible monomial ordering on a given ring $R = \mathbb{K}[x_1,\dots,x_n]$ where $\mathbb{K}$ is a field.
+			Let $\prec$ be an admissible monomial ordering on a given ring $R$.
 Given a sequence of homogeneous (resp.
@@
 $\Mac{\leq D}{m}{\mathcal{F}}$), and defined as follows: the columns of the matrix are indexed by the monomials of degree $D$ (resp.
-$\leq D$) sorted in decreasing order from the left to right using the defined monomial ordering.
+$\leq D$) sorted in decreasing order from the left to the right using the defined monomial ordering.
 Each row of the matrix is labeled by a tag, also called signature, ($u, f_i$) where u is a monomial in $R$ and $f_i \in \mathcal{F}$ such that $deg(uf_i) = D$ (resp.
@@
 We denote by $\Mact{D}{m}{F}$ (resp.
-$\Mact{\leq D}{m}{F}$) the Macaulay matrix on which we apply the row echelon form without swapping columns to preserve the monomial ordering.
+$\Mact{\leq D}{m}{F}$) the row echelon form of $\Mac{D}{m}{F}$ (resp.
+$\Mac{\leq D}{m}{F}$) without swapping the columns to mantain the monomial order.
 		\end{definition}
@@
 		\end{proposition}
+
+		\begin{definition}
+			The Hilbert function of a homogeneous ideal $I = \langle f_1,\dots f_m \rangle$ at degree $s$ is defined as
+			$$
+				HF_{s, m, d}(n) = dim(R/I)_s = dim((R)_s/I_s) = dim(R)_s - dim(I)_s
+			$$
+			where $d = (d_1,\dots,d_m)$ and $d_i$ is the degree of the homogeneous polynomial $f_i$.
+		\end{definition}
+
+		The index of regularity is when this function of $s$ is equal to Hilbert's polynomial of $s$.
+The degree of this polynomial is exactly the dimension of the ideal.
+We denote by $H(I)$ the index of regularity of $I$.
+
+		\begin{definition}[Semi-regularity \cite{Bardet04}]
+			Let $R = \mathbb{K}[x_1,\dots,x_n]$ a polynomial ring with $\mathbb{K}$ a field.
+			A homogeneous sequence $f_1,\dots,f_m \in R$ is semi-regular if the following conditions are verified
+			\begin{itemize}
+				\item $I = \langle f_1,\dots,f_m \rangle \neq R$
+				\item For $i \in [1;m]$, if $g_if_i = 0$ in $R/\langle f_1,\dots,f_m \rangle$ and $deg(g_if_i) < H(I)$ then $g_i = 0$ in $R/\langle f_1,\dots,f_m \rangle$.
+			\end{itemize} 
+		\end{definition}
 

### 2025-04-08 article_bihom_v2.tex
small changes
ajout de proposition semi-regularité

--- 2025-04-04 article_bihom_v2.tex
@@
 \usepackage{amsthm}
+\usepackage{amssymb}
 \usepackage{mathtools}
@@
 We denote by $\Mact{D}{m}{F}$ (resp.
-$\Mact{\leq D}{m}{F}$) the Macaulay matrix on which we apply the row echelon form without swapping columns to preserve the monomial ordering.
+$\Mact{\leq D}{m}{F}$) the row echelon form of $\Mac{D}{m}{\mathcal{F}}$ (resp.
+$\Mac{\leq D}{m}{\mathcal{F}}$) without swapping columns to preserve the monomial ordering.
 		\end{definition}
@@
 In our case, $n_x = n_y = (m-1)/2$ and $m = k$.
-As we mentioned earlier, our polynomials systems can be semi-regular, so we get no reductions to 0 during the F5 algorithm using the Frobenius criterion for boolean polynomial systems until the Macaulay matrix of degree $D_{reg}$.
+As we mentioned earlier, our polynomial systems can be semi-regular, so we get no reductions to 0 during the F5 algorithm using the Frobenius criterion for boolean polynomial systems until the Macaulay matrix of degree $D_{reg}$.
 
@@
 
-		The Algorithms \ref{alg:crit_calc} will give us the signatures of reductions to 0 during F5-matrix algorithm.
+		We will also need some properties of semi-regular sequences from \cite{Bardet04}
+
+		\begin{proposition}[Proposition 3.2.5 in \cite{Bardet04}]
+			let $f_1,\dots,f_m$ a semi-regular homogeneous sequence, $f_i$ is of degree $d_i$. Then the following proprieties are verified:
+			\begin{itemize}
+				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if it's Hilbert's series is
+				$$
+					 \prod_{i=1}^{m}(1-z^d_i)/(1-z)^n,
+				$$
+				The series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^d_i)/(1-z)^n$ is called the generating series of $f_1,\dots,f_m$.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if it's Hilbert series is
+				$$
+				(1-z)^n/\prod_{i=1}^{m}(1+z^d_i),
+				$$
+				the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1-z)^n/\prod_{i=1}^{m}(1+z^d_i)$ is called the generating series of $f_1,\dots,f_m$.
+				\item Every permutation $f_{\sigma(1)},\dots,f_{\sigma(m)}$ est a semi-regular sequence.
+				\item When the associated variety is zero-dimensional, $H(I)$ is characterized by
+				$$
+					\forall d < H(I), h_{d, m}(n) > 0 \text{ and } h_{H(I), m}(n) \leq 0.
+				$$
+				\item The sequence $f_1,\dots,f_m$ is semi-regular $\nRightarrow \forall i \in [1;m]$, the sequence $f_1,\dots,f_i$ is semi-regular. 
+			\end{itemize}
+		\end{proposition}
+
+		The Algorithm \ref{alg:crit_calc} will give us the signatures of reductions to 0 during F5-matrix algorithm.
 
@@
 
-		 The following theorem, ensures that there are no reductions to 0 with extended the F5 criterion.
+		 \begin{definition}[Definition 8 in \cite{FSS11}]
+		 	Let $\prec$ be a monomial ordering such that its restriction to $\mathbb{K}[x_1,\dots,x_{n_x}]$ (resp.
+$\mathbb{K}[y_1,\dots,y_{n_y}]$) is the grevlex ordering.
+Let $m \leq n_x + n_y$ and $f_1,\dots,f_m$ be bilinear polynomials or $R$.
+We say that the polynomial sequence ($f_1,\dots,f_m$) is a \textit{bi-regular} sequence if $m = 1$ or if ($f_1,\dots,f_m$) is a \textit{bi-regular} sequence and
+		 	$$
+		 		LM(I_{m-1} : f_m) = \langle Monomials_{m - n_y - 2}^x(n_y + 1) \rangle + \langle Monomials_{m - n_x - 2}^y(n_x + 1) + LM(I_{m-1})\rangle.
+		 	$$
+		 \end{definition}
+
+		 The following theorem, ensures that there are no reductions to 0 with extended the F5 criterion. Ce théorème est en suspens pour utiliser une autre formulation sans utiliser la topologie de Zariski
 
@@
 		 \end{theorem}
-
-		 For a formal definition of a \textit{bi-regular} sequence, we refer to definition 8 of \cite{FSS11}.
 

### 2025-04-09 article_bihom_v2.tex
typo

@@
 			\begin{itemize}
-				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if it's Hilbert's series is
-				$$
-					 \prod_{i=1}^{m}(1-z^d_i)/(1-z)^n,
-				$$
-				The series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^d_i)/(1-z)^n$ is called the generating series of $f_1,\dots,f_m$.
-				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if it's Hilbert series is
-				$$
-				(1-z)^n/\prod_{i=1}^{m}(1+z^d_i),
-				$$
-				the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1-z)^n/\prod_{i=1}^{m}(1+z^d_i)$ is called the generating series of $f_1,\dots,f_m$.
-				\item Every permutation $f_{\sigma(1)},\dots,f_{\sigma(m)}$ est a semi-regular sequence.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if it's Hilbert's series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^d_i)/(1-z)^n$ is called the generating series of $f_1,\dots,f_m$.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if it's Hilbert series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^d_i)$ is called the generating series of $f_1,\dots,f_m$.
+				\item Every permutation $f_{\sigma(1)},\dots,f_{\sigma(m)}$ is a semi-regular sequence.
 				\item When the associated variety is zero-dimensional, $H(I)$ is characterized by

### 2025-04-09 article_bihom_v2.tex
ajout compréhension papier FSS

@@
 
-		\begin{theorem}[Theorem 2 in \cite{FSS11}]
+		\begin{theorem}[Theorem 2 in \cite{FSS11}]\label{theoremMaxMin}
 			Let $i > n_x + 1$ (resp. $i > n_y + 1$) and let $s$ be a linear combination of maximal minors of $jac_{x}(F_{i-1})$ (resp. $jac_{y}(F_{i-1})$). Then $s \in I_{i-1} : f_i$
@@
 		\end{proposition}
+
+		As theorem \ref{theoremMaxMin} indicates, reductions to 0 can also arise from the maximal minors of the jacobian matrix, so the signatures ($s, f_i$) where $s$ is a linear combination of maximal minors of $jac_{x}(F_{i-1})$ or $jac_{y}(F_{i-1})$ is a reduction to 0.
+		This explains Algorithm \ref{alg:crit_calc} where we compute the maximal minors the jacobian matrices before applying Algorithm \ref{alg:reduc} where we compute linear combinations of these maximal minors.
+		This means that Algorithm \ref{alg:crit_calc} computes all the signatures that will arise in reductions to 0 that involve the maximal minors of jacobian matrices instead of being incremental like F5 criterions \cite{F02}.
 

### 2025-04-10 article_bihom_v2.tex
minor changes

@@
 
-		In \cite{Lazard83}, Lazard introduced the connection between linear algebra and Gröbner basis theory via the usage of Macaulay matrices which is defined as follows.
+		In \cite{Lazard83}, Lazard introduced the connection between linear algebra and Gröbner basis theory via the usage of Macaulay matrices \cite{Mac} which is defined as follows.
 
@@
 
-		\begin{proposition}[Proposition 3.2.5 in \cite{Bardet04}]
+		\begin{proposition}[Proposition 3.2.5 in \cite{Bardet04}]\label{PropSemi}
 			let $f_1,\dots,f_m$ a semi-regular homogeneous sequence, $f_i$ is of degree $d_i$. Then the following proprieties are verified:
 			\begin{itemize}
-				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if it's Hilbert's series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^d_i)/(1-z)^n$ is called the generating series of $f_1,\dots,f_m$.
-				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if it's Hilbert series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^d_i)$ is called the generating series of $f_1,\dots,f_m$.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if it's Hilbert's series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^{d_i})/(1-z)^n$ is called the generating series of $f_1,\dots,f_m$.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if it's Hilbert series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^{d_i})$ is called the generating series of $f_1,\dots,f_m$.
 				\item Every permutation $f_{\sigma(1)},\dots,f_{\sigma(m)}$ is a semi-regular sequence.
@@
 
+		Notes pour moi: On a quelque chose de bizarre où si on prends la 2 de la proposition \ref{PropSemi}, on a pas semi-régularité voir rien de particulier.
+Mais si on prends la 1, donc oublier qu'on est sur $\mathbb{F_2}$, on a toujours pas de semi-regularité d'après la proposition, mais la 5 est vérifié avec la série generatrice, donc je soupçonne que la façon de laquelle sagemath calcule la série de hilbert est pas forcément correct, à vérifier avec des exemples de suites semi-regulière.
+
 		As theorem \ref{theoremMaxMin} indicates, reductions to 0 can also arise from the maximal minors of the jacobian matrix, so the signatures ($s, f_i$) where $s$ is a linear combination of maximal minors of $jac_{x}(F_{i-1})$ or $jac_{y}(F_{i-1})$ is a reduction to 0.
-		This explains Algorithm \ref{alg:crit_calc} where we compute the maximal minors the jacobian matrices before applying Algorithm \ref{alg:reduc} where we compute linear combinations of these maximal minors.
+		This explains Algorithm \ref{alg:crit_calc} where we compute the maximal minors of jacobian matrices before applying Algorithm \ref{alg:reduc} where we compute linear combinations of these maximal minors.
 		This means that Algorithm \ref{alg:crit_calc} computes all the signatures that will arise in reductions to 0 that involve the maximal minors of jacobian matrices instead of being incremental like F5 criterions \cite{F02}.
-
-		The Algorithm \ref{alg:crit_calc} will give us the signatures of reductions to 0 during F5-matrix algorithm.
 
@@
 
-		 First, we can clearly see that the complexity of algorithm \ref{alg:crit_calc} is exponential and is what we will try to show.
+		 First, we can clearly see that the complexity of algorithm \ref{alg:crit_calc} is exponential and we will show that in the following proposition \ref{complexCrit}.
 
-		 \begin{proposition}
+		 \begin{proposition}\label{complexCrit}
 		 	The complexity of Algorithm \ref{alg:crit_calc} is

### 2025-04-10 article_bihom_v2.tex
petits ajouts pour moi

@@
 		Notes pour moi: On a quelque chose de bizarre où si on prends la 2 de la proposition \ref{PropSemi}, on a pas semi-régularité voir rien de particulier.
-Mais si on prends la 1, donc oublier qu'on est sur $\mathbb{F_2}$, on a toujours pas de semi-regularité d'après la proposition, mais la 5 est vérifié avec la série generatrice, donc je soupçonne que la façon de laquelle sagemath calcule la série de hilbert est pas forcément correct, à vérifier avec des exemples de suites semi-regulière.
+Mais si on prends la 1, donc oublier qu'on est sur $\mathbb{F}_2$, on a toujours pas de semi-regularité d'après la proposition, mais la 5 est vérifié avec la série generatrice, donc je soupçonne que la façon de laquelle sagemath calcule la série de hilbert est pas forcément correct, à vérifier avec des exemples de suites semi-regulière.
+Update: Il semblerait qu'avec des systèmes aléatoires, donc qui devrait être semi-régulier, sagemath contredit la 5 de \ref{PropSemi}.
+Donc il faut voir si la méthode .hilbert\_series est la bonne ou s'il faut utiliser autre chose.
+Update 2: Sage tronque la série de hilbert jusqu'au degré de régularité très certainement et ça correspond à la série génératrice mais seulement avec la formule en 1 de \ref{PropSemi} et non la 2 ce qui est pas normal vu qu'on est sur $\mathbb{F}_2$.
 

### 2025-04-11 article_bihom_v2.tex
v2 envoyé à Sorina

@@
 
-		The most straightforward cryptanalysis of this type of systems is to perform a Gröbner basis algorithm like Faugère's F5 \cite{F02}.
+		The most straightforward cryptanalysis of this type of systems is to perform a Gröbner basis algorithm like Buchberger's algorithm \cite{Buc} or Faugère's F4 \cite{F99} or F5 \cite{F02} or their multiple variants \cite{EF19}.
 
@@
 
-		We will introduce the F5 criterions and the notions of regularity and semi-regularity in the following.
+		We will introduce the F5 criterions and the notions of semi-regularity in the following.
 
@@
 
+		The Frobenius criterion \ref{Frob} is the analogue of the General criterion \ref{F5Crit} but for polynomial systems defined over $\mathbb{F}_2$.
+
 		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $D$ which is the degree of regularity of $I$ which is the ideal generated by the polynomials of the system.
@@
 
-		However, after some tests and after applying F5-matrix algorithm on polynomial systems generated by NSBC instances, we get reductions to 0, this means that the polynomial systems are not semi-regular when we only multiply by the $x$ variables or by the $y$ variables.
-If we multiply by the $x$ variables and at least one $y$ variable, we get no reductions to 0 which means that the system could be semi-regular.
+		We will also need some properties of semi-regular sequences from \cite{Bardet04}
+
+		\begin{proposition}[Proposition 3.2.5 in \cite{Bardet04}]\label{PropSemi}
+			let $f_1,\dots,f_m$ a semi-regular homogeneous sequence, $f_i$ is of degree $d_i$. Then the following proprieties are verified:
+			\begin{itemize}
+				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if it's Hilbert's series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^{d_i})/(1-z)^n$ is called the generating series of $f_1,\dots,f_m$.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if it's Hilbert series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^{d_i})$ is called the generating series of $f_1,\dots,f_m$.
+				\item Every permutation $f_{\sigma(1)},\dots,f_{\sigma(m)}$ is a semi-regular sequence.
+				\item When the associated variety is zero-dimensional, $H(I)$ is characterized by
+				$$
+				\forall d < H(I), h_{d, m}(n) > 0 \text{ and } h_{H(I), m}(n) \leq 0.
+				$$
+				\item The sequence $f_1,\dots,f_m$ is semi-regular $\nRightarrow \forall i \in [1;m]$, the sequence $f_1,\dots,f_i$ is semi-regular. 
+			\end{itemize}
+		\end{proposition}
+
+		We will also need the next definition of $\gamma$-semi-regularity.
+
+		 \begin{definition}[Definition 4 of \cite{VID24}]
+			Let $\mathcal{F} = {f_1,...,f_m}$ be a semi-regular sequence of polynomials in $\mathbb{F}_2[x_1,...,x_n]$ and let $0 \leq \gamma \leq 1$ such that $k = (1 - \gamma)n$.
+We say that this sequence is $\gamma$-strong semi-regular if
+			$$
+			\mathcal{S}(I) = \{(a_{k+1},...,a_n) \in \mathbb{F}_{2}^{n-k} |
+			$$
+			$$
+			\{f_1(x_1,...,x_k,a_{k+1},...,a_n),...,f_m(x_1,...,x_k,a_{k+1},...,a_n)\} \text{ is not semi-regular}\}
+			$$
+			has cardinality $\mathcal{O}(2^{-\gamma n})$
+		\end{definition}
+
+		When applying Macaulay matrix generation code from hpXbred \cite{BS23} on NSBC polynomial systems, we notice that they are not semi-regular when we only multiply by the $x$ variables or only by the $y$ variables.
+If we multiply by the $x$ variables and at least one $y$ variable, we get no reductions to 0 which means that the system could be semi-regular or $\gamma$-semi-regular.
+
+		Notes pour moi: On a quelque chose de bizarre où si on prends la 2 de la proposition \ref{PropSemi}, on a pas semi-régularité voir rien de particulier.
+Mais si on prends la 1, donc oublier qu'on est sur $\mathbb{F}_2$, on a toujours pas de semi-regularité d'après la proposition, mais la 5 est vérifié avec la série generatrice, donc je soupçonne que la façon de laquelle sagemath calcule la série de hilbert est pas forcément correct, à vérifier avec des exemples de suites semi-regulière.
+Update: Il semblerait qu'avec des systèmes aléatoires, donc qui devrait être semi-régulier, sagemath contredit la 5 de \ref{PropSemi}.
+Donc il faut voir si la méthode .hilbert\_series est la bonne ou s'il faut utiliser autre chose.
+Update 2: Sage tronque la série de hilbert jusqu'au degré de régularité très certainement et ça correspond à la série génératrice mais seulement avec la formule en 1 de \ref{PropSemi} et non la 2 ce qui est pas normal vu qu'on est sur $\mathbb{F}_2$.
+Il est possible que ça soit la méthode de sage qui fait n'importe quoi quand on est sur $\mathbb{F}_2$.
+Après vérification, c'est bien ce que sagemath fait donc il prends pas en compte qu'on est sur $\mathbb{F}_2$.
 
@@
 			Let $E, F, G$ be vector spaces on the same base field $\mathbb{K}$. 
-			Let $f: E \times F \leftarrow G$ an application. We say that $f$ is bilinear if
+			Let $f: E \times F \rightarrow G$ an application. We say that $f$ is bilinear if
 			$$
@@
 
-		The article \cite{FSS11} and the thesis \cite{Spaen2012} give a partial answer to this problem by providing a new criterion for bilinear determined systems and a complexity analysis of their algorithm (The proofs in the end of \cite{FSS11} are not exact, the authors recommended to refer to \cite{Spaen2012} for better proofs.).
+		The article \cite{FSS11} and the thesis \cite{Spaen2012} provide a new criterion for bilinear determined systems and a complexity analysis of their algorithm (the proofs in the end of \cite{FSS11} are not exact, the authors recommended to refer to \cite{Spaen2012} for better proofs) that can help remove reductions to 0 for bilinear systems that are not semi-regular.
 
@@
 		\end{theorem}
-
-		We will also need some properties of semi-regular sequences from \cite{Bardet04}
-
-		\begin{proposition}[Proposition 3.2.5 in \cite{Bardet04}]\label{PropSemi}
-			let $f_1,\dots,f_m$ a semi-regular homogeneous sequence, $f_i$ is of degree $d_i$. Then the following proprieties are verified:
-			\begin{itemize}
-				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if it's Hilbert's series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^{d_i})/(1-z)^n$ is called the generating series of $f_1,\dots,f_m$.
-				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if it's Hilbert series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^{d_i})$ is called the generating series of $f_1,\dots,f_m$.
-				\item Every permutation $f_{\sigma(1)},\dots,f_{\sigma(m)}$ is a semi-regular sequence.
-				\item When the associated variety is zero-dimensional, $H(I)$ is characterized by
-				$$
-					\forall d < H(I), h_{d, m}(n) > 0 \text{ and } h_{H(I), m}(n) \leq 0.
-				$$
-				\item The sequence $f_1,\dots,f_m$ is semi-regular $\nRightarrow \forall i \in [1;m]$, the sequence $f_1,\dots,f_i$ is semi-regular. 
-			\end{itemize}
-		\end{proposition}
-
-		Notes pour moi: On a quelque chose de bizarre où si on prends la 2 de la proposition \ref{PropSemi}, on a pas semi-régularité voir rien de particulier.
-Mais si on prends la 1, donc oublier qu'on est sur $\mathbb{F}_2$, on a toujours pas de semi-regularité d'après la proposition, mais la 5 est vérifié avec la série generatrice, donc je soupçonne que la façon de laquelle sagemath calcule la série de hilbert est pas forcément correct, à vérifier avec des exemples de suites semi-regulière.
-Update: Il semblerait qu'avec des systèmes aléatoires, donc qui devrait être semi-régulier, sagemath contredit la 5 de \ref{PropSemi}.
-Donc il faut voir si la méthode .hilbert\_series est la bonne ou s'il faut utiliser autre chose.
-Update 2: Sage tronque la série de hilbert jusqu'au degré de régularité très certainement et ça correspond à la série génératrice mais seulement avec la formule en 1 de \ref{PropSemi} et non la 2 ce qui est pas normal vu qu'on est sur $\mathbb{F}_2$.
 
@@
 
-		 \begin{definition}[Definition 4 of \cite{VID24}]
-		 	Let $\mathcal{F} = {f_1,...,f_m}$ be a semi-regular sequence of polynomials in $\mathbb{F}_2[x_1,...,x_n]$ and let $0 \leq \gamma \leq 1$ such that $k = (1 - \gamma)n$.
-We say that this sequence is $\gamma$-strong semi-regular if
-		 	$$
-		 	\mathcal{S}(I) = \{(a_{k+1},...,a_n) \in \mathbb{F}_{2}^{n-k} |
-		 	$$
-		 	$$
-		 	\{f_1(x_1,...,x_k,a_{k+1},...,a_n),...,f_m(x_1,...,x_k,a_{k+1},...,a_n)\} \text{ is not semi-regular}\}
-		 	$$
-		 	has cardinality $\mathcal{O}(2^{-\gamma n})$
-		 \end{definition}
-
-		 Note pour moi-même, il faut maintenant essayer de réaliser une implémentation de ce critère en sage ou avec hpXbred et BeanPolE pour voir comment ça se comporterait avec des petits systèmes et voir quel serait l'avantage ou le désavantage sachant que la supposition que toutes les lignes se réduisent à 0 est extrêment peu probable.
+		 Note pour moi-même, il faut maintenant essayer de réaliser une implémentation de ce critère en sage ou avec hpXbred \cite{BS23} et BeanPolE \cite{Bou24} pour voir comment ça se comporterait avec des petits systèmes et voir quel serait l'avantage ou le désavantage sachant que la supposition que toutes les lignes se réduisent à 0 est extrêment peu probable.
+Donc les lignes se réduisent pas à 0 quand on a les x et les y donc le paramètre v de hpXbred est strictement supérieur à $n_x$.
+		 On peut ensuite utiliser XL \cite{CKPS00} avec l'implantation \cite{NNY17}
 

### 2025-04-14 article_bihom_v2.tex
ajout pdf sorina et premier batch de corrections

@@
 
+	\section{Preliminaries and notations}
+		In this section, we introduce some standard notations of the field and notions that we will use.
+
+		Let $\mathbb{F}$ be a finite field and $\mathbb{K}$ a fiend not necessarily finite. 
+		We denote the multivariate polynomial ring defined over a field in $n$ variables $\mathbb{K}[x_1,\dots,x_n] = R$ where $n$ is a positive integer.
+		In the following we will use the standard notation for matrix multiplication complexity for a square matrix of size $n$, $\mathcal{O}(n^\omega)$,
+		where $2 \leq \omega \leq 3$.
+
 	\section{Signature Scheme}
@@
 		\newline
-		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent overs $\mathbb{F}_q$, find two vectors $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})((\vec{v} \cdot \vec{x})$$
+		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent overs $\mathbb{F}_q$, find two vectors $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})$$
+		where $(\cdot)$ is the dot product between 2 vectors.
 
 		\subsection{NSBC}
+
 		The authors also introduce a particular version of this problem called normalized SBC (NSBC) where $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ and
@@
 		One of the known attacks given by the authors is through algebraic cryptanalysis.
-		One can model the system by taking $g$ as the following polynomial
+		One can model the attack by taking $g$ as the following polynomial
 		$$
-		g(x_1,...,x_n,y_1,...,y_n) := (\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) - (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})
+		g(x_1,...,x_n,y_1,...,y_n) := (\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) - (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x}).
 		$$
-		or if we consider the NSBC problem
+		We consider
 		$$
-		g(x_1,..,x_{n-2},y_1,..,y_{n-2}) := (\sum_{i = 1}^{n-2}u_{i}x_{i} + u_{n-1})(\sum_{i = 1}^{n-2}v_{i}y_{i} + v_{n}) - (\sum_{i = 1}^{n-2}u_{i}y_{i} + u_{n})(\sum_{i = 1}^{n-2}v_{i}x_{i} + v_{n-1})
+		g(x_1,..,x_{n-2},y_1,..,y_{n-2}) := (\sum_{i = 1}^{n-2}u_{i}x_{i} + u_{n-1})(\sum_{i = 1}^{n-2}v_{i}y_{i} + v_{n}) - (\sum_{i = 1}^{n-2}u_{i}y_{i} + u_{n})(\sum_{i = 1}^{n-2}v_{i}x_{i} + v_{n-1}).
 		$$
@@
 
-		The most straightforward cryptanalysis of this type of systems is to perform a Gröbner basis algorithm like Buchberger's algorithm \cite{Buc} or Faugère's F4 \cite{F99} or F5 \cite{F02} or their multiple variants \cite{EF19}.
+		where $g_i$ are polynomials with coefficients in $\mathbb{F}_q$.
 
-		The authors cite \cite{FSS11} to argue their parameters choice, we will dive deeper into this article's results in a later section.
+		The most straightforward cryptanalysis of this type of systems is to perform a Gröbner basis algorithm like Buchberger's algorithm \cite{Buc}, Faugère's F4 \cite{F99}, F5 \cite{F02} or their multiple variants \cite{EF19}.
+
+		Huth and Joux cite a complexity result in \cite{FSS11} to argue their parameters choice, we will dive deeper into this article's results in a later section.
 
@@
 		We will introduce the F5 criterions and the notions of semi-regularity in the following.
+
+		\subsubsection{Faugère's F5}
+
+		\subsubsection{Hilbert's series}
 
@@
 			\begin{itemize}
-				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if it's Hilbert's series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^{d_i})/(1-z)^n$ is called the generating series of $f_1,\dots,f_m$.
-				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if it's Hilbert series is the series $S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^{d_i})$ is called the generating series of $f_1,\dots,f_m$.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if its Hilbert's series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^{d_i})/(1-z)^n$$ is called the generating series of $f_1,\dots,f_m$.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if its Hilbert series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^{d_i})$$ is called the generating series of $f_1,\dots,f_m$.
 				\item Every permutation $f_{\sigma(1)},\dots,f_{\sigma(m)}$ is a semi-regular sequence.
@@
 
-		We will note $n_x$ (resp.
+		A bilinear system is a polynomial system where each equation is a bilinear form.
+
+		We will denote $n_x$ (resp.
 $n_y$) the number of variables in $x$ (resp.
-number of variables in y).
+number of variables in $y$).
 In our case, $n_x = n_y = (m-1)/2$ and $m = k$.
@@
 
-		The article \cite{FSS11} and the thesis \cite{Spaen2012} provide a new criterion for bilinear determined systems and a complexity analysis of their algorithm (the proofs in the end of \cite{FSS11} are not exact, the authors recommended to refer to \cite{Spaen2012} for better proofs) that can help remove reductions to 0 for bilinear systems that are not semi-regular.
+		The article \cite{FSS11} and the thesis \cite{Spaen2012} provide a new criterion for bilinear determined systems and a complexity analysis of their algorithm (the proofs in the end of \cite{FSS11} are not exact, the authors recommended to refer to \cite{Spaen2012} for better proofs) that can help more remove reductions to 0 in F5 Algorithm for bilinear systems which are not semi-regular.
 
@@
 			\item[-] Let $M$ be a $l \times c$ matrix with $l > c$. We call maximal minors of $M$ the determinants of the $c \times c$ sub-matrices of $M$. 
-			\item[-] $I_{i-1} : f_i$ is the ideal spanned by $\{g \in R \mid gf_i \in I_{i-1}\}$.
+			\item[-] $I_{i-1} : f_i$ is the ideal spanned by $\{g \in R \mid gf_i \in I_{i-1}\}$. We denote the maximal minors of $M$ by $MaxMinors(M)$.
 		\end{itemize}
@@
 
-		As theorem \ref{theoremMaxMin} indicates, reductions to 0 can also arise from the maximal minors of the jacobian matrix, so the signatures ($s, f_i$) where $s$ is a linear combination of maximal minors of $jac_{x}(F_{i-1})$ or $jac_{y}(F_{i-1})$ is a reduction to 0.
+		As Theorem \ref{theoremMaxMin} indicates, reductions to 0 may also arise from the maximal minors of the jacobian matrix, so the signatures ($s, f_i$) where $s$ is a linear combination of maximal minors of $jac_{x}(F_{i-1})$ or $jac_{y}(F_{i-1})$ is a reduction to 0.
 		This explains Algorithm \ref{alg:crit_calc} where we compute the maximal minors of jacobian matrices before applying Algorithm \ref{alg:reduc} where we compute linear combinations of these maximal minors.
@@
 				\Ensure $T$ a reduced set of homogeneous polynomials of degree $q$.
-				\State $M \gets Macaulay_{\prec}(S, q)$.
+				\State $M \gets Mac_{\prec}(S, q)$.
 				\State $M \gets RowEchelonForm(M)$.
@@
 		 \begin{algorithm}
-		 	\caption{BilinF5Critetion}\label{algo:crit}
+		 	\caption{BilinF5Criterion}\label{algo:crit}
 		 	\begin{algorithmic}[1]
@@
 		 The arithmetic complexity of computing a Gröbner basis of a generic bilinear system $f_1,...,f_{n_x + n_y}$ with the F5 algorithm is bounded by
-		 $$\mathcal{O}\left(\binom{n_x + n_y + min(n_x + 1, n_y + 1)}{min(n_x + 1, n_y + 1)}^\omega\right)$$
-		 where $2 \leq \omega \leq 3$ is the linear algebra constant.
+		 $$\mathcal{O}\left(\binom{n_x + n_y + min(n_x + 1, n_y + 1)}{min(n_x + 1, n_y + 1)}^\omega\right)$$.
 		 \end{corollary}
 
-		 First, we can clearly see that the complexity of algorithm \ref{alg:crit_calc} is exponential and we will show that in the following proposition \ref{complexCrit}.
+		 First, we can clearly see that the complexity of Algorithm \ref{alg:crit_calc} is exponential and we will show that in the following proposition/.
 
@@
 		 	$$
-		 	\mathcal{O}\left(\left(\sum_{i = n_x}^{m} \binom{i}{n_x}n_x^{\omega} + \binom{2n_x + 1}{n_x}^\omega\right) + \left(\sum_{i = n_y}^{m} \binom{i}{n_y}n_y^{\omega} + \binom{2n_y + 1}{n_y}^\omega\right)\right)
+		 	\mathcal{O}\left(\left(\sum_{i = n_x}^{m} \binom{i}{n_x}n_x^{\omega} + \binom{2n_x + 1}{n_x}^\omega\right) + \left(\sum_{i = n_y}^{m} \binom{i}{n_y}n_y^{\omega} + \binom{2n_y + 1}{n_y}^\omega\right)\right).
 		 	$$
@@
 $jac_x(F_{i-1})$) which is of size $n_y \times n_y$ (resp.
-$n_x \times n_x$) for the first one and $(m-1) \times n_y$ (resp.
-$(m-1) \times n_x$) for the last one.
+$n_x \times n_x$) for the first one and $m-1 \times n_y$ (resp.
+$m-1 \times n_x$) for the last one i.e.
+$i = m$.
 We suppose here that the construction of such matrices is free.
@@
 		 	The size of Macauly matrix is bounded by $\binom{n + D}{D}$ where D is the degree and n the number variables.
-So the complexity of \textbf{Reduce} is $\mathcal{O}\left(\binom{n_y + n_y + 1}{n_y + 1}^\omega \right)$ (resp.
-$\mathcal{O}\left(\binom{n_x + n_x + 1}{n_x + 1}^\omega \right)$).
+So the complexity of \textbf{Reduce} is $$\mathcal{O}\left(\binom{n_y + n_y + 1}{n_y + 1}^\omega \right)$$ (resp.
+$$\mathcal{O}\left(\binom{n_x + n_x + 1}{n_x + 1}^\omega \right)$$).
 Hence the second term in the sum, because we call \textbf{Reduce} at each iteration of the for loop.
-We finally suppose that the iteration on the elements of $T$ and of $T'$ is negligible in the big O notation.
+We finally suppose that the iteration on the elements of the output of \textbf{Reduce} is negligible in the big $\mathcal{O}$ notation.
 		 \end{proof}

### 2025-04-15 article_bihom_v2.tex
deuxième batch et rearrangement

@@
 \newtheorem{corollary}{Corollary}[section]
+\newtheorem{example}{Example}[section]
 
@@
 
+		\subsection{Gröbner basis theory}
+		Gröbner basis theory was first introduced by Buchberger in his thesis \cite{Buc} and he gave an algorithm to compute the Gröbner basis of an ideal.
+		Gröbner basis algorithms like Buchberger's Algorithm \cite{Buc}, are used to find solutions for polynomial systems using algebraic geometry tools.
+For a gentle introduction to this subject \cite{CLS}.
+
+		In \cite{Lazard83}, Lazard introduced the connection between linear algebra and Gröbner basis theory via the usage of Macaulay matrices \cite{Mac} which is defined as follows.
+
+		\begin{definition}
+			Let $\prec$ be an admissible monomial ordering on a given ring $R = \mathbb{K}[x_1,\dots,x_n]$ where $\mathbb{K}$ is a field.
+Given a sequence of homogeneous (resp.
+affine) polynomials $\mathcal{F} = (f_1,\dots,f_m) \in R$, we associate to it the Macaulay matrix of degree $D$ (resp.
+$\leq D$), denoted by $\Mac{D}{m}{\mathcal{F}}$ (resp.
+$\Mac{\leq D}{m}{\mathcal{F}}$), and defined as follows: the columns of the matrix are indexed by the monomials of degree $D$ (resp.
+$\leq D$) sorted in decreasing order from the left to right using the defined monomial ordering.
+Each row of the matrix is labeled by a tag, also called signature, ($u, f_i$) where u is a monomial in $R$ and $f_i \in \mathcal{F}$ such that $deg(uf_i) = D$ (resp.
+$deg(uf_i) \leq D$), and the polynomial $uf_i$ is written as vector of coefficients of monomials on the row.
+We denote by $\Mact{D}{m}{F}$ (resp.
+$\Mact{\leq D}{m}{F}$) the row echelon form of $\Mac{D}{m}{\mathcal{F}}$ (resp.
+$\Mac{\leq D}{m}{\mathcal{F}}$) without swapping columns to preserve the monomial ordering.
+		\end{definition}
+
+		\subsection{Faugère's F5}
+
+		The major "issue" of Gröbner basis algorithms is the reductions to 0 in the Macaulay matrix associated to a polynomial sequence $\Mac{D}{m}{F}$, because the size of the matrices are huge, multiple millions for practical examples,
+		reductions to 0 can cost a lot during the linear algebra process.
+F5 \cite{F02} was a major improvement because it provided a criterion that avoided reduction to 0 during the linear algebra part for all regular sequences.
+
+		\begin{proposition}[General criterion \cite{F02}]
+			\label{F5Crit}
+			Let $d_i$ be the total degree of $f_i$.
+			$\forall j < m$, if a row of signature ($u, f_j$) in the matrix $\Mact{D-d_j}{m-1}{F}$ has as leading term $t'$, then the row ($t', f_m$) in $\Mac{D}{m}{F}$ (resp.
+$\Mac{\leq D}{m}{F}$) will reduce to 0 i.e.
+is a linear combination of the predecessors.
+		\end{proposition}
+
+		\begin{proposition}[Frobenius criterion \cite{F02}]
+			\label{Frob}
+			If a row of signature ($t, f_m$) in $\Mact{D - d_m}{m}{F}$ has as leading term $t'$, then the row ($t', f_m$) in $\Mac{D}{m}{F}$ (resp.
+$\Mac{\leq D}{m}{F}$) will reduce to 0 i.e.
+is a linear combination of the predecessors.
+		\end{proposition}
+
+		The Frobenius criterion \ref{Frob} is the analogue of the General criterion \ref{F5Crit} but for polynomial systems defined over $\mathbb{F}_2$.
+		F5 crtierions remove the rows corresponding to the trivial syzygies, i.e.
+the syzygies $(s_1, \dots, s_m)$ such that for a row with signature $(s_i, f_i)$, we get $s_i \in \langle f_1,\dots,f_{i-1},f_{i+1},\dots,f_m \rangle$.
+
+		Bardet introduced a variant of F5 Algorithm that is simple to understand and analyze called Matrix F5 in \cite{Bardet04}.
+		We give the description of Matrix F5 Algorithm.
+
+		\begin{algorithm}
+			\caption{Matrix F5}\label{alg:matrix_f5}
+			\begin{algorithmic}[1]
+				\Require 
+				\begin{tabular}[t]{@{}l@{}}
+					$\left\{ 
+					\begin{array}{l}
+						m \text{ homogeneous polynomials } f_1, \dots, f_m \text{ of degree } d_1 \leq d_2 \leq \dots \leq d_m, \\
+						D \text{ an integer}, \\
+						\text{a monomial ordering } \prec
+					\end{array}
+					\right.$
+				\end{tabular}
+				\Ensure 
+				\begin{tabular}[t]{@{}l@{}}
+					$\left\{ 
+					\begin{array}{l}
+						G \text{ is a Gröbner basis of } \langle f_1,\dots,f_m \rangle.
+					\end{array}
+					\right.$
+				\end{tabular}
+				\State $G \gets {f_1,\dots,f_m}$
+				\For{$d$ from $d_1$ to $D$}
+					\State $\widetilde{Mac}_{d, 0} \gets$ matrix with 0 rows
+					\For{$i$ from $1$ to $m$}
+						\State Construct ${Mac}_{d, i}$ by adding to $\widetilde{Mac}_{d, i-1}$ the following rows:
+						\If{$d_i = d$}
+							\State add the row $f_i$ with signature $(1, f_i)$
+						\EndIf
+						\If{$d > d_i$}
+							for all $f$ from $\widetilde{Mac}_{d, i-1}$ with signature $(e, f_i)$, such that $x_\lambda$ is the greatest variable of $e$, add the $n - \lambda + 1$ rows $x_\lambda f,x_{\lambda + 1}f,\dots,x_n f$ with the signatures $(x_\lambda e, f),(x_{\lambda + 1}e, f),\dots,(x_n e, f)$ except those who satisfy the F5 criterion with $((x_{\lambda + k}e, fi), \widetilde{Mac}_{d-d_i, i-1}$
+						\EndIf
+						\State Compute $\widetilde{Mac}_{d, i}$ the row echelon form of ${Mac}_{d, i}$
+						\State Add to $G$ the polynomials corresponding to rows of $\widetilde{Mac}_{d, i}$ such that their leading
+						\State monomials is different from the leading monomial of the row with 
+						\State same signature in ${Mac}_{d, i}$
+				 	\EndFor
+				\EndFor
+				\State \textbf{Return} $G$
+			\end{algorithmic}
+		\end{algorithm}
+
+		\subsection{Hilbert's series}
+
 	\section{Signature Scheme}
@@
 		where $(\cdot)$ is the dot product between 2 vectors. 
+		We denote an instance of the SBC problem given by two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$ by $\text{SBC}[\vec{u}, \vec{v}]$.
+If the vectors $\vec{x},\vec{y} \in (\mathbb{F}_q)^n$ are solution of $\text{SBC}[\vec{u},\vec{v}]$, we use the notation $(\vec{x}, \vec{y}) \in \text{SBC}[\vec{u}, \vec{v}]$.
 
@@
 
-		\subsection{Gröbner basis algorithms}
-		Gröbner basis algorithms are used to find solutions for polynomial systems using algebraic geometry tools. For a gentle introduction to this subject \cite{CLS}.
-
-		In \cite{Lazard83}, Lazard introduced the connection between linear algebra and Gröbner basis theory via the usage of Macaulay matrices \cite{Mac} which is defined as follows.
-
-		\begin{definition}
-			Let $\prec$ be an admissible monomial ordering on a given ring $R = \mathbb{K}[x_1,\dots,x_n]$ where $\mathbb{K}$ is a field.
-Given a sequence of homogeneous (resp.
-affine) polynomials $\mathcal{F} = (f_1,\dots,f_m) \in R$, we associate to it the Macaulay matrix of degree $D$ (resp.
-$\leq D$), denoted by $\Mac{D}{m}{\mathcal{F}}$ (resp.
-$\Mac{\leq D}{m}{\mathcal{F}}$), and defined as follows: the columns of the matrix are indexed by the monomials of degree $D$ (resp.
-$\leq D$) sorted in decreasing order from the left to right using the defined monomial ordering.
-Each row of the matrix is labeled by a tag, also called signature, ($u, f_i$) where u is a monomial in $R$ and $f_i \in \mathcal{F}$ such that $deg(uf_i) = D$ (resp.
-$deg(uf_i) \leq D$), and the polynomial $uf_i$ is written as vector of coefficients of monomials on the row.
-We denote by $\Mact{D}{m}{F}$ (resp.
-$\Mact{\leq D}{m}{F}$) the row echelon form of $\Mac{D}{m}{\mathcal{F}}$ (resp.
-$\Mac{\leq D}{m}{\mathcal{F}}$) without swapping columns to preserve the monomial ordering.
-		\end{definition}
-
-		The major "issue" of Gröbner basis algorithms is the reductions to 0 in the Macaulay matrix associated to a polynomial sequence $\Mac{D}{m}{F}$, because the size of the matrices are huge, multiple millions for practical examples,
-		reductions to 0 can cost a lot during the linear algebra process.
-F5 \cite{F02} was a major improvement because it provided a criterion that avoided reduction to 0 during the linear algebra part for all regular sequences.
-
-		We will introduce the F5 criterions and the notions of semi-regularity in the following.
-
-		\subsubsection{Faugère's F5}
-
-		\subsubsection{Hilbert's series}
-
-		\begin{proposition}[General criterion \cite{F02}]
-			\label{F5Crit}
-			Let $d_i$ be the total degree of $f_i$.
-			$\forall j < m$, if a row of signature ($u, f_j$) in the matrix $\Mact{D-d_j}{m-1}{F}$ has as leading term $t'$, then the row ($t', f_m$) in $\Mac{D}{m}{F}$ (resp.
-$\Mac{\leq D}{m}{F}$) will reduce to 0 i.e.
-is a linear combination of the predecessors.
-		\end{proposition}
-
-		\begin{proposition}[Frobenius criterion \cite{F02}]
-			\label{Frob}
-			If a row of signature ($t, f_m$) in $\Mact{D - d_m}{m}{F}$ has as leading term $t'$, then the row ($t', f_m$) in $\Mac{D}{m}{F}$ (resp.
-$\Mac{\leq D}{m}{F}$) will reduce to 0 i.e.
-is a linear combination of the predecessors.
-		\end{proposition}
-
-		The Frobenius criterion \ref{Frob} is the analogue of the General criterion \ref{F5Crit} but for polynomial systems defined over $\mathbb{F}_2$.
+		\subsection{Algebraic cryptanalysis}
 
@@
 			$$
-				\forall (x, x') \in E^2 , \forall (y, y') \in F^2 , \forall \lambda \in \mathbb{K}
+				\forall (x, x') \in E^2 , \forall (y, y') \in F^2 , \forall \lambda, \mu \in \mathbb{K}
 			$$
 			$$
-				f(x + x',y) = f(x, y) + f(x', y)
+				f(x + x',y) = f(x, y) + f(x', y),
 			$$
 			$$
-				f(x, y + y') = f(x, y) + f(x, y')
+				f(x, y + y') = f(x, y) + f(x, y'),
 			$$
 			$$
-				f(\lambda x, y) = \lambda f(x, y)
+				f(\lambda x, \mu y) = \lambda \mu f(x, y).
 			$$

### 2025-04-18 article_bihom_v2.tex
ajouts biblio

@@
 
+		\subsection{Boolean systems}
+
 	\section{Signature Scheme}
@@
 			\begin{itemize}
-				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if its Hilbert's series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^{d_i})/(1-z)^n$$ is called the generating series of $f_1,\dots,f_m$.
-				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if its Hilbert series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^{d_i})$$ is called the generating series of $f_1,\dots,f_m$.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if its Hilbert's series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^{d_i})/(1-z)^n.$$ $S_{m, n}(z)$ is called the generating series of $f_1,\dots,f_m$.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if its Hilbert series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^{d_i}).$$ $S_{m, n}(z)$ is called the generating series of $f_1,\dots,f_m$.
 				\item Every permutation $f_{\sigma(1)},\dots,f_{\sigma(m)}$ is a semi-regular sequence.
@@
 			$$
-			has cardinality $\mathcal{O}(2^{-\gamma n})$
+			has cardinality $\mathcal{O}(2^{-\gamma n})$.
 		\end{definition}
@@
 		\begin{definition}
-			Let $E, F, G$ be vector spaces on the same base field $\mathbb{K}$. 
-			Let $f: E \times F \rightarrow G$ an application. We say that $f$ is bilinear if
+			A homogeneous polynomial $f \in \mathbb{K}[x_1,\dots,x_{n_x},y_1,\dots,y_{n_y}]$ is called bilinear if we have:
 			$$
-				\forall (x, x') \in E^2 , \forall (y, y') \in F^2 , \forall \lambda, \mu \in \mathbb{K}
-			$$
-			$$
-				f(x + x',y) = f(x, y) + f(x', y),
-			$$
-			$$
-				f(x, y + y') = f(x, y) + f(x, y'),
-			$$
-			$$
-				f(\lambda x, \mu y) = \lambda \mu f(x, y).
+				\forall \lambda, \mu \in \mathbb{K}, f(\lambda x_1,\dots,\lambda x_{n_x}, \mu y_1,\dots, \mu y_{n_y}) = \lambda \mu f(x_1,\dots,x_{n_x},y_1,\dots,y_{n_y}).
 			$$
@@
 		\begin{itemize}
-			\item[-] Let $\mathbb{K}$ be a field. In our case in cryptography $\mathbb{K}$ is always finite.
-			\item[-] $R = \mathbb{K}[x_1,...,x_{n_x},y_1,...,y_{n_y}]$
 			\item[-] Let $f_1,...,f_m \in R$ be a bilinear polynomial. We denote by $F_i$ the polynomial sequence $(f_1,...,f_i)$ and $I_i$ the ideal spanned by this sequence $\langle F_i \rangle$.
 			\item[-] $jac_{x}(F_{i-1})$ (resp. $jac_{y}(F_{i-1})$) is the jacobian matrix with respect to the two subsets of variables $x_1,...,x_{n_x}$ (resp. $y_1,...,y_{n_y}$).
-			\item[-] Let $M$ be a $l \times c$ matrix with $l > c$. We call maximal minors of $M$ the determinants of the $c \times c$ sub-matrices of $M$. 
-			\item[-] $I_{i-1} : f_i$ is the ideal spanned by $\{g \in R \mid gf_i \in I_{i-1}\}$. We denote the maximal minors of $M$ by $MaxMinors(M)$.
+			\item[-] Let $M$ be a $l \times c$ matrix with $l > c$.
+We call maximal minors of $M$ the determinants of the $c \times c$ sub-matrices of $M$.
+We denote the maximal minors of $M$ by $MaxMinors(M)$.
+			\item[-] $I_{i-1} : f_i$ is the ideal spanned by $\{g \in R \mid gf_i \in I_{i-1}\}$.
 		\end{itemize}
@@
 		\begin{theorem}[Theorem 2 in \cite{FSS11}]\label{theoremMaxMin}
-			Let $i > n_x + 1$ (resp. $i > n_y + 1$) and let $s$ be a linear combination of maximal minors of $jac_{x}(F_{i-1})$ (resp. $jac_{y}(F_{i-1})$). Then $s \in I_{i-1} : f_i$
+			Let $i > n_x + 1$ (resp. $i > n_y + 1$) and let $s$ be a linear combination of maximal minors of $jac_{x}(F_{i-1})$ (resp. $jac_{y}(F_{i-1})$). Then $s \in I_{i-1} : f_i$.
 		\end{theorem}
@@
 		 The arithmetic complexity of computing a Gröbner basis of a generic bilinear system $f_1,...,f_{n_x + n_y}$ with the F5 algorithm is bounded by
-		 $$\mathcal{O}\left(\binom{n_x + n_y + min(n_x + 1, n_y + 1)}{min(n_x + 1, n_y + 1)}^\omega\right)$$.
+		 $$\mathcal{O}\left(\binom{n_x + n_y + min(n_x + 1, n_y + 1)}{min(n_x + 1, n_y + 1)}^\omega\right).$$
 		 \end{corollary}
 
-		 First, we can clearly see that the complexity of Algorithm \ref{alg:crit_calc} is exponential and we will show that in the following proposition/.
+		 First, we can clearly see that the complexity of Algorithm \ref{alg:crit_calc} is exponential and we will show that in the following proposition.
 
@@
 		 $$
-		 \mathcal{O}\left(2n_x^{\omega}\left(\sum_{i = n_x}^{m} \binom{i}{n_x} + \binom{2n_x + 1}{n_x}^\omega\right)\right)
+		 \mathcal{O}\left(2n_x^{\omega}\left(\sum_{i = n_x}^{m} \binom{i}{n_x} + \binom{2n_x + 1}{n_x}^\omega\right)\right).
 		 $$
@@
 		 $$
-		 \mathcal{O}\left(\binom{n + 4}{4}^\omega\right)
+		 \mathcal{O}\left(\binom{n + 4}{4}^\omega\right).
 		 $$
@@
 		 \begin{equation}
-		 Adv(n, m) = 2\left(\sum_{i = n/2}^{m}\binom{i}{n/2}\left(\frac{n}{2}\right)^\omega + \binom{n + 1}{n/2}^\omega\right) - \binom{n + 4}{4}^\omega
+		 Adv(n, m) = 2\left(\sum_{i = n/2}^{m}\binom{i}{n/2}\left(\frac{n}{2}\right)^\omega + \binom{n + 1}{n/2}^\omega\right) - \binom{n + 4}{4}^\omega.
 		 \label{equation_complex}
@@
 
-		 \subsection{Boolean systems and Crossbred}
-
 \printbibliography

### 2025-05-07 article_bihom_v2.tex
ajout de partie sur series de hilbert

@@
 
-		Let $\mathbb{F}$ be a finite field and $\mathbb{K}$ a fiend not necessarily finite. 
-		We denote the multivariate polynomial ring defined over a field in $n$ variables $\mathbb{K}[x_1,\dots,x_n] = R$ where $n$ is a positive integer.
-		In the following we will use the standard notation for matrix multiplication complexity for a square matrix of size $n$, $\mathcal{O}(n^\omega)$,
+		\subsubsection{Notations}
+
+		\begin{itemize}
+			\item Let $\mathbb{F}$ be a finite field and $\mathbb{K}$ a field not necessarily finite. 
+			\item We denote the multivariate polynomial ring defined over a field in $n$ variables $\mathbb{K}[x_1,\dots,x_n] = R$ where $n$ is a positive integer.
+			\item We will use the standard notation for matrix multiplication complexity for a square matrix of size $n$, $\mathcal{O}(n^\omega)$,
 			where $2 \leq \omega \leq 3$.
+			\item Let $R_{\leq s}$ be the polynomials of total degree $\leq s$. $I_{\leq s} = I \cap R_{\leq s}$.
+		\end{itemize}
 
@@
 
-		\subsection{Hilbert's series}
+		\subsection{Hilbert series}
+		Hilbert-Poincaré series (or just Hilbert series) is a very important tool in algebraic geometry to describe ideals and varieties and their dimensions.
+		\begin{definition}[\cite{CLS}]
+			Let $I$ be an ideal in $R$. The affine Hilbert function of $I$ is the function on the nonnegative integers s defined by
+			$$
+				{}^a HF_{R/I}(s) = dim R_{\leq s} / dim I_{\leq s} = dim R_{\leq s} - dim I_{\leq s}.
+			$$
+		\end{definition}
+
+		\begin{proposition}
+			${}^a HF_{R/I}(s)$ is the number of monomials not in $I$ of total degree $s$.
+		\end{proposition}
+		\begin{proof}
+			First note that the monomials ${x^\alpha | \alpha \leq s}$ is a basis of $R_{\leq s}$ as a vector space.
+We also have that ${x^\alpha | \alpha \leq s | x^\alpha \in I}$ is a basis of $I_{\leq s}$.
+Therefore, the monomials ${x^\alpha | \alpha \leq s | x^\alpha \notin I}$ are the ones missing to form a basis for $R$, so they form a basis for $R_{\leq s} / I_{\leq s}$ which concludes the proof.
+		\end{proof}
+
+		\begin{definition}
+			Let I be a homogeneous ideal. The Hilbert series is defined as follows
+			$$
+					HS_{R/I}(t) = \sum_{m=0}^{\infty} HF_{R/I}(m) t^m.
+			$$	
+		\end{definition}
+
+		The smallest integer $s_0$ such that $HF_{R/I}(s_0) \leq 0$ is called the degree of regularity of the ideal $I$.
+
+		Faugère showed that no reductions to 0 occur in the Macaulay matrix during the F5 algorithm \ref{alg:matrix_f5} until the degree of regularity \cite{F02}.
+
+		\subsection{Semi-regularity}
 
@@
 		\newline
-		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent overs $\mathbb{F}_q$, find two vectors $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})$$
+		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent overs $\mathbb{F}_q$, find two non-colinear vectors $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})$$
 		where $(\cdot)$ is the dot product between 2 vectors.
@@
 		The authors also introduce a particular version of this problem called normalized SBC (NSBC) where $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ and
-		$\vec{x} = (\vec{x'}, 1, 0), \vec{y} = (\vec{y'}, 0, 1)$.
+		$\vec{x} = (\vec{x'}, 1, 0), \vec{y} = (\vec{y'}, 0, 1)$ to ensure that they are not colinear without needing to regenerate each time they are not.
 

### 2025-05-13 article_bihom_v2.tex
fini regularité ?

@@
 			where $2 \leq \omega \leq 3$.
-			\item Let $R_{\leq s}$ be the polynomials of total degree $\leq s$. $I_{\leq s} = I \cap R_{\leq s}$.
+			\item Let $R_{s}$ be the polynomials of total degree $s$. $I_{s} = I \cap R_{s}$.
 		\end{itemize}
@@
 		\begin{definition}[\cite{CLS}]
-			Let $I$ be an ideal in $R$. The affine Hilbert function of $I$ is the function on the nonnegative integers s defined by
+			Let $I$ be a homogeneous ideal in $R$. The affine Hilbert function of $I$ is the function on the nonnegative integers s defined by
 			$$
-				{}^a HF_{R/I}(s) = dim R_{\leq s} / dim I_{\leq s} = dim R_{\leq s} - dim I_{\leq s}.
+				{}^a HF_{R/I}(s) = dim R_{s} / dim I_{s} = dim R_{s} - dim I_{s}.
 			$$
@@
 
-		The smallest integer $s_0$ such that $HF_{R/I}(s_0) \leq 0$ is called the degree of regularity of the ideal $I$.
+		The index of regularity is the smallest integer $s_0$ such that $HF_{R/I}(s_0)$ is equal to a certain polynomial called the Hilbert polynomial.
+We will denote $H(I)$ as the index or degree of regularity of I.
 

### 2025-05-13 rapport.tex
ajout semi-regularity, crossbred et corrections

--- 2025-05-13 article_bihom_v2.tex
@@
 
-\title{Ce que j'ai compris jusqu'à maintenant\\Début de rapport ?}
+\title{Cryptanalysis of post-quantum signature scheme while breaking dlog}
 
@@
 		The index of regularity is the smallest integer $s_0$ such that $HF_{R/I}(s_0)$ is equal to a certain polynomial called the Hilbert polynomial.
+It is also the smallest integer $s_0$ such that the coefficient of the Hilbert series is less or equal to 0.
 We will denote $H(I)$ as the index or degree of regularity of I.
@@
 		\subsection{Semi-regularity}
+		One problem with the definition of a regular sequence, is that it applies for only underdetermined  polynomial systems because on of the proprieties of a regular sequence of a polynomial system $f_1,\dots,f_m \in \mathbb{K}[x_1,\dots,x_n]$ is that the dimension of the ideal $\langle f_1,\dots,f_m \rangle$ is $n - m$.
+This propriety can not be verified for overdetermined systems because the dimension of the ideal can not be negative.
+
+		Thus, Bardet came with a new concept called semi-regularity for overdetermined system to allow their study and analysis. The definition of a semi-regular sequence is given by Bardet \cite{Bardet04}.
+
+		\begin{definition}
+			A homogeneous sequence $f_1,\dots,f_m \subset R$ is semi-regular if the following conditions are verified:
+			\begin{itemize}
+				\item $I = \langle f_1,\dots,f_m \rangle \neq R$,
+				\item For $i \in [1:m]$, if $g_if_i = 0$ in $R/\langle f_1,\dots,f_m \rangle$ and $deg(g_if_i) < H(I)$ then $g_i = 0$ in $R/\langle f_1,\dots,f_m \rangle$.
+			\end{itemize} 
+		\end{definition}
+
+		Bardet also showed that if we apply matrix-F5 \ref{alg:matrix_f5} to a semi-regular system, there are no reductions to 0 until the degree $d = H(I) - 1$. We also have a certain reciprocity.
+
+		One interesting thing shown by Bardet \cite{Bardet04}, is that a random polynomial system is very likely to be semi-regular.
+Most of the cryptosystems using random polynomial systems assume the semi-regularity of the system to ease their analysis using well established tools.
+For example, one can compute the Hilbert series of a semi-regular system easily.
+
+		\begin{proposition}[Proposition 3.2.5 in \cite{Bardet04}]\label{PropSemi}
+			let $f_1,\dots,f_m$ a semi-regular homogeneous sequence, $f_i$ is of degree $d_i$. Then the following proprieties are verified:
+			\begin{itemize}
+				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if its Hilbert's series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^{d_i})/(1-z)^n.$$ $S_{m, n}(z)$ is called the generating series of $f_1,\dots,f_m$.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if its Hilbert series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^{d_i}).$$ $S_{m, n}(z)$ is called the generating series of $f_1,\dots,f_m$.
+				\item Every permutation $f_{\sigma(1)},\dots,f_{\sigma(m)}$ is a semi-regular sequence.
+				\item When the associated variety is zero-dimensional, $H(I)$ is characterized by
+				$$
+				\forall d < H(I), h_{d, m}(n) > 0 \text{ and } h_{H(I), m}(n) \leq 0.
+				$$
+				\item The sequence $f_1,\dots,f_m$ is semi-regular $\nRightarrow \forall i \in [1;m]$, the sequence $f_1,\dots,f_i$ is semi-regular. 
+			\end{itemize}
+		\end{proposition}
 
 		\subsection{Boolean systems}
+
+		The problem of finding solutions to a multivariate polynomial system over a finite field $\mathbb{F}_p$ is proven NP-complete in \cite{FY79}.
+It is also well known that solving a quadratic polynomial system over any finite-field is NP-hard.
+This does not mean that any random instance is hard to solve.
+The problem of the MQ problem is that some instances can be very easy to solve, so the goal of cryptographers is to imagine a system with parameters that can be hard to solve in general and not have easy cases to solve most of the time.
+
+		Sometimes, the finite field used is $\mathbb{F}_2$ because it is still supposedly hard to solve but is very easy to use and manipulate on computers.
+But in such small fields one can begin to use exhaustive search to obtain the solution to the system.
+The best algorithm known to the authors for exhaustive search for multivariate polynomial system $f_1,\dots,f_m$ over $\mathbb{F}_2[x_1,\dots,x_n]$ is \cite{BCC+10} with a complexity $\mathcal{O}(ln(n)2^n)$, what is remarkable is that the complexity does not implicate the number of polynomials in the system, $m$.
+There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, libFES.
+
+		Another type of algorithms uses exhaustive search combined with probabilistic algorithms.
+\cite{LPT+17} gave an algorithm that finds the solution to a polynomial system with complexity $\tilde{\mathcal{O}}(2^{0.8765n})$ which is better than exhaustive search.
+Even though it's asymptotically better, there are no known implementations because of the huge constant factors present in the complexity that makes it worse that exhaustive search for any systems that can be solved on modern computers.
+New algorithms proposed by Itai Dinur \cite{Itai21A, Itai21B} where the complexity is brought down to $\tilde{\mathcal{O}}(2^{0.6943n})$, but still no implementation known to the authors.
+
+		The last type is a hybrid algorithms using algebraic geometry and gröbner basis algorithms and exhaustive search to find solutions.
+The most known and the best one is BooleanSolve hybrid by Bardet, Faugère, Salvy and Spaenlehauer \cite{BFSS13} where we guess a certain number of variables $k$ and apply F5 algorithm on the specialized systems.
+Under semi-regularity assumption, the algorithm finds the solution in $\mathcal{O}^(2^{0.792n})$ if $n = m$ and $k = 0.55n$.
+This is better than exhaustive search, but still, the constant factors are too big for it to be better for systems smaller with less than 200 variables.
+
+		The algorithm on which I focused on during my internship is Crossbred by Joux and Vitse \cite{JV17}.
+It mixes the usage of exhaustive search and algebraic algorithms like BooleanSolve hybrid but with a pre-processing step before doing the exhaustive search and applying algebraic algorithms like XL \cite{CKPS00} or it's variants or F5 \cite{F02}.
+
+		\begin{algorithm}
+			\caption{The CrossBred algorithm}\label{alg:crossbred}
+			\begin{algorithmic}[1]
+				\Require Polynomials $\mathcal{F} = {f_1, \dots, f_m}$ of $n$ variables and integers $D,d,k$
+				\Ensure  A solution for the system if it exists or nothing otherwise.
+				\State Construct $\Mac{arg1}{arg2}{arg3}$
+				\State Compute a basis $(v_1,\dots,v_r)$ if the left kernel of ...
+				\State Construct the polynomials $p_1,\dots,p_r$ corresponding ti ...
+				\For{$a = (a_{k+1},\dots,a_n) \in \mathbb{F}_{2}^{n-k}$}
+					\State a
+					\State a
+					\State a
+					\State a
+					\If{$S*$ is consistent using XL for example}
+						\State \textbf{Return} solution
+					\EndIf
+				\EndFor
+			\end{algorithmic}
+		\end{algorithm}
 
@@
 		In \cite{BFS15}, the authors show that the complexity of computing the Gröbner basis of a semi-regular system $f_1,...,f_m \in \mathbb{K}[x_1,...,x_n]$ is $O\left(mD\binom{n+D-1}{D}^\omega \right)$.
-
-		We will also need some properties of semi-regular sequences from \cite{Bardet04}
-
-		\begin{proposition}[Proposition 3.2.5 in \cite{Bardet04}]\label{PropSemi}
-			let $f_1,\dots,f_m$ a semi-regular homogeneous sequence, $f_i$ is of degree $d_i$. Then the following proprieties are verified:
-			\begin{itemize}
-				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if its Hilbert's series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^{d_i})/(1-z)^n.$$ $S_{m, n}(z)$ is called the generating series of $f_1,\dots,f_m$.
-				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if its Hilbert series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^{d_i}).$$ $S_{m, n}(z)$ is called the generating series of $f_1,\dots,f_m$.
-				\item Every permutation $f_{\sigma(1)},\dots,f_{\sigma(m)}$ is a semi-regular sequence.
-				\item When the associated variety is zero-dimensional, $H(I)$ is characterized by
-				$$
-				\forall d < H(I), h_{d, m}(n) > 0 \text{ and } h_{H(I), m}(n) \leq 0.
-				$$
-				\item The sequence $f_1,\dots,f_m$ is semi-regular $\nRightarrow \forall i \in [1;m]$, the sequence $f_1,\dots,f_i$ is semi-regular. 
-			\end{itemize}
-		\end{proposition}
 

### 2025-05-20 rapport.tex
version envoyé à Sorina

@@
 
-\author{Paul Mekhail\\ \textbf{Encadrante:} Sorina Ionica}
+\author{Paul Mekhail\\ \textbf{Advisor:} Sorina Ionica}
 
@@
 
-		The major "issue" of Gröbner basis algorithms is the reductions to 0 in the Macaulay matrix associated to a polynomial sequence $\Mac{D}{m}{F}$, because the size of the matrices are huge, multiple millions for practical examples,
+		The major "issue" of Gröbner basis algorithms like Bucherberger's \cite{Buc}, F4 \cite{F02} are the reductions to 0 in the Macaulay matrix associated to a polynomial sequence $\Mac{D}{m}{F}$, because the size of the matrices are huge, multiple millions for practical examples,
 		reductions to 0 can cost a lot during the linear algebra process.
@@
 		\subsection{Hilbert series}
-		Hilbert-Poincaré series (or just Hilbert series) is a very important tool in algebraic geometry to describe ideals and varieties and their dimensions.
+		Hilbert-Poincaré series (or just Hilbert series) is a very important tool in algebraic geometry to describe ideals, varieties and their dimensions.
 		\begin{definition}[\cite{CLS}]
@@
 
-		Thus, Bardet came with a new concept called semi-regularity for overdetermined system to allow their study and analysis. The definition of a semi-regular sequence is given by Bardet \cite{Bardet04}.
+		Thus, Bardet studied a new concept called semi-regularity for overdetermined system to allow their study and analysis. The definition of a semi-regular sequence is given by Bardet \cite{Bardet04}.
 
@@
 		\end{proposition}
-
-		\subsection{Boolean systems}
-
-		The problem of finding solutions to a multivariate polynomial system over a finite field $\mathbb{F}_p$ is proven NP-complete in \cite{FY79}.
-It is also well known that solving a quadratic polynomial system over any finite-field is NP-hard.
-This does not mean that any random instance is hard to solve.
-The problem of the MQ problem is that some instances can be very easy to solve, so the goal of cryptographers is to imagine a system with parameters that can be hard to solve in general and not have easy cases to solve most of the time.
-
-		Sometimes, the finite field used is $\mathbb{F}_2$ because it is still supposedly hard to solve but is very easy to use and manipulate on computers.
-But in such small fields one can begin to use exhaustive search to obtain the solution to the system.
-The best algorithm known to the authors for exhaustive search for multivariate polynomial system $f_1,\dots,f_m$ over $\mathbb{F}_2[x_1,\dots,x_n]$ is \cite{BCC+10} with a complexity $\mathcal{O}(ln(n)2^n)$, what is remarkable is that the complexity does not implicate the number of polynomials in the system, $m$.
-There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, libFES.
-
-		Another type of algorithms uses exhaustive search combined with probabilistic algorithms.
-\cite{LPT+17} gave an algorithm that finds the solution to a polynomial system with complexity $\tilde{\mathcal{O}}(2^{0.8765n})$ which is better than exhaustive search.
-Even though it's asymptotically better, there are no known implementations because of the huge constant factors present in the complexity that makes it worse that exhaustive search for any systems that can be solved on modern computers.
-New algorithms proposed by Itai Dinur \cite{Itai21A, Itai21B} where the complexity is brought down to $\tilde{\mathcal{O}}(2^{0.6943n})$, but still no implementation known to the authors.
-
-		The last type is a hybrid algorithms using algebraic geometry and gröbner basis algorithms and exhaustive search to find solutions.
-The most known and the best one is BooleanSolve hybrid by Bardet, Faugère, Salvy and Spaenlehauer \cite{BFSS13} where we guess a certain number of variables $k$ and apply F5 algorithm on the specialized systems.
-Under semi-regularity assumption, the algorithm finds the solution in $\mathcal{O}^(2^{0.792n})$ if $n = m$ and $k = 0.55n$.
-This is better than exhaustive search, but still, the constant factors are too big for it to be better for systems smaller with less than 200 variables.
-
-		The algorithm on which I focused on during my internship is Crossbred by Joux and Vitse \cite{JV17}.
-It mixes the usage of exhaustive search and algebraic algorithms like BooleanSolve hybrid but with a pre-processing step before doing the exhaustive search and applying algebraic algorithms like XL \cite{CKPS00} or it's variants or F5 \cite{F02}.
-
-		\begin{algorithm}
-			\caption{The CrossBred algorithm}\label{alg:crossbred}
-			\begin{algorithmic}[1]
-				\Require Polynomials $\mathcal{F} = {f_1, \dots, f_m}$ of $n$ variables and integers $D,d,k$
-				\Ensure  A solution for the system if it exists or nothing otherwise.
-				\State Construct $\Mac{arg1}{arg2}{arg3}$
-				\State Compute a basis $(v_1,\dots,v_r)$ if the left kernel of ...
-				\State Construct the polynomials $p_1,\dots,p_r$ corresponding ti ...
-				\For{$a = (a_{k+1},\dots,a_n) \in \mathbb{F}_{2}^{n-k}$}
-					\State a
-					\State a
-					\State a
-					\State a
-					\If{$S*$ is consistent using XL for example}
-						\State \textbf{Return} solution
-					\EndIf
-				\EndFor
-			\end{algorithmic}
-		\end{algorithm}
 
@@
 
-		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $D$ which is the degree of regularity of $I$ which is the ideal generated by the polynomials of the system.
+		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $H(I) - 1$, where $I$ is the ideal generated by the polynomials of the system.
 
@@
 
-		When applying Macaulay matrix generation code from hpXbred \cite{BS23} on NSBC polynomial systems, we notice that they are not semi-regular when we only multiply by the $x$ variables or only by the $y$ variables.
-If we multiply by the $x$ variables and at least one $y$ variable, we get no reductions to 0 which means that the system could be semi-regular or $\gamma$-semi-regular.
-
-		Notes pour moi: On a quelque chose de bizarre où si on prends la 2 de la proposition \ref{PropSemi}, on a pas semi-régularité voir rien de particulier.
-Mais si on prends la 1, donc oublier qu'on est sur $\mathbb{F}_2$, on a toujours pas de semi-regularité d'après la proposition, mais la 5 est vérifié avec la série generatrice, donc je soupçonne que la façon de laquelle sagemath calcule la série de hilbert est pas forcément correct, à vérifier avec des exemples de suites semi-regulière.
-Update: Il semblerait qu'avec des systèmes aléatoires, donc qui devrait être semi-régulier, sagemath contredit la 5 de \ref{PropSemi}.
-Donc il faut voir si la méthode .hilbert\_series est la bonne ou s'il faut utiliser autre chose.
-Update 2: Sage tronque la série de hilbert jusqu'au degré de régularité très certainement et ça correspond à la série génératrice mais seulement avec la formule en 1 de \ref{PropSemi} et non la 2 ce qui est pas normal vu qu'on est sur $\mathbb{F}_2$.
-Il est possible que ça soit la méthode de sage qui fait n'importe quoi quand on est sur $\mathbb{F}_2$.
-Après vérification, c'est bien ce que sagemath fait donc il prends pas en compte qu'on est sur $\mathbb{F}_2$.
-
 		\subsection{Bilinear systems case}
-		Our NSBC instances are quadratic systems that are bilinear in the variables of degree 2, overdetermined and can be semi-regular.
+		Our NSBC instances are quadratic systems that are bilinear in the variables of degree 2 and overdetermined.
 
@@
 
-		We will denote $n_x$ (resp.
-$n_y$) the number of variables in $x$ (resp.
-number of variables in $y$).
-In our case, $n_x = n_y = (m-1)/2$ and $m = k$.
-As we mentioned earlier, our polynomial systems can be semi-regular, so we get no reductions to 0 during the F5 algorithm using the Frobenius criterion for boolean polynomial systems until the Macaulay matrix of degree $D_{reg}$.
+		We will denote $n_x$ (resp. $n_y$) the number of variables in $x$ (resp. in $y$). In our case, $n_x = n_y = (m-1)/2$ and $m = k$.
 
@@
 
-		 No need to say that it's useless to use the criterion as shown in \cite{FSS11} without adapting it to get an advantage.
-		 This can be explained by the fact that the extended criterion computes reduced polynomials of degree $n_x$ and $n_y$ and in our case, $n_x$ and $n_y$ can be too big to compute.
-Maybe there could be a way to compute reduced polynomials of only degree 2 for our use case if we stop Macaulay matrix computations at degree 4 with our quadratic systems.
+	\section{Boolean systems}
+		Sometimes, the finite field used is $\mathbb{F}_2$ because it is still supposedly hard to solve but is very easy to use and manipulate on computers.
+But in such small fields one can begin to use exhaustive search to obtain the solution to the system.
+The best algorithm known to the authors for exhaustive search for multivariate polynomial system $f_1,\dots,f_m$ over $\mathbb{F}_2[x_1,\dots,x_n]$ is \cite{BCC+10} with a complexity $\mathcal{O}(ln(n)2^n)$, what is remarkable is that the complexity does not implicate the number of polynomials in the system, $m$.
+There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, \textbf{libFES}.
 
-		 As the authors of \cite{FSS11} noted at the end of their article, generic bilinear systems are regular.
-They didn't show that for overdetermined systems but we saw that experimentally but it's yet to be proven.
+		Another type of algorithms uses exhaustive search combined with probabilistic algorithms.
+\cite{LPT+17} gave an algorithm that finds the solution to a polynomial system with complexity $\tilde{\mathcal{O}}(2^{0.8765n})$ which is better than exhaustive search.
+Even though it's asymptotically better, there are no known implementations because of the huge constant factors present in the complexity that makes it worse that exhaustive search for any systems that can be solved on modern computers.
+New algorithms proposed by Itai Dinur \cite{Itai21A, Itai21B} where the complexity is brought down to $\tilde{\mathcal{O}}(2^{0.6943n})$, but still no implementation known to the authors.
 
-		 Note pour moi-même, il faut maintenant essayer de réaliser une implémentation de ce critère en sage ou avec hpXbred \cite{BS23} et BeanPolE \cite{Bou24} pour voir comment ça se comporterait avec des petits systèmes et voir quel serait l'avantage ou le désavantage sachant que la supposition que toutes les lignes se réduisent à 0 est extrêment peu probable.
-Donc les lignes se réduisent pas à 0 quand on a les x et les y donc le paramètre v de hpXbred est strictement supérieur à $n_x$.
-		 On peut ensuite utiliser XL \cite{CKPS00} avec l'implantation \cite{NNY17}
+		The last type is a hybrid algorithms using gröbner basis algorithms and exhaustive search to find solutions.
+The most known and the best one is BooleanSolve hybrid by Bardet, Faugère, Salvy and Spaenlehauer \cite{BFSS13} where we guess a certain number of variables $k$ and apply F5 algorithm on the specialized systems.
+Under semi-regularity assumption, the algorithm finds the solution in $\mathcal{O}^(2^{0.792n})$ if $n = m$ and $k = 0.55n$.
+This is better than exhaustive search, but still, the constant factors are too big for it to be better for systems smaller with less than 200 variables.
+
+		The algorithm on which I focused on during my internship is Crossbred by Joux and Vitse \cite{JV17}.
+It mixes the usage of exhaustive search and algebraic algorithms like BooleanSolve hybrid but with a pre-processing step before doing the exhaustive search and applying algebraic algorithms like XL \cite{CKPS00} or it's variants or F5 \cite{F02}.
+
+		\begin{algorithm}
+			\caption{The CrossBred algorithm}\label{alg:crossbred}
+			\begin{algorithmic}[1]
+				\Require Polynomials $\mathcal{F} = {f_1, \dots, f_m}$ of $n$ variables and integers $D,d,k$
+				\Ensure  A solution for the system if it exists or nothing otherwise.
+				\State Construct $\Mac{arg1}{arg2}{arg3}$
+				\State Compute a basis $(v_1,\dots,v_r)$ if the left kernel of ...
+				\State Construct the polynomials $p_1,\dots,p_r$ corresponding ti ...
+				\For{$a = (a_{k+1},\dots,a_n) \in \mathbb{F}_{2}^{n-k}$}
+				\State Evaluate the last $n-k$ variables in each $f_i \in \mathcal{F}$ at $(a_{k+1},\dots,a_n)$ and compute $\mathcal{F}*$
+				\State Compute $Mac_{\leq d, m}(\mathcal{F}*)$
+				\State Compute $\mathcal{F}'*$ as the partial evaluation of $F' = {p_1,\dots,p_r}$ at $(a_{k+1},\dots,a_n)$
+				\State Consider the system $S*$ consisting of $Mac_{\leq d, m}(\mathcal{F}*) \cup \mathcal{F}'*$
+				\If{$S*$ is consistent using XL for example}
+				\State \textbf{Return} solution
+				\EndIf
+				\EndFor
+			\end{algorithmic}
+		\end{algorithm}
+
+		This algorithm have shown it's effectiveness in practice with good implementations like Bouillaguet and Sauvage's implementation \cite {BS23} that broke records for solving boolean systems on the Fukuaka MQ challenge \cite{MQFUK}.
+
+	\section{Experimental results}
+		We tested if small instances of our problem are semi-regular or not using Proposition \ref{PropSemi} and they were not semi-regular.
+We also remarked that the bilinear criterion \ref{algo:crit} detected reductions to 0 using the original F5 criterion \ref{F5Crit} and Frobenius criterion for systems over $\mathbb{F}_2$ but also detected reductions to 0 on polynomials of degree $n_x$ or $n_y$ according to the algorithms \ref{alg:crit_calc} and \ref{alg:reduc}.
+In our case, we will use the Crossbred algorithm and we do not think that we will get to these degrees in the Macaulay matrices, so no need to use the bilinear criterion.
+
+		Even though no concrete complexity analysis is available for the Crossbred algorithm because of it's wide parameters choice, there are some articles that give formal tools to analyze admissible parameters and try to estimate the complexity for these parameters.
+We choose to focus on the article \cite{VID24} to select valid parameters and estimate the complexity of the algorithm using them.
+
+		\begin{theorem}
+			Let $H_{m, n}^k(X,Y) = \sum_{d_1 \geq 0, d_2 \geq 0} h_{d1, d2, m}^k X^{d_1}Y^{d_2}$ be the bivariate series with coefficient which is the number of new "independent" polynomials generated during the pre-processing.
+This series is given by:
+			$$
+				H_{m, n}^k(X, Y) = \frac{1}{Y}\left((1+X)^{n-k} - \frac{(1+XY)^k(1+X)^{n-k}}{(1+X^2Y^2)^m}\right).
+			$$
+		\end{theorem}
+
+		\begin{proposition}
+			For fixed values of $m, n \text{ and } k$ the bivariate series $G^{k}_{m,n}(X, Y) = $ \\ $\sum_{d_1 \geq 0, d_2 \geq 0}(\sum_{d_{1}' \leq d_1, d_{2}' \geq d_2} g_{d_{1}',d_{2}',m}^k)X^{d_1}Y^{d_2}$ is given by the formula:
+			$$
+				G^{k}_{m,n}(X, Y) = - \frac{YH_{m, n}^k(X,Y) - H_{m, n}^k(X, 1)}{(1-X)(1-Y)}.
+			$$
+
+			the coefficient in front of $X^{d_1}Y^{d_2}$ is the number of polynomials generated during the pre-processing step for a given pair $(d_1, d_2)$.
+		\end{proposition}
+
+		\begin{theorem}
+			Let $\mathcal{F} = {f_1,\dots,f_m}$ be a $\gamma$-strong semi-regular sequence of polynomials in $\mathbb{F}_2[x1,\dots,x_n]$ and consider $D < D_{reg}, d < d_{reg}(k).$ Then $k, D$ and $d$ are potentially admissible parameters for the Crossbred algorithm if the coefficient corresponding to $X^DY^d$ of the following bivariate series
+			$$
+			J_{m, n}^k(X, Y) = \frac{1}{(1-X)(1-Y)}\left(\frac{(1+X^{n-k})(1+XY)^k}{(1+X^2Y^2)^m} - \frac{(1+X)^n}{(1+X^2)^m} - \frac{(1+Y)^k}{(1+Y^2)^m}\right)
+			$$
+			is non-negative.
+		\end{theorem}
+
+		We wrote sage scripts that automated the search of admissible parameters using the bivariate series given in \cite{VID24}.
+We searched for different key sizes to have a rough estimate on the complexity.
+
+		For the complexity estimate for the precomputation step we took for reference the complexity of block-lanczos and block-wiedemann algorithms which is $\mathcal{O}(n*nnz)$ where $n$ is the number of rows and $nnz$ is the number of non zero elements of the matrix.
+
+		For the space complexity we thought that the best is to consider sparse data storage using CSR format where we store 2 arrays, one array is storing the number of non zero elements per row and the second array stores the columns of the non zero elements.
+No need to store the coefficients because they are always $1$.
+So the storage complexity is $\mathcal{O}(nb\_rows + nnz)$ bytes.
 

### 2025-05-21 rapport.tex
ajout preuve de sec UB-KOA et attaques (pas complet)

@@
 \newtheorem{corollary}{Corollary}[section]
+\newtheorem{lemma}{Lemma}[section]
 \newtheorem{example}{Example}[section]
@@
 
-	\section{Experimental results}
+	\subsection{Admissible parameters for Crossbred}
 		We tested if small instances of our problem are semi-regular or not using Proposition \ref{PropSemi} and they were not semi-regular.
@@
 
-		\begin{theorem}
-			Let $H_{m, n}^k(X,Y) = \sum_{d_1 \geq 0, d_2 \geq 0} h_{d1, d2, m}^k X^{d_1}Y^{d_2}$ be the bivariate series with coefficient which is the number of new "independent" polynomials generated during the pre-processing.
+		\begin{theorem}\label{serie_=H}
+			Let $H_{m, n}^k(X,Y) = \sum_{d_1 \geq 0, d_2 \geq 0} h_{d_1, d_2, m}^k X^{d_1}Y^{d_2}$ be the bivariate series with coefficient which is the number of new "independent" polynomials generated during the pre-processing.
 This series is given by:
@@
 
-		\begin{proposition}
+		\begin{proposition}\label{serie_G}
 			For fixed values of $m, n \text{ and } k$ the bivariate series $G^{k}_{m,n}(X, Y) = $ \\ $\sum_{d_1 \geq 0, d_2 \geq 0}(\sum_{d_{1}' \leq d_1, d_{2}' \geq d_2} g_{d_{1}',d_{2}',m}^k)X^{d_1}Y^{d_2}$ is given by the formula:
@@
 			$$
-
 			the coefficient in front of $X^{d_1}Y^{d_2}$ is the number of polynomials generated during the pre-processing step for a given pair $(d_1, d_2)$.
@@
 
-		\begin{theorem}
+		We also have that the definition of the coefficient \( h_{d_1, d_2, m}^k \) is given by:
+		\begin{align}
+			h_{d_1, d_2, m}^k =
+			\begin{cases}
+				U_{d_1, d_2, m}^k - M_{d_1, d_2 + 1}^k, & \text{if } d_1 \geq d_2 \geq 0,\ m \geq 1, \\
+				- M_{d_1, 0}, & \text{if } d_1 \geq 0,\ d_2 = -1,\ m \geq 0, \\
+				- M_{d_1, d_2 + 2}^k, & \text{if } d_1 \geq d_2 \geq 0,\ m = 0, \\
+				0, & \text{in all other cases}.
+			\end{cases}
+		\end{align}
+
+		\begin{theorem}\label{serie_J}
 			Let $\mathcal{F} = {f_1,\dots,f_m}$ be a $\gamma$-strong semi-regular sequence of polynomials in $\mathbb{F}_2[x1,\dots,x_n]$ and consider $D < D_{reg}, d < d_{reg}(k).$ Then $k, D$ and $d$ are potentially admissible parameters for the Crossbred algorithm if the coefficient corresponding to $X^DY^d$ of the following bivariate series
@@
 
-		For the complexity estimate for the precomputation step we took for reference the complexity of block-lanczos and block-wiedemann algorithms which is $\mathcal{O}(n*nnz)$ where $n$ is the number of rows and $nnz$ is the number of non zero elements of the matrix.
+		We discuss more the attacks complexity using Crossbred and different algorithms in the next section.
 
-		For the space complexity we thought that the best is to consider sparse data storage using CSR format where we store 2 arrays, one array is storing the number of non zero elements per row and the second array stores the columns of the non zero elements.
-No need to store the coefficients because they are always $1$.
-So the storage complexity is $\mathcal{O}(nb\_rows + nnz)$ bytes.
+	\section{Security analysis}
+		This section is heavily inspired from the security analysis of the MQOM signature scheme \cite{MQOMv2}.
+
+		\subsection{Security proofs}
+		To prove the security of a signature schema, one can simulate a game with assumptions and prove that the scheme is not breakable with assumptions on the adversary capabilities and the security goals.
+
+		There exists 3 main security goals:
+		\begin{itemize}
+			\item \textit{Unbreakability}: the attacker recovers the secret key from the verification key(or public key). It's denoted \textbf{UB}.
+			\item \textit{Universal Unforgeability}: the attacker can produce a valid signature of any message in the message space without necessarily recovering the verification key.
+It's denoted \textbf{UUF}.
+			\item \textit{Existential Unforgeability}: the attacker creates a message and produces a valid signature of it without necessarily having a choice for the message.
+			It's denoted \textbf{EUF}.
+		\end{itemize}
+		Finally, there are 3 main adversarial models:
+		\begin{itemize}
+			\item \textit{Key-Only Attacks}: The adversary has only access to the verification keys, which is the simplest and unavoidable scenario. It's denoted \textbf{KOA}.
+			\item \textit{Known-Message Attacks}: The adversary has access to a set of messages and their valid signatures. It's denoted \textbf{KMA}. 
+			\item \textit{Chosen-Message Attacks}: The adversary can probe an oracle that can sign any message of his choice and he can request as much as he wants. It's denoted \textbf{CMA}.
+		\end{itemize}
+		The definitions are inspired from Damien Vergnaud's lectures at Sorbonne-Université.
+
+		The goal of any signature scheme is to be \textbf{EUF-CMA}, which is the strongest security proof. First we prove that the schema is \textbf{UB-KOA}, which is the weakest proof but the easiest.
+
+		\begin{lemma}\label{UB-KOA}
+			Suppose that the SBC function $\text{SBC}: \mathbb{F}_{q^k}^{2n} \rightarrow \mathbb{F}_{q}^{2n}$ is ($\tau, \epsilon$)-hard, i.e.
+there are no $\mathcal{PPT}$ $\mathcal{A}$ with running time upper-bounded by $\tau$ such that the SBC problem is hard to solve for a certain parameter set $(q, n, m)$,
+			\begin{align*}
+				\text{Succ}(\mathcal{A}) = \mathcal{P}\left[
+				\begin{array}{l}
+					(u, v) \leftarrow \mathbb{F}_{q^k}^{2n}, \\
+					(x, y) \leftarrow \mathcal{A}(u, v), \\
+					(x, y) \in \text{SBC}[u, v]
+				\end{array}
+				\right] < \epsilon,
+			\end{align*}
+			then the signature scheme is \textbf{UB-KOA}.
+		\end{lemma}
+		The proof of \ref{UB-KOA} can be found in \ref{appendix:proof}.
+
+		\begin{theorem}\label{EUF-CMA}
+			Suppose that no adversary can distinguish the PRF used by the protocol from real random bits
+			$$
+				| \mathcal{P}_{seed \in \{0,1\}^n}[b=0 | p \leftarrow PRF(seed)] - \mathcal{P}[b=1 | p \overset{{\scriptscriptstyle\$}}{\leftarrow} \{0,1\}^{2n}] | \leq \frac{1}{2}.
+			$$
+			Suppose also that the protocol uses a cryptographic hash function which is one-way and does not have collision attacks with better complexity than $\frac{1}{\lambda}$ using a $\mathcal{PPT}$, where $\lambda$ is the number security bits needed for the signature scheme.
+			Finally, suppose that the SBC problem is hard, i.e. the complexity the best attack of the SBC problem has complexity $\frac{1}{\lambda}$ using a $\mathcal{PPT}$.
+			Then, the SBC signature scheme is \textbf{EUF-CMA}.
+		\end{theorem}
+		The proof of \ref{EUF-CMA} can be found in the appendix \ref{appendix:proof}.
+
+		\subsection{Attacks against SBC signatures}
+		First we need to enumerate the parameters for the signature scheme.
+The finite field used is $\mathbb{F}_2$ and the vector length $n$ is 130 but we consider the NSBC version so it's 128 in practice with the 2 last coordinates implicitly defined.
+The field extension used $k$ is 257 which is prime to avoid subfield attacks \cite{Peb24}.
+
+		For the linear algebra contant $\omega$, we take $\omega = 2.81$ for dense linear algebra using the Strassen algorithm, which is implemented in dense linear boolean algebra software like M4RI and M4RIE \cite{M4RI}.
+
+		For sparse linear algebra, we suppose the usage of block-Wiedemann algorithm which is the most used widely in software like CADO-NFS \cite{cadonfs} and in XL parallel implementation \cite{NNY17}.
+The complexity of this algorithm depend on the "blocking factor" that we will denote $\tilde{m}$ and $\tilde{n}$, the size of the matrix denoted $N$ and the number of non-zero elements denoted $nnz$.
+Most of the algorithm boils down to (sparse $N \times N$ matrix) $\times$ (dense $N$ vector) products.
+There are 3 phases in the block-Wiedemann algorithm BW1, BW2 and BW3.
+The reader can refer to \cite{NNY17} for a clear description of these phases.
+The most costly phase is BW1 where it does $(1/\tilde{n} + 1/\tilde{m})N$ (dense $\tilde{m} \times N$ matrix) x vector products that require $2N\tilde{m}$ operations.
+The complexity of BW2 and BW3 phases can be negligible compared to BW1.
+So we estimate the total number of arithmetic operations of the block-Wiedemann algorithm to be $(2(1 + \tilde{n}/\tilde{m})N(nnz + \tilde{m}N))$.
+\textbf{A préciser et comprendre si possible}.
+
+		We need also to clarify some properties that we use for the Macaulay matrices.
+First we know that the number of monomials of degree exactly $d$ (resp.
+of $\leq d$) in $n$ variables, which is the number of columns of the Macaulay matrix of degree $d$ (resp.
+of degree $\leq d$), is $\binom{n}{d}$ (resp.
+$\sum_{i=0}^{d}\binom{n}{d}$) because of the frobenius identity that reduces their number.
+A good remark is that the number of monomials of system modelisation of SBC problems contain exactly $\frac{n^2}{4} + \frac{n}{2} + 1$ monomials.
+For the number of rows, we will use bivariate serie G \ref{serie_G}.
+Finally, we know that Macaulay matrices are very sparse and each row contain at most $(n+1)(n+2)/2$ non zero coefficients.
+
+		The first attack is exhaustive search which can be done in $\mathcal{O}(ln(n)2^n)$ operations in the field with negligible space complexity \cite{BCC+10}.
+		There is very optimized implementations like libFES by Charles Bouillaguet and there have been implementations on FPGA that broke real world cryptography \cite{DDVY20}.
+
+		There is also the Hybrid Solve algorithm by Bardet et al.
+\cite{BFSS13} with average running time $\tilde{\mathcal{O}}(2^{0.792n})$ under some algebraic assumptions like semi-regularity.
+We do not seem to have this assumption on our systems, but we know thanks to \cite{FSS11} that the degree of regularity upper-bound is lower than for semi-regular systems \textbf{A écrire dans la section de l'article car important et referencer}, but the cost the bilinear criterion seems to be superior to that of the cost of solving \ref{complexCrit} so we will not consider this attack.
+
+		One type of attack is the probabilistic attacks. The best ones are Dinur's first and second algorithm \cite{Itai21A, Itai21B}.
+		Their complexity is...
+
+		Finally, the Crossbred algorithm...
 
@@
 
+\appendix
+\section{Security proof}\label{appendix:proof}
+	\begin{lemma}
+		Suppose that the SBC function $\text{SBC}: \mathbb{F}_{q^k}^{2n} \rightarrow \mathbb{F}_{q}^{2n}$ is ($\tau, \epsilon$)-hard, i.e.
+there are no $\mathcal{PPT}$ $\mathcal{A}$ with running time upper-bounded by $\tau$ such that the SBC problem is hard to solve for a certain parameter set $(q, n, m)$,
+		\begin{align*}
+			\text{Succ}(\mathcal{A}) = \mathcal{P}\left[
+			\begin{array}{l}
+				(u, v) \leftarrow \mathbb{F}_{q^k}^{2n}, \\
+				(x, y) \leftarrow \mathcal{A}(u, v), \\
+				(x, y) \in \text{SBC}[u, v]
+			\end{array}
+			\right] < \epsilon,
+		\end{align*}
+		then the signature scheme is \textbf{UB-KOA}.
+	\end{lemma}
+	\begin{proof}
+		We need to show that if a $\mathcal{PPT}$ adversary $\mathcal{A}$ can retrieve the private key from the public key, then we can break the SBC function.
+		It's straightforward to show, because if $\mathcal{A}$ can retrieve the private key $(x, y) \in \mathbb{F}_{q}^{2n}$ of a given valid public key $(u, v) \in \mathbb{F}_{q^k}^{2n}$ then no need to convert anything because $(x, y) \in \text{SBC}[u, v]$ directly.
+So the signature scheme is \textbf{UB-KOA} if the SBC function is ($\tau, \epsilon$)-hard.
+	\end{proof}
+
+	\begin{theorem}
+		Suppose that no adversary can distinguish the PRF used by the protocol from real random bits
+		$$
+		| \mathcal{P}_{seed \in \{0,1\}^n}[b=0 | p \leftarrow PRF(seed)] - \mathcal{P}[b=1 | p \overset{{\scriptscriptstyle\$}}{\leftarrow} \{0,1\}^{2n}] | \leq \frac{1}{2}.
+		$$
+		Suppose also that the protocol uses a cryptographic hash function which is one-way and does not have collision attacks with better complexity than $\frac{1}{\lambda}$ using a $\mathcal{PPT}$, where $\lambda$ is the number security bits needed for the signature scheme.
+		Finally, suppose that the SBC problem is hard, i.e. the complexity the best attack of the SBC problem has complexity $\frac{1}{\lambda}$ using a $\mathcal{PPT}$.
+		Then, the SBC signature scheme is \textbf{EUF-CMA}.
+	\end{theorem}
+	\begin{proof}
+
+	\end{proof}
+
 \end{document}

### 2025-05-21 rapport.tex
corrections par Marize

@@
 
-\title{Cryptanalysis of post-quantum signature scheme while breaking dlog}
+\title{Cryptanalysis of post-quantum signature scheme while breaking Dlog}
 
@@
 			where $2 \leq \omega \leq 3$.
-			\item Let $R_{s}$ be the polynomials of total degree $s$. $I_{s} = I \cap R_{s}$.
+			\item Let $R_{s}$ be the polynomials of total degree $s$. $I_{s} = I \cap R_{s}$ where $I$ is an ideal.
 		\end{itemize}
@@
 
-		The major "issue" of Gröbner basis algorithms like Bucherberger's \cite{Buc}, F4 \cite{F02} are the reductions to 0 in the Macaulay matrix associated to a polynomial sequence $\Mac{D}{m}{F}$, because the size of the matrices are huge, multiple millions for practical examples,
+		The major "issue" of Gröbner basis algorithms like Bucherberger's \cite{Buc}, F4 \cite{F02} are the reductions to 0 in the Macaulay matrix associated to a polynomial sequence $\Mac{D}{m}{F}$, because the size of the matrices are huge, multiple millions rows and columns for practical examples,
 		reductions to 0 can cost a lot during the linear algebra process.
@@
 			\label{F5Crit}
-			Let $d_i$ be the total degree of $f_i$.
+			Let $d_i$ be the total degree of $f_i \forall i \in [1,m]$.
 			$\forall j < m$, if a row of signature ($u, f_j$) in the matrix $\Mact{D-d_j}{m-1}{F}$ has as leading term $t'$, then the row ($t', f_m$) in $\Mac{D}{m}{F}$ (resp.
@@
 			\label{Frob}
-			If a row of signature ($t, f_m$) in $\Mact{D - d_m}{m}{F}$ has as leading term $t'$, then the row ($t', f_m$) in $\Mac{D}{m}{F}$ (resp.
-$\Mac{\leq D}{m}{F}$) will reduce to 0 i.e.
-is a linear combination of the predecessors.
+			If a row of signature ($t, f_m$) in $\Mact{D - d_m}{m}{F}$ has as leading term $t'$, then the row ($t', f_m$) in $\Mac{D}{m}{F}$ (resp. $\Mac{\leq D}{m}{F}$) will reduce to 0.
 		\end{proposition}
@@
 		The Frobenius criterion \ref{Frob} is the analogue of the General criterion \ref{F5Crit} but for polynomial systems defined over $\mathbb{F}_2$.
-		F5 crtierions remove the rows corresponding to the trivial syzygies, i.e.
+		F5 criterion and Frobenius criterion remove the rows corresponding to the trivial syzygies, i.e.
 the syzygies $(s_1, \dots, s_m)$ such that for a row with signature $(s_i, f_i)$, we get $s_i \in \langle f_1,\dots,f_{i-1},f_{i+1},\dots,f_m \rangle$.
@@
 						\If{$d_i = d$}
-							\State add the row $f_i$ with signature $(1, f_i)$
+							\State Add the row $f_i$ with signature $(1, f_i)$
 						\EndIf
 						\If{$d > d_i$}
-							for all $f$ from $\widetilde{Mac}_{d, i-1}$ with signature $(e, f_i)$, such that $x_\lambda$ is the greatest variable of $e$, add the $n - \lambda + 1$ rows $x_\lambda f,x_{\lambda + 1}f,\dots,x_n f$ with the signatures $(x_\lambda e, f),(x_{\lambda + 1}e, f),\dots,(x_n e, f)$ except those who satisfy the F5 criterion with $((x_{\lambda + k}e, fi), \widetilde{Mac}_{d-d_i, i-1}$
+							For all $f$ from $\widetilde{Mac}_{d, i-1}$ with signature $(e, f_i)$, such that $x_\lambda$ is the greatest variable of $e$, add the $n - \lambda + 1$ rows $x_\lambda f,x_{\lambda + 1}f,\dots,x_n f$ with the signatures $(x_\lambda e, f),(x_{\lambda + 1}e, f),\dots,(x_n e, f)$ except those who satisfy the F5 criterion with $((x_{\lambda + k}e, fi), \widetilde{Mac}_{d-d_i, i-1}$
 						\EndIf
 						\State Compute $\widetilde{Mac}_{d, i}$ the row echelon form of ${Mac}_{d, i}$
-						\State Add to $G$ the polynomials corresponding to rows of $\widetilde{Mac}_{d, i}$ such that their leading
-						\State monomials is different from the leading monomial of the row with 
+						\State Add to $G$ the polynomials corresponding to the rows of $\widetilde{Mac}_{d, i}$ such that their leading
+						\State monomials is different from the leading monomial of the row with the
 						\State same signature in ${Mac}_{d, i}$
@@
 		\subsection{Hilbert series}
-		Hilbert-Poincaré series (or just Hilbert series) is a very important tool in algebraic geometry to describe ideals, varieties and their dimensions.
+		Hilbert-Poincaré series (or Hilbert series) is a very important tool in algebraic geometry to describe ideals, varieties and their dimensions.
 		\begin{definition}[\cite{CLS}]
-			Let $I$ be a homogeneous ideal in $R$. The affine Hilbert function of $I$ is the function on the nonnegative integers s defined by
+			Let $I$ be a homogeneous ideal in $R$. The affine Hilbert function of $I$ is the function on the nonnegative integers is defined by
 			$$
@@
 
-		Faugère showed that no reductions to 0 occur in the Macaulay matrix during the F5 algorithm \ref{alg:matrix_f5} until the degree of regularity \cite{F02}.
+		Faugère showed that no reductions to 0 occur in the Macaulay matrix during the F5 algorithm \ref{alg:matrix_f5} until the degree of regularity \cite{F02} for regular sequences.
 
 		\subsection{Semi-regularity}
-		One problem with the definition of a regular sequence, is that it applies for only underdetermined  polynomial systems because on of the proprieties of a regular sequence of a polynomial system $f_1,\dots,f_m \in \mathbb{K}[x_1,\dots,x_n]$ is that the dimension of the ideal $\langle f_1,\dots,f_m \rangle$ is $n - m$.
-This propriety can not be verified for overdetermined systems because the dimension of the ideal can not be negative.
+		One problem with the definition of a regular sequence, is that it applies for only underdetermined  polynomial systems because on of the properties of a regular sequence of a polynomial system $f_1,\dots,f_m \in \mathbb{K}[x_1,\dots,x_n]$ is that the dimension of the ideal $\langle f_1,\dots,f_m \rangle$ is $n - m$.
+This property can not be verified for overdetermined systems because the dimension of the ideal can not be negative.
 
@@
 		\begin{proposition}[Proposition 3.2.5 in \cite{Bardet04}]\label{PropSemi}
-			let $f_1,\dots,f_m$ a semi-regular homogeneous sequence, $f_i$ is of degree $d_i$. Then the following proprieties are verified:
+			Let $f_1,\dots,f_m$ a semi-regular homogeneous sequence, $f_i$ is of degree $d_i$. Then the following properties are verified:
 			\begin{itemize}
@@
 
-	\section{Signature Scheme}
+	\section{SBC Signature Scheme}
 		\subsection{SBC Problem}
@@
 		\newline
-		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent overs $\mathbb{F}_q$, find two non-colinear vectors $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})$$
+		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent over $\mathbb{F}_q$, find two non-colinear vectors $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})$$
 		where $(\cdot)$ is the dot product between 2 vectors.
@@
 
-		\subsection{NSBC}
+		\subsection{NSBC variant}
 
@@
 		$\vec{x} = (\vec{x'}, 1, 0), \vec{y} = (\vec{y'}, 0, 1)$ to ensure that they are not colinear without needing to regenerate each time they are not.
+We also denote $(\vec{x}, \vec{y}) \in NSBC[\vec{u}, \vec{v}]$ if $(\vec{x}, \vec{y})$ are solutions to the NSBC problem on $(\vec{u}, \vec{v})$.
 
@@
 			\begin{itemize}
-				\item[-] If the sequence is semi-regular, then no reductions to 0 are performed during the F5-matrix algorithm until the degree of regularity $D - 1$.
+				\item[-] If the sequence is semi-regular, then no reductions to 0 are performed during the F5-matrix algorithm until the degree $d = H(I) - 1$
 				\item[-] If there are no reductions to 0 during the F5-matrix algorithm until the degree $D - 1$, and if the matrix of degree D is full rank and is the first matrix to have more rows than columns, then the sequence is semi-regular and its index of regularity is $H(I) = D$.
@@
 
-		We will also need the next definition of $\gamma$-semi-regularity.
+		We will also need the next definition of $\gamma$-strong semi-regularity.
 
@@
 
-		The article \cite{FSS11} and the thesis \cite{Spaen2012} provide a new criterion for bilinear determined systems and a complexity analysis of their algorithm (the proofs in the end of \cite{FSS11} are not exact, the authors recommended to refer to \cite{Spaen2012} for better proofs) that can help more remove reductions to 0 in F5 Algorithm for bilinear systems which are not semi-regular.
+		The article \cite{FSS11} and the thesis \cite{Spaen2012} provide a new criterion for bilinear determined systems and a complexity analysis of their algorithm (the proofs in the end of \cite{FSS11} are not exact, the authors recommended to refer to \cite{Spaen2012} for better proofs) that removes all reductions to 0 in F5 Algorithm for bilinear systems which are not semi-regular.
 
@@
 
-		 The following theorem, ensures that there are no reductions to 0 with extended the F5 criterion. Ce théorème est en suspens pour utiliser une autre formulation sans utiliser la topologie de Zariski
+		 The following theorem, ensures that there are no reductions to 0 with extended the F5 criterion.
+\textbf{Ce théorème est en suspens pour utiliser une autre formulation sans utiliser la topologie de Zariski}
 
@@
 		 	First, if $i > n_y$ (resp.
-$i > n_x$), we construct that jacobian matrix $jac_y(F_{i-1})$ (resp.
+$i > n_x$), we construct the jacobian matrix $jac_y(F_{i-1})$ (resp.
 $jac_x(F_{i-1})$) which is of size $n_y \times n_y$ (resp.
@@
 $n_x \times n_x$) and for a matrix $M$ of size $l \times c$ with $l > c$ there are $\binom{l}{c}$ sub-matrices of size $c \times c$, hence the term $\binom{i}{n_y}n_y^{\omega}$ in the sum if we suppose the complexity of computing the determinant of a square multivariate polynomial matrix of size $c \times c$ to be $\mathcal{O}(c^\omega)$, which is not the case but let's suppose the best case ever.
-faudrait trouver la vraie complexité de ça pour être plus juste.
-		 	For now, we have a sequence of polynomials and we apply the algorithm \textbf{Reduce} on them with degree $n_y + 1$ (resp. $n_x + 1$).
+\textbf{faudrait trouver la vraie complexité de ça pour être plus juste.}
+		 	For now, we have a sequence of polynomials and we apply the algorithm \textbf{Reduce} \ref{alg:reduc} on them with degree $n_y + 1$ (resp. $n_x + 1$).
 		 	In \textbf{Reduce}, we construct the Macaulay matrix of the sequence of degree $n_y + 1$ (resp. $n_x + 1$) and compute it's row echelon form.
@@
 
-		 Since we have quadratic systems, reductions to 0 won't appear before degree 4 of the Macaulay matrix.
-Also, in practice we are using $q=2$ and there's much better algorithms in the literature for boolean systems than basic Gröbner basis algorithms like F5.
+		 Since we have quadratic systems, reductions to 0 would not appear before degree 4 of the Macaulay matrix.
+Also, in practice we are using $q=2$ and there are much better algorithms in the literature for boolean systems than basic Gröbner basis algorithms like F5.
 The best algorithm used at the moment is CrossBred \cite{JV17}.
 Most of the time in practical implementations, we stop at degree 4 of the Macaulay in pre-processing step.
-So, to know what advantage we could have if we use the extended F5 criterion, we will suppose the worst, that all of our entries will reduce to 0 at degree 4 of the Macaulay matrix, which is unlikely.
+So, to know what advantage we could have if we use the extended F5 criterion, we will suppose the worst, meaning that all of our entries will reduce to 0 at degree 4 of the Macaulay matrix, which is unlikely.
 
@@
 
-		 Using the computer algebra system SageMath \cite{sagemath} to test numerically when using the F5 extended criterion of \cite{FSS11} becomes negligible compared to the computations of reductions to 0, we see experimentally that $Adv(10, 12) = 2498903$, i.e.
+		 We used the computer algebra system SageMath \cite{sagemath} to test numerically when using the F5 extended criterion of \cite{FSS11} becomes negligible compared to the computations of reductions to 0, we see experimentally that $Adv(10, 12) = 2498903$, i.e.
 the theoretical advantage becomes negligible and it would be advantageous to simply compute reductions to 0 instead of applying the F5 extended criterion.
 This number grows exponentially with $n$ and $m$.
-For those who wonder, what it would be the estimated theoretical advantage computing reductions to 0 for the parameters given by \cite{HJ23} $Adv(256, 257)$ is a 514 bit number.
+For those who wonder, what would be the estimated theoretical advantage computing reductions to 0 for the parameters given by \cite{HJ23} $Adv(256, 257)$ is a 514 bit number.
 
@@
 But in such small fields one can begin to use exhaustive search to obtain the solution to the system.
-The best algorithm known to the authors for exhaustive search for multivariate polynomial system $f_1,\dots,f_m$ over $\mathbb{F}_2[x_1,\dots,x_n]$ is \cite{BCC+10} with a complexity $\mathcal{O}(ln(n)2^n)$, what is remarkable is that the complexity does not implicate the number of polynomials in the system, $m$.
+The best algorithm known to the authors for exhaustive search for multivariate polynomial system $f_1,\dots,f_m$ over $\mathbb{F}_2[x_1,\dots,x_n]$ is \cite{BCC+10} with a complexity $\mathcal{O}(ln(n)2^n)$.
+What is remarkable is that the complexity does not implicate the number of polynomials in the system, $m$.
 There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, \textbf{libFES}.
@@
 \cite{LPT+17} gave an algorithm that finds the solution to a polynomial system with complexity $\tilde{\mathcal{O}}(2^{0.8765n})$ which is better than exhaustive search.
-Even though it's asymptotically better, there are no known implementations because of the huge constant factors present in the complexity that makes it worse that exhaustive search for any systems that can be solved on modern computers.
-New algorithms proposed by Itai Dinur \cite{Itai21A, Itai21B} where the complexity is brought down to $\tilde{\mathcal{O}}(2^{0.6943n})$, but still no implementation known to the authors.
+Even though it's asymptotically better, there are no known implementations because of the huge constant factors present in the complexity that makes it worse than exhaustive search for any systems that can be solved on modern computers.
+New algorithms have been proposed by Itai Dinur \cite{Itai21A, Itai21B} where the complexity is brought down to $\tilde{\mathcal{O}}(2^{0.6943n})$, but still no implementation known to the authors.
 
@@
 
-		We also have that the definition of the coefficient \( h_{d_1, d_2, m}^k \) is given by:
+		We also have that by definition \( h_{d_1, d_2, m}^k \) is defined as:
 		\begin{align}
@@
 		\subsection{Security proofs}
-		To prove the security of a signature schema, one can simulate a game with assumptions and prove that the scheme is not breakable with assumptions on the adversary capabilities and the security goals.
+		To prove the security of a signature scheme, one can simulate a game with assumptions and prove that the scheme is not breakable with assumptions on the adversary capabilities and the security goals.
 
@@
 			$$
-			Suppose also that the protocol uses a cryptographic hash function which is one-way and does not have collision attacks with better complexity than $\frac{1}{\lambda}$ using a $\mathcal{PPT}$, where $\lambda$ is the number security bits needed for the signature scheme.
-			Finally, suppose that the SBC problem is hard, i.e. the complexity the best attack of the SBC problem has complexity $\frac{1}{\lambda}$ using a $\mathcal{PPT}$.
+			Suppose also that the protocol uses a cryptographic hash function which is one-way and does not have collision attacks with better complexity than $2^{\lambda}$ using a $\mathcal{PPT}$, where $\lambda$ is the number security bits needed for the signature scheme.
+
+			Finally, suppose that the SBC problem is hard, i.e. the complexity of the best attack of the SBC problem has complexity $2^{\lambda}$ using a $\mathcal{PPT}$.
 			Then, the SBC signature scheme is \textbf{EUF-CMA}.
@@
 
-		For the linear algebra contant $\omega$, we take $\omega = 2.81$ for dense linear algebra using the Strassen algorithm, which is implemented in dense linear boolean algebra software like M4RI and M4RIE \cite{M4RI}.
+		For the linear algebra constant $\omega$, we take $\omega = 2.81$ for dense linear algebra using the Strassen algorithm, which is implemented in dense linear boolean algebra software like M4RI and M4RIE \cite{M4RI}.
 
-		For sparse linear algebra, we suppose the usage of block-Wiedemann algorithm which is the most used widely in software like CADO-NFS \cite{cadonfs} and in XL parallel implementation \cite{NNY17}.
-The complexity of this algorithm depend on the "blocking factor" that we will denote $\tilde{m}$ and $\tilde{n}$, the size of the matrix denoted $N$ and the number of non-zero elements denoted $nnz$.
+		For sparse linear algebra, we suppose the use of block-Wiedemann algorithm which is the most used in software like CADO-NFS \cite{cadonfs} and in XL parallel implementation \cite{NNY17}.
+The complexity of this algorithm depends on the "blocking factor" that we will denote $\tilde{m}$ and $\tilde{n}$, the size of the matrix denoted $N$ and the number of non-zero elements denoted $nnz$.
 Most of the algorithm boils down to (sparse $N \times N$ matrix) $\times$ (dense $N$ vector) products.
@@
 
-		We need also to clarify some properties that we use for the Macaulay matrices.
+		We also need to clarify some properties that we use for the Macaulay matrices.
 First we know that the number of monomials of degree exactly $d$ (resp.
@@
 A good remark is that the number of monomials of system modelisation of SBC problems contain exactly $\frac{n^2}{4} + \frac{n}{2} + 1$ monomials.
-For the number of rows, we will use bivariate serie G \ref{serie_G}.
+For the number of rows, we will use the bivariate series G \ref{serie_G}.
 Finally, we know that Macaulay matrices are very sparse and each row contain at most $(n+1)(n+2)/2$ non zero coefficients.
@@
 \cite{BFSS13} with average running time $\tilde{\mathcal{O}}(2^{0.792n})$ under some algebraic assumptions like semi-regularity.
-We do not seem to have this assumption on our systems, but we know thanks to \cite{FSS11} that the degree of regularity upper-bound is lower than for semi-regular systems \textbf{A écrire dans la section de l'article car important et referencer}, but the cost the bilinear criterion seems to be superior to that of the cost of solving \ref{complexCrit} so we will not consider this attack.
+We do not seem to have this assumption on our systems, but we know thanks to \cite{FSS11} that the degree of regularity upper-bound is lower than for semi-regular systems \textbf{A écrire dans la section de l'article car important et referencer}, but the cost of the bilinear criterion seems to be superior to that of the cost of solving \ref{complexCrit} so we will not consider this attack.
 

### 2025-05-22 rapport.tex
ajout idées utiles

@@
 
+		We need to first compute the number of rows of $\mathcal{M}_{\leq D, \geq d, m}^k(\mathcal{F})$ before and the number of nonzero elements per column for our systems.
+		The number of rows is bounded by $$nb\_monomials\_\mathcal{M}_{\leq D, \geq d, m}^k = \sum_{i=d+1}^{D}\sum_{j=0}^{d-i} \binom{k}{i} \cdot \binom{n-k}{j}$$
+		Finally, we saw by experiment that the number of monomials present in the our system polynomials is $\frac{n^2}{4} + \frac{n}{2} + 1$ which is the upper-bound for the number of nonzero elements per row.
+
 \printbibliography

### 2025-05-23 rapport.tex
minor changes

--- 2025-05-21 rapport.tex
@@
 
+		We need to first compute the number of rows of $\mathcal{M}_{\leq D, \geq d, m}^k(\mathcal{F})$ before and the number of nonzero elements per column for our systems.
+We focus on this matrix because it's the one on which we apply block-lancszos to find vectors in its co-kernel.
+		The number of rows is bounded by $$nb\_monomials\_\mathcal{M}_{\leq D, \geq d, m}^k = \sum_{i=d+1}^{D}\sum_{j=0}^{d-i} \binom{k}{i} \cdot \binom{n-k}{j}$$
+		Finally, we saw by experiment that the number of monomials present in the our system polynomials is $\frac{n^2}{4} + \frac{n}{2} + 1$ which is the upper-bound for the number of nonzero elements per row instead of the standard $\mathcal{O}(n^2)$.
+
+		After finding the vectors, comes the exhaustive search part. We will
+
 \printbibliography

### 2025-05-27 rapport.tex
ajout des parties sur crossbred et début preuves

@@
 
-\title{Cryptanalysis of post-quantum signature scheme while breaking Dlog}
+\title{Cryptanalysis of SBC based post-quantum signature scheme}
 
@@
 
+		The signature protocol is described in .. with the signature and verification algorithms.
+
 		\section{Cryptanalysis of the scheme}
@@
 
-		 Finally, the authors give the following corollary
+		 An important result of \cite{FSS11} is on the index of regularity of bi-regular systems.
 
-		 \begin{corollary}[Corollary 3 in \cite{FSS11}]
+		 \begin{theorem}\label{reg_bi}
+		 	If the system $\mathcal{F}$ is generic, then the degree of regularity of $\mathcal{V}(I)$ is upper bounded by $d_{reg} \leq min(n_x, n_y)$.
+		 \end{theorem}
+
+		 Finally, the authors give the following corollary which is the consequence of \ref{reg_bi} by applying the complexity of computing a Gröbner basis given by \cite{Bardet04}.
+
+		 \begin{corollary}[Corollary 3 in \cite{FSS11}]\label{complex_bi}
 		 The arithmetic complexity of computing a Gröbner basis of a generic bilinear system $f_1,...,f_{n_x + n_y}$ with the F5 algorithm is bounded by
@@
 
+		To understand this algorithm, we will need definitions of special Macaulay matrices used by it.
+
+		We will denote by $deg^k(p)$ the total degree of $p$ in the first $k$ variables.
+
+		\begin{definition}
+			For a given $\mathcal{F} = \{f_1,\dots,f_m\}$, $Mac_{\leq D, \geq d, m}^k$ is the submatrix of $Mac_{\leq D, m}$ whose rows correspond to a product $uf_i$, $1 \leq i \leq m$, with $deg^k(u) \geq d-1$.
+Let $\mathcal{M}_{\leq D, \geq d, m}^k$ be the submatrix of $Mac_{\leq D, \geq d, m}^k$ with columns correspond to monomials $m$ such that, $deg^k(m) \geq d+1$.
+		\end{definition}
+
 		\begin{algorithm}
@@
 				\Ensure  A solution for the system if it exists or nothing otherwise.
-				\State Construct $\Mac{arg1}{arg2}{arg3}$
-				\State Compute a basis $(v_1,\dots,v_r)$ if the left kernel of ...
-				\State Construct the polynomials $p_1,\dots,p_r$ corresponding ti ...
+				\State Construct $Mac_{\leq D, \geq d, m}^k(\mathcal{F})$ and $\mathcal{M}_{\leq D, \geq d, m}^k(\mathcal{F})$
+				\State Compute a basis $(v_1,\dots,v_r)$ if the left kernel of $\mathcal{M}_{\leq D, \geq d, m}^k(\mathcal{F})$
+				\State Construct the polynomials $p_1,\dots,p_r$ corresponding to $v_i \cdot Mac_{\leq D, \geq d, m}^k(\mathcal{F})$.
 				\For{$a = (a_{k+1},\dots,a_n) \in \mathbb{F}_{2}^{n-k}$}
@@
 		This algorithm have shown it's effectiveness in practice with good implementations like Bouillaguet and Sauvage's implementation \cite {BS23} that broke records for solving boolean systems on the Fukuaka MQ challenge \cite{MQFUK}.
+Despite that, there are no proven complexity analysis for this algorithm due to it's huge number of possible parameters but we can do some estimations with admissible parameters as we will see in the next section.
 
@@
 \cite{BFSS13} with average running time $\tilde{\mathcal{O}}(2^{0.792n})$ under some algebraic assumptions like semi-regularity.
-We do not seem to have this assumption on our systems, but we know thanks to \cite{FSS11} that the degree of regularity upper-bound is lower than for semi-regular systems \textbf{A écrire dans la section de l'article car important et referencer}, but the cost of the bilinear criterion seems to be superior to that of the cost of solving \ref{complexCrit} so we will not consider this attack.
+We do not seem to have this assumption on our systems, but we know thanks to \cite{FSS11} that the degree of regularity upper-bound is lower than for semi-regular systems \ref{reg_bi}, but the cost of the bilinear criterion seems to be superior to that of the cost of solving \ref{complexCrit} so we will not consider this attack.
 
 		One type of attack is the probabilistic attacks. The best ones are Dinur's first and second algorithm \cite{Itai21A, Itai21B}.
-		Their complexity is...
+		The time complexity of Diner's first algorithm is estimated to be $2^{229.19}$.
+For the second algorithm it is $2^{221.78}$.
+Of course, these are asymptotic estimations and it can and is very probably an underestimation according to some authors like those of MQOM in their security analysis using these algorithms.
+We got these estimations from the CryptographicEstimator python library \cite{EVZ+24}.
 
-		Finally, the Crossbred algorithm...
-
-		We need to first compute the number of rows of $\mathcal{M}_{\leq D, \geq d, m}^k(\mathcal{F})$ before and the number of nonzero elements per column for our systems.
-We focus on this matrix because it's the one on which we apply block-lancszos to find vectors in its co-kernel.
+		Now we will estimate attacks based on the Crossbred algorithm. Even though we do not have tight complexity analysis of it, we can do estimations based on some algorithms and common assumptions.
+		We need to first compute the number of rows of $\mathcal{M}_{\leq D, \geq d, m}^k(\mathcal{F})$ and the number of nonzero elements per column for our systems.
+We focus on this matrix because it's the one on which we apply block-lancszos to find r vectors in its co-kernel.
 		The number of rows is bounded by $$nb\_monomials\_\mathcal{M}_{\leq D, \geq d, m}^k = \sum_{i=d+1}^{D}\sum_{j=0}^{d-i} \binom{k}{i} \cdot \binom{n-k}{j}$$
 		Finally, we saw by experiment that the number of monomials present in the our system polynomials is $\frac{n^2}{4} + \frac{n}{2} + 1$ which is the upper-bound for the number of nonzero elements per row instead of the standard $\mathcal{O}(n^2)$.
+The time complexity of this operation is
+		$$
+		\mathcal{O}\left(\frac{r}{W}max(\frac{n^2}{4} + \frac{n}{2} + 1)c^2/W, c^2\right),
+		$$
+		where $W$ is the machine word size which is 64 on most common machines nowadays and $c$ is the number of rows of $\mathcal{M}(\leq D, \geq d, m)^k$ which is bounded by \textbf{Citer article quand il sera sortie}
+		$$
+			c \leq \left(\sum_{i=0}^{D}\binom{n}{i} - \sum_{i=0}^{d-2}\binom{k}{i}\right)m.
+		$$
+		The authors of \cite{MQOMv2} noticed that if $d \geq 1$, then $\mathcal{M}_{\leq D, \geq d, m}^k$ is a dense matrix, so it's better to just apply strassen to find these vectors.
+We compute both complexities and take the minimum for our estimations.
+		After finding the vectors, comes the exhaustive search part.
+First, we compute the specialized system $\mathcal{F}*$ and compute $Mac_{\leq d, m}(\mathcal{F}*)$ and concatenate to the Macaulay matrix the polynomials $\{p_1,\dots,p_r\}$ evaluated and finally we check if the system is full rank by applying XL algorithm using block-wiedemann or just strassen or simply gaussian elimination which is done in hpXbred \cite{BS23}.
+The most computationally intensive part here is the gaussian elimination that we apply on a matrix almost square matrix of size $\sum_{i=0}^{d}\binom{n-k}{i}$ with complexity $\mathcal{O}(n^3)$ or $\mathcal{O}(n^{2.81})$ using strassen or M4RI algorithms.
+We then apply these algorithms $2^{n-k}$ times to get an estimation for the time complexity of exhaustive search part.
 
-		After finding the vectors, comes the exhaustive search part. We will
+		With all of these estimation, we get a broad estimation for the Crossbred algorithm for a admissible parameters.
+We wrote sage scripts to compute these estimation which are available in a public github repository
+		\begin{center}
+			\href{http://www.github.com/paolo/MPCitH_SBC}{http://www.github.com/paolo/MPCitH\_SBC}.
+		\end{center}
+		Our best estimations are...\textbf{A suivre}.
 
+		Finally, the simplest and best attack to our knowledge is to do exhaustive search on all of the $y$ or $x$ variables, then the specialized systems $\mathcal{F}*$ are just linear systems and we could just apply gaussian elimination on it in $\mathcal{O}(n^3)$ operations.
+The system is of size $257 \times 129$.
+If the system is full rank, we give the solution, else we continue the exhaustive search.
+The cost of this attack is
+		$$
+		\mathcal{O}(2^{128} \cdot 257^{3}) \approx \mathcal{O}(2^{152}).
+		$$
+		One advantage of this attack is that it's memory footprint is negligible.
 \printbibliography
@@
 \appendix
+\section{Signature protocol}
+
 \section{Security proof}\label{appendix:proof}
@@
 
+	The following proofs have been inspired by the proof of \textbf{EUF-CMA} of SDitH by Thibauld Feneuil \cite{feneu2023}.
+
 	\begin{theorem}
@@
 	\begin{proof}
-
+		\textbf{Il suffit de modeler le schema avec les énoncés de \cite{feneu2023} et hop on peut simplement utiliser ses preuves qui sont assez gnérales pour tout système de signature MPCitH avec une oneway function et une PRF.}
 	\end{proof}

### 2025-05-27 rapport.tex
ajout libfes-lite dans bib

@@
 What is remarkable is that the complexity does not implicate the number of polynomials in the system, $m$.
-There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, \textbf{libFES}.
+There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, \textit{libFES-lite} \cite{Bouilla22}.
 
@@
 		The first attack is exhaustive search which can be done in $\mathcal{O}(ln(n)2^n)$ operations in the field with negligible space complexity \cite{BCC+10}.
-		There is very optimized implementations like libFES by Charles Bouillaguet and there have been implementations on FPGA that broke real world cryptography \cite{DDVY20}.
+		There is very optimized implementations like \textit{libFES-lite} \cite{Bouilla22} by Charles Bouillaguet and there have been implementations on FPGA that broke real world cryptography \cite{DDVY20}.
 
@@
 		After finding the vectors, comes the exhaustive search part.
-First, we compute the specialized system $\mathcal{F}*$ and compute $Mac_{\leq d, m}(\mathcal{F}*)$ and concatenate to the Macaulay matrix the polynomials $\{p_1,\dots,p_r\}$ evaluated and finally we check if the system is full rank by applying XL algorithm using block-wiedemann or just strassen or simply gaussian elimination which is done in hpXbred \cite{BS23}.
+First, we compute the specialized system $\mathcal{F}*$ and compute $Mac_{\leq d, m}(\mathcal{F}*)$ and concatenate to the Macaulay matrix the polynomials $\{p_1,\dots,p_r\}$ evaluated and finally we check if the system is full rank by applying XL algorithm using block-wiedemann or just strassen or simply gaussian elimination which is done in \textit{hpXbred} \cite{BS23}.
 The most computationally intensive part here is the gaussian elimination that we apply on a matrix almost square matrix of size $\sum_{i=0}^{d}\binom{n-k}{i}$ with complexity $\mathcal{O}(n^3)$ or $\mathcal{O}(n^{2.81})$ using strassen or M4RI algorithms.
@@
 
-	The following proofs have been inspired by the proof of \textbf{EUF-CMA} of SDitH by Thibauld Feneuil \cite{feneu2023}.
+	The following proofs have been inspired by the proof of \textbf{EUF-CMA} of a generic 5-round protocols by Thibauld Feneuil \cite{feneu2023}.
 

### 2025-05-27 rapport.tex
phase 1 presque terminée

@@
 \usepackage{hyperref}
+\usepackage{tcolorbox}
+\usepackage{enumitem}
+\usepackage{fancyvrb}
+\usepackage{multicol}
+\usepackage{stmaryrd}
 
@@
 \section{Signature protocol}
+	Here we will show to signature protocol adapted from the identification protocol from \cite{HJ23} in the same format as in \cite{feneu2023} as a 5-round protocol to be able to apply his theorem on EUF-CMA security proof of such protocol.
+
+		\begin{tcolorbox}[title=Signature Scheme -- Signing algorithm, colback=white, colframe=black]
+			\textbf{Inputs:} A public key $x = (\vec{u}, \vec{v})$, a private key $w = (\vec{x}, \vec{y}) \in \text{NSBC}[\vec{u}, \vec{v}]$ and a message $m \in \{0,1\}^*$.
+
+			\vspace{0.5em}
+			\textbf{Phase 0: Initialization.}
+			\begin{enumerate}[label=\arabic*.]
+				\item Sample a random salt $\texttt{salt} \leftarrow \{0,1\}^{\lambda}$.
+				\item Sample a root seed $\rho \leftarrow \{0,1\}^{\lambda}$.
+			\end{enumerate}
+
+			\vspace{0.5em}
+			\textbf{Phase 1: Preparation of the MPC-in-the-Head inputs.} For each iteration $e \in [1:\tau]$:
+			\begin{enumerate}[label=\arabic*.]
+				\item Derive randomness $r^{[e]}_{w}, r^{[e]}_r, r^{[e]}_{\psi}$ from the root seed $\rho$: \\
+				\quad $(r^{[e]}_w, r^{[e]}_r, r^{[e]}_{\psi}) \leftarrow \text{PRG}(\texttt{salt}, e, \rho)$.
+
+				\item Share the witness $w$ into an $(N, N)$-threshold linear secret sharing $[w^{[e]}_i]$: \\
+				\quad $[X_1^{[e]}], [X_2^{[e]}], [Y_1^{[e]}], [Y_2^{[e]}], [R_A^{[e]}], [R_B^{[e]}] \leftarrow \text{Share}(w; r^{[e]}_w)$ \#The shares are in $\mathbb{F}_{q^k}$. \\
+				\quad $[\vec{R}_y^{[e]}], [\vec{R}_y^{[e]}] \leftarrow \text{Share}(w; r^{[e]}_w)$ \#The shares are in $(\mathbb{F}_{q})^n$.
+
+				\item Compute $X_1, X_2, Y_1, Y_2$ \\
+				\quad $X_1 \leftarrow \sum_{i=1}^{N} [X_1^{[e]}]^{\llbracket i \rrbracket}, X_2 \leftarrow \sum_{i=1}^{N} [X_2^{[e]}]^{\llbracket i \rrbracket}, Y_1 \leftarrow \sum_{i=1}^{N} [Y_1^{[e]}]^{\llbracket i \rrbracket}, Y_2 \leftarrow \sum_{i=1}^{N} [Y_2^{[e]}]^{\llbracket i \rrbracket}$
+
+				\item Compute $A^{[e]}$ and $B^{[e]}$ and share them: \\
+				\quad $A \leftarrow X_1Y_1 - X_2Y_2$. \\
+				\quad $B \leftarrow X_1(\vec{v} \cdot \vec{y}) + Y_1(\vec{u} \cdot \vec{x}) - X_2(\vec{u} \cdot \vec{y}) - Y_2(\vec{v} \cdot \vec{x})$. \\
+				\quad $\delta_A \leftarrow A - \sum_{i=1}^{N} [R_A^{[e]}]^{\llbracket i \rrbracket}$. \\
+				\quad $\delta_B \leftarrow B - \sum_{i=1}^{N} [R_B^{[e]}]^{\llbracket i \rrbracket}$. \\
+				\quad $\vec{\delta}_x \leftarrow A - \sum_{i=1}^{N} [\vec{R}_x^{[e]}]^{\llbracket i \rrbracket}$. \\
+				\quad $\vec{\delta}_y \leftarrow A - \sum_{i=1}^{N} [\vec{R}_y^{[e]}]^{\llbracket i \rrbracket}$.
+
+				\item Compute the commitments \\
+				\quad $\texttt{com}^{[e]}_i := \text{Hash}_0(\texttt{salt}, e, i, [w^{[e]}_i], [g^{[e]}_i])$ \\
+				for all $i \in [1:N]$, and compute the Merkle root \\
+				\quad $\tilde{h}^{[e]} := \text{MerkleTree}(\texttt{com}^{[e]}_1, \dots, \texttt{com}^{[e]}_N)$.
+			\end{enumerate}
+
+			\vspace{0.5em}
+			\textbf{Phase 2: First challenge (randomness for the MPC protocol).}
+			\begin{enumerate}[label=\arabic*.]
+				\item Compute $h_1 := \text{Hash}(m, \texttt{salt}, \tilde{h}^{[1]}, \dots, \tilde{h}^{[\tau]})$.
+				\item Expand $h_1$ as $(e^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h_1)$.
+			\end{enumerate}
+
+			\vspace{0.5em}
+			\textbf{Phase 3: Simulation of the MPC protocol.} For each $e \in [1:\tau]$:
+			\begin{enumerate}[label=\arabic*.]
+				\item Computes, for $i \in S$, \\
+				\quad $[\alpha^{[e]}_i] := \varphi_{x,e,f}([w^{[e]}_i], [g^{[e]}_i])$ \\
+				and recomposes $\alpha^{[e]}$. \\
+
+				\textit{This step is repeated as many times as in the MPC protocol (cf Protocol 11).}
+			\end{enumerate}
+
+			\vspace{0.5em}
+			\textbf{Phase 4: Second challenge (parties to be opened).}
+			\begin{enumerate}[label=\arabic*.]
+				\item Compute $h_2 := \text{Hash}(\texttt{salt}, h_1, [\alpha^{[1]}]_{S}, \dots, [\alpha^{[\tau]}]_{S})$.
+				\item Expand $h_2$ as $(r^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h_2)$.
+			\end{enumerate}
+
+			\vspace{0.5em}
+			\textbf{Phase 5: Building of the signature.} Output the signature $\sigma$ built as:
+
+			\[
+			\sigma := \texttt{salt} \parallel h_1 \parallel h_2 \parallel (([w^{[e]}_i], [g^{[e]}_i])_{i \in r^{[e]}}, \texttt{auth}^{[e]}, \alpha^{[e]})_{e \in [1:\tau]}
+			\]
+
+			where $\texttt{auth}^{[e]}$ is the authentication path for $\{\texttt{com}^{[e]}_i\}_{i \in r^{[e]}}$ w.r.t. Merkle root $\tilde{h}^{[e]}$ and $i^{[e]} \notin r^{[e]}$.
+		\end{tcolorbox}
 
@@
 		\textbf{Il suffit de modeler le schema avec les énoncés de \cite{feneu2023} et hop on peut simplement utiliser ses preuves qui sont assez gnérales pour tout système de signature MPCitH avec une oneway function et une PRF.}
+
 	\end{proof}

### 2025-05-28 rapport.tex
ajout théorème thibauld feneuil

@@
 \usepackage{hyperref}
-\usepackage{tcolorbox}
+\usepackage[most]{tcolorbox}
 \usepackage{enumitem}
@@
 
-		\begin{tcolorbox}[title=Signature Scheme -- Signing algorithm, colback=white, colframe=black]
+		\begin{tcolorbox}[title=Protocol 15: Signing Algorithm, breakable, enhanced, colback=white, colframe=black]
 			\textbf{Inputs:} A public key $x = (\vec{u}, \vec{v})$, a private key $w = (\vec{x}, \vec{y}) \in \text{NSBC}[\vec{u}, \vec{v}]$ and a message $m \in \{0,1\}^*$.
@@
 
-				\item Compute $X_1, X_2, Y_1, Y_2$ \\
-				\quad $X_1 \leftarrow \sum_{i=1}^{N} [X_1^{[e]}]^{\llbracket i \rrbracket}, X_2 \leftarrow \sum_{i=1}^{N} [X_2^{[e]}]^{\llbracket i \rrbracket}, Y_1 \leftarrow \sum_{i=1}^{N} [Y_1^{[e]}]^{\llbracket i \rrbracket}, Y_2 \leftarrow \sum_{i=1}^{N} [Y_2^{[e]}]^{\llbracket i \rrbracket}$
+				\item Compute $X_1^{[e]}, X_2^{[e]}, Y_1^{[e]}, Y_2^{[e]}$ \\
+				\quad $X_1^{[e]} \leftarrow \sum_{i=1}^{N} [X_1^{[e]}]^{\llbracket i \rrbracket}, X_2^{[e]} \leftarrow \sum_{i=1}^{N} [X_2^{[e]}]^{\llbracket i \rrbracket}, Y_1^{[e]} \leftarrow \sum_{i=1}^{N} [Y_1^{[e]}]^{\llbracket i \rrbracket}, Y_2^{[e]} \leftarrow \sum_{i=1}^{N} [Y_2^{[e]}]^{\llbracket i \rrbracket}$
 
 				\item Compute $A^{[e]}$ and $B^{[e]}$ and share them: \\
-				\quad $A \leftarrow X_1Y_1 - X_2Y_2$. \\
-				\quad $B \leftarrow X_1(\vec{v} \cdot \vec{y}) + Y_1(\vec{u} \cdot \vec{x}) - X_2(\vec{u} \cdot \vec{y}) - Y_2(\vec{v} \cdot \vec{x})$. \\
-				\quad $\delta_A \leftarrow A - \sum_{i=1}^{N} [R_A^{[e]}]^{\llbracket i \rrbracket}$. \\
-				\quad $\delta_B \leftarrow B - \sum_{i=1}^{N} [R_B^{[e]}]^{\llbracket i \rrbracket}$. \\
-				\quad $\vec{\delta}_x \leftarrow A - \sum_{i=1}^{N} [\vec{R}_x^{[e]}]^{\llbracket i \rrbracket}$. \\
-				\quad $\vec{\delta}_y \leftarrow A - \sum_{i=1}^{N} [\vec{R}_y^{[e]}]^{\llbracket i \rrbracket}$.
+				\quad $A^{[e]} \leftarrow X_1Y_1 - X_2Y_2$. \\
+				\quad $B^{[e]} \leftarrow X_1(\vec{v} \cdot \vec{y}) + Y_1(\vec{u} \cdot \vec{x}) - X_2(\vec{u} \cdot \vec{y}) - Y_2(\vec{v} \cdot \vec{x})$. \\
+				\quad $\delta_A^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [R_A^{[e]}]^{\llbracket i \rrbracket}$. \\
+				\quad $\delta_B^{[e]} \leftarrow B^{[e]} - \sum_{i=1}^{N} [R_B^{[e]}]^{\llbracket i \rrbracket}$. \\
+				\quad $\vec{\delta}_x^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [\vec{R}_x^{[e]}]^{\llbracket i \rrbracket}$. \\
+				\quad $\vec{\delta}_y^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [\vec{R}_y^{[e]}]^{\llbracket i \rrbracket}$.
+
+				\item Compute random value $t_0^{[e]}$ and share it. \\
+				\quad $[t_0^{[e]}]^{\llbracket i \rrbracket} \leftarrow Hash_0(i || [vec{R}_x^{[e]}]^{\llbracket i \rrbracket}, [\vec{R}_y^{[e]}]^{\llbracket i \rrbracket}, [X_1^{[e]}]^{\llbracket i \rrbracket}, [X_2^{[e]}]^{\llbracket i \rrbracket}, [Y_1^{[e]}]^{\llbracket i \rrbracket}, [Y_2^{[e]}]^{\llbracket i \rrbracket},$ \\ $[R_A^{[e]}]^{\llbracket i \rrbracket}, [R_B^{[e]}]^{\llbracket i \rrbracket} || \vec{\delta}_x^{[e]}, \vec{\delta}_y^{[e]}, \delta_A^{[e]}, \delta_B^{[e]})$.
+\\
+				\quad $t_0^{[e]} \leftarrow \sum_{i=1}^{N} [t_0^{[e]}]^{\llbracket i \rrbracket}$.
 
@@
 
+	\begin{theorem}[\cite{feneu2023}]
+		Let $\text{Hash}_0, \text{Hash}_1, \text{Hash}_m$ and $\text{Expand}$ be modeled as random oracles, and let $(N, \tau, \lambda, p)$ be parameters of the signature scheme.
+Let $\mathcal{A}$ be an adversary against the \textbf{EUF-CMA} security of the scheme running in time $t_{\mathcal{A}}$ and making a total of $Q_{RO}$ random oracle queries and $Q_{Sig}$ signing queries.
+Assuming that $F$ is an $(t_{\mathcal{A}}, \epsilon_{OWF})$-hard one-way function and that PRG is a $(t_{\mathcal{A}}, \epsilon_{PRG})$-secure psuedorandom generator then $\mathcal{A}$'s advantage in the \textbf{EUF-CMA} game is
+		$$
+			\epsilon_{EUF-CMA} \leq \epsilon_{OWF} + \epsilon_{PRG} + \frac{(\tau N + 2)Q^2}{2^{2\lambda}} + Pr[X+Y = \tau],
+		$$
+		where $Q = Q_{RO} + N_{Hash} \cdot Q_{Sig}$ with $N_{Hash} = 2 + \tau(2N - 1)$ the number of hash computations in a signature generation, and where $X, Y$ are defined as
+		\begin{itemize}
+			\item $X = max_{q1 \in \mathcal{Q}_1} \{X_{q1}\} \text{ with } X_{q1} \sim \mathcal{B}\left(\tau, \binom{N}{l+1}p\right)$ and
+			\item $Y = max_{q2 \in \mathcal{Q}_2} \{Y_{q2}\} \text{ with } Y_{q2} \sim \mathcal{B}\left(\tau - X, 1 / \binom{N}{l}\right)$,
+		\end{itemize}
+		where $\mathcal{B}(n_0, p_0)$ denotes the binomial distribution with $n_0$ the number of trials and $p_0$ the success probability of each trial.
+	\end{theorem}
+
 	\begin{theorem}

### 2025-05-28 rapport.tex
minor updates
memory footprint ajouté

@@
 			\begin{itemize}
-				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if its Hilbert's series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^{d_i})/(1-z)^n.$$ $S_{m, n}(z)$ is called the generating series of $f_1,\dots,f_m$.
+				\item The sequence $f_1,\dots,f_m$ is semi-regular if and only if its Hilbert series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = \prod_{i=1}^{m}(1-z^{d_i})/(1-z)^n.$$ $S_{m, n}(z)$ is called the generating series of $f_1,\dots,f_m$.
 				\item The sequence $f_1,\dots,f_m$ is semi-regular on $\mathbb{F}_2$, if and only if its Hilbert series is the series $$S_{m, n}(z) = \sum_{d \geq 0}h_{d, m}(n)z^d = (1+z)^n/\prod_{i=1}^{m}(1+z^{d_i}).$$ $S_{m, n}(z)$ is called the generating series of $f_1,\dots,f_m$.
@@
 
-		Huth and Joux cite a complexity result in \cite{FSS11} to argue their parameters choice, we will dive deeper into this article's results in a later section.
+		Huth and Joux cite a complexity result in \cite{FSS11} to argue their parameters choice, we will dive deeper into this article  results in a later section.
 
@@
 $MaxMinors(jac_x(F_{i-1}))$)means that we have to compute the determinants of all the sub-matrices of size $n_y \times n_y$ (resp.
-$n_x \times n_x$) and for a matrix $M$ of size $l \times c$ with $l > c$ there are $\binom{l}{c}$ sub-matrices of size $c \times c$, hence the term $\binom{i}{n_y}n_y^{\omega}$ in the sum if we suppose the complexity of computing the determinant of a square multivariate polynomial matrix of size $c \times c$ to be $\mathcal{O}(c^\omega)$, which is not the case but let's suppose the best case ever.
+$n_x \times n_x$) and for a matrix $M$ of size $l \times c$ with $l > c$ there are $\binom{l}{c}$ sub-matrices of size $c \times c$, hence the term $\binom{i}{n_y}n_y^{\omega}$ in the sum if we suppose the complexity of computing the determinant of a square multivariate polynomial matrix of size $c \times c$ to be $\mathcal{O}(c^\omega)$, which is not the case but let us suppose the best case ever.
 \textbf{faudrait trouver la vraie complexité de ça pour être plus juste.}
@@
 
-		 It's an asymptotic complexity which doesn't take into account the constant factor in implementations that could slow down a lot the algorithm.
+		 It is an asymptotic complexity which does not take into account the constant factor in implementations that could slow down a lot the algorithm.
 		 We also recall that in our case with NSBC, $n_x = n_y$. So we get
@@
 \cite{LPT+17} gave an algorithm that finds the solution to a polynomial system with complexity $\tilde{\mathcal{O}}(2^{0.8765n})$ which is better than exhaustive search.
-Even though it's asymptotically better, there are no known implementations because of the huge constant factors present in the complexity that makes it worse than exhaustive search for any systems that can be solved on modern computers.
+Even though it is asymptotically better, there are no known implementations because of the huge constant factors present in the complexity that makes it worse than exhaustive search for any systems that can be solved on modern computers.
 New algorithms have been proposed by Itai Dinur \cite{Itai21A, Itai21B} where the complexity is brought down to $\tilde{\mathcal{O}}(2^{0.6943n})$, but still no implementation known to the authors.
@@
 		\begin{itemize}
-			\item \textit{Unbreakability}: the attacker recovers the secret key from the verification key(or public key). It's denoted \textbf{UB}.
+			\item \textit{Unbreakability}: the attacker recovers the secret key from the verification key(or public key). It is denoted \textbf{UB}.
 			\item \textit{Universal Unforgeability}: the attacker can produce a valid signature of any message in the message space without necessarily recovering the verification key.
-It's denoted \textbf{UUF}.
+It is denoted \textbf{UUF}.
 			\item \textit{Existential Unforgeability}: the attacker creates a message and produces a valid signature of it without necessarily having a choice for the message.
-			It's denoted \textbf{EUF}.
+			It is denoted \textbf{EUF}.
 		\end{itemize}
@@
 		\begin{itemize}
-			\item \textit{Key-Only Attacks}: The adversary has only access to the verification keys, which is the simplest and unavoidable scenario. It's denoted \textbf{KOA}.
-			\item \textit{Known-Message Attacks}: The adversary has access to a set of messages and their valid signatures. It's denoted \textbf{KMA}. 
-			\item \textit{Chosen-Message Attacks}: The adversary can probe an oracle that can sign any message of his choice and he can request as much as he wants. It's denoted \textbf{CMA}.
+			\item \textit{Key-Only Attacks}: The adversary has only access to the verification keys, which is the simplest and unavoidable scenario. It is denoted \textbf{KOA}.
+			\item \textit{Known-Message Attacks}: The adversary has access to a set of messages and their valid signatures. It is denoted \textbf{KMA}. 
+			\item \textit{Chosen-Message Attacks}: The adversary can probe an oracle that can sign any message of his choice and he can request as much as he wants. It is denoted \textbf{CMA}.
 		\end{itemize}
-		The definitions are inspired from Damien Vergnaud's lectures at Sorbonne-Université.
+		The definitions are inspired from Professor Damien Vergnaud's lectures at Sorbonne-Université.
 
@@
 		First we need to enumerate the parameters for the signature scheme.
-The finite field used is $\mathbb{F}_2$ and the vector length $n$ is 130 but we consider the NSBC version so it's 128 in practice with the 2 last coordinates implicitly defined.
+The finite field used is $\mathbb{F}_2$ and the vector length $n$ is 130 but we consider the NSBC version so it is 128 in practice with the 2 last coordinates implicitly defined.
 The field extension used $k$ is 257 which is prime to avoid subfield attacks \cite{Peb24}.
@@
 So we estimate the total number of arithmetic operations of the block-Wiedemann algorithm to be $(2(1 + \tilde{n}/\tilde{m})N(nnz + \tilde{m}N))$.
+In \cite{NNY17}, which describes the only and best implementation of XL using also a good implementation of the block-wiedemann algorithm, they suggest $\tilde{n} = \tilde{m} = 512$ for $\mathbb{F}_2$ and that is what we used for our estimations.
 \textbf{A préciser et comprendre si possible}.
@@
 		We need to first compute the number of rows of $\mathcal{M}_{\leq D, \geq d, m}^k(\mathcal{F})$ and the number of nonzero elements per column for our systems.
-We focus on this matrix because it's the one on which we apply block-lancszos to find r vectors in its co-kernel.
+We focus on this matrix because it is the one on which we apply block-lancszos to find r vectors in its co-kernel.
 		The number of rows is bounded by $$nb\_monomials\_\mathcal{M}_{\leq D, \geq d, m}^k = \sum_{i=d+1}^{D}\sum_{j=0}^{d-i} \binom{k}{i} \cdot \binom{n-k}{j}$$
@@
 		$$
-		The authors of \cite{MQOMv2} noticed that if $d \geq 1$, then $\mathcal{M}_{\leq D, \geq d, m}^k$ is a dense matrix, so it's better to just apply strassen to find these vectors.
+		The authors of \cite{MQOMv2} noticed that if $d \geq 1$, then $\mathcal{M}_{\leq D, \geq d, m}^k$ is a dense matrix, so it is better to just apply strassen to find these vectors.
 We compute both complexities and take the minimum for our estimations.
@@
 
-		Finally, the simplest and best attack to our knowledge is to do exhaustive search on all of the $y$ or $x$ variables, then the specialized systems $\mathcal{F}*$ are just linear systems and we could just apply gaussian elimination on it in $\mathcal{O}(n^3)$ operations.
+		Finally, the simplest and best attack to our knowledge is to do exhaustive search on all of the $y$ or $x$ variables, then the specialized systems $\mathcal{F}*$ are just linear systems and we could just apply gaussian elimination or system solving using M4RI \cite{M4RI} on it in $\mathcal{O}(n^{log(7)})$ operations.
 The system is of size $257 \times 129$.
@@
 		$$
-		\mathcal{O}(2^{128} \cdot 257^{3}) \approx \mathcal{O}(2^{152}).
+		\mathcal{O}(2^{128} \cdot 257^{log(7)}) \approx \mathcal{O}(2^{150.5}).
 		$$
 		One advantage of this attack is that it's memory footprint is negligible.
+
+		\subsection{Memory footprint of the attacks}
+
+		One important thing to consider for practical implementations of attacks is the memory footprint.
+There exist a lot of algorithms that are either galactic, i.e.
+that have good asymptotic complexity but will never be used in practice because of big-Oh factors, famous examples of galactic algorithms are matrix multiplication in $\mathcal{O}(n^2.371552)$ \cite{WXXZ23} or integer multiplication in $\mathcal{O}(nlog(n))$ \cite{HH21}.
+Other algorithms can be very practical but necessitate a huge amount of memory like birthday attacks or rainbow table attacks on hash functions thus making them impractical.
+
+		In this section we discuss the memory needed by the Crossbred algorithm to attack the SBC signature scheme.
+
+		First, we need to construct $Mac_{\leq D, \geq d, m}^k$ and $\mathcal{M}_{\leq D, \geq d, m}^k$, but the second is by definition a sub matrix of the first, so we would only need to construct the first.
+We can construct $\mathcal{M}_{\leq D, \geq d, m}^k$ by only manipulating pointers on $Mac_{\leq D, \geq d, m}^k$ for example.
+So, the important memory footprint that we need to consider in the first part is the one of $Mac_{\leq D, \geq d, m}^k$.
+Let us consider the following cases
+		\begin{itemize}
+			\item We use the sparse representation.
+Then we can use the CSR representation which needs to store an array of column index of the non-zero $col\_index$ elements and an array containing the number of non-zero elements per row $row\_index$.
+We do not need a third array of coefficients because all the non-zero coefficients are equal to 1.
+			The size of $col\_index$ is $nnz$ and the size of $row\_index$ is the number of rows of the matrix.
+			So, the memory footprint for the storage of $Mac_{\leq D, \geq d, m}^k$ is in our case
+			$$
+				memory\_footprint \leq \left(\left(\frac{n^2}{4} + \frac{n}{2} + 1\right) + 1\right) \times \left(\sum_{i=0}^{D}\binom{n}{i} - \sum_{i=0}^{d-2}\binom{k}{i}\right)m \times 4 \text{ bytes}.
+			$$
+			One important thing to take into account is the size of the coefficients in the arrays.
+They have to be at least of size $log\left(\sum_{i=0}^{D} \binom{n}{i}\right)$ \textbf{à affiner comme borne}, because it's the number of columns of $Mac_{\leq D, \geq d, m}^k$.
+So the most accurate estimation is rather
+			$$
+				memory\_footprint \leq \left(\left(\frac{n^2}{4} + \frac{n}{2} + 1\right) + 1\right) \times \left(\sum_{i=0}^{D}\binom{n}{i} - \sum_{i=0}^{d-2}\binom{k}{i}\right)m \times log\left(\sum_{i=0}^{D} \binom{n}{i}\right) \text{ bytes}.
+			$$
+			\item If we use dense representation, then we could
+		\end{itemize}
+
 \printbibliography
@@
 		We need to show that if a $\mathcal{PPT}$ adversary $\mathcal{A}$ can retrieve the private key from the public key, then we can break the SBC function.
-		It's straightforward to show, because if $\mathcal{A}$ can retrieve the private key $(x, y) \in \mathbb{F}_{q}^{2n}$ of a given valid public key $(u, v) \in \mathbb{F}_{q^k}^{2n}$ then no need to convert anything because $(x, y) \in \text{SBC}[u, v]$ directly.
+		It is straightforward to show, because if $\mathcal{A}$ can retrieve the private key $(x, y) \in \mathbb{F}_{q}^{2n}$ of a given valid public key $(u, v) \in \mathbb{F}_{q^k}^{2n}$ then no need to convert anything because $(x, y) \in \text{SBC}[u, v]$ directly.
 So the signature scheme is \textbf{UB-KOA} if the SBC function is ($\tau, \epsilon$)-hard.

### 2025-05-30 rapport.tex
fini complexité spatiale crossbred

@@
 			$$
-				memory\_footprint \leq \left(\left(\frac{n^2}{4} + \frac{n}{2} + 1\right) + 1\right) \times \left(\sum_{i=0}^{D}\binom{n}{i} - \sum_{i=0}^{d-2}\binom{k}{i}\right)m \times log\left(\sum_{i=0}^{D} \binom{n}{i}\right) \text{ bytes}.
+				memory\_footprint \leq \left(\left(\frac{n^2}{4} + \frac{n}{2} + 1\right) + 1\right) \times \left(\sum_{i=0}^{D}\binom{n}{i} - \sum_{i=0}^{d-2}\binom{k}{i}\right)m \times \lceil \lceil log\left(\sum_{i=0}^{D} \binom{n}{i}\right) \rceil \times \frac{1}{8} \rceil \text{ bytes}.
 			$$
-			\item If we use dense representation, then we could
+			\begin{example}
+				If we have $D = 12, d = 10, n = 256, m = 257, k=176$, the memory\_footprint would be aboult
+				$$
+					\left(\left( \frac{256^2}{4} + \frac{256}{2} + 1 \right) + 1 \right) \times \left(\sum_{i=0}^{12} \binom{256}{i} - \sum_{i=0}^{8} \binom{176}{i}\right) \times 257 \times \lceil \lceil log\left(\sum_{i=0}^{12} \binom{256}{i}\right) \rceil \times \frac{1}{8} \rceil = $$ $$ 5112102637167689483266493844 \approx 2^{92} \text{ bytes}.
+				$$
+			\end{example}
+			\item If we use dense representation, then we could represent each row as a single integer of size $log\left(\sum_{i=0}^{D} \binom{n}{i}\right)$ bits and we would have $\left(\sum_{i=0}^{D}\binom{n}{i} - \sum_{i=0}^{d-2}\binom{k}{i}\right)m$ of these integers, which is the number of rows.
+So the memory\_footprint would be
+			$$
+				memory\_footprint \leq \lceil \lceil log\left(\sum_{i=0}^{D} \binom{n}{i}\right) \rceil \times \frac{1}{8} \rceil \times \left(\sum_{i=0}^{D}\binom{n}{i} - \sum_{i=0}^{d-2}\binom{k}{i}\right)m \text{ bytes}.
+			$$
+			\begin{example}
+				If we apply the same last example, this time the footprint would be
+				$$
+					\lceil \lceil log\left(\sum_{i=0}^{12} \binom{256}{i}\right) \rceil \times \frac{1}{8} \rceil \times \left(\sum_{i=0}^{12}\binom{256}{i} - \sum_{i=0}^{8}\binom{176}{i}\right) \times 257 = $$ $$ 309561743803299593270346 \approx 2^{78} \text{ bytes}.
+				$$
+			\end{example}
 		\end{itemize}

### 2025-05-30 rapport.tex
petite remarque

@@
+% !TeX spellcheck = <none>
 % !TXS template
@@
 
+		These are theoretical estimations and in practice it can be much more.
+For example, to represent a 67 bit number using the bignum implementation in the GMP library, we would need at least 32 bytes instead of the 9 that we estimated because of memory alignement and the total coast of the structure, but this can be optimized to use less memory for an array usage.
+
+		\subsubsection{Space complexity of other attacks}
+
 \printbibliography

### 2025-06-02 rapport.tex
minor changes

@@
 There exist a lot of algorithms that are either galactic, i.e.
-that have good asymptotic complexity but will never be used in practice because of big-Oh factors, famous examples of galactic algorithms are matrix multiplication in $\mathcal{O}(n^2.371552)$ \cite{WXXZ23} or integer multiplication in $\mathcal{O}(nlog(n))$ \cite{HH21}.
+that have good asymptotic complexity but will never be used in practice because of big-Oh factors, famous examples of galactic algorithms are matrix multiplication in $\mathcal{O}(n^{2.371552})$ \cite{WXXZ23} or integer multiplication in $\mathcal{O}(nlog(n))$ \cite{HH21}.
 Other algorithms can be very practical but necessitate a huge amount of memory like birthday attacks or rainbow table attacks on hash functions thus making them impractical.

### 2025-06-04 rapport.tex
bcp de changements à venir

@@
 		$\vec{x} = (\vec{x'}, 1, 0), \vec{y} = (\vec{y'}, 0, 1)$ to ensure that they are not colinear without needing to regenerate each time they are not.
-We also denote $(\vec{x}, \vec{y}) \in NSBC[\vec{u}, \vec{v}]$ if $(\vec{x}, \vec{y})$ are solutions to the NSBC problem on $(\vec{u}, \vec{v})$.
+We also denote $(\vec{x}, \vec{y}) \in \text{NSBC}[\vec{u}, \vec{v}]$ if $(\vec{x}, \vec{y})$ are solutions to the NSBC problem on $(\vec{u}, \vec{v})$.
 
@@
 
-		There exists 3 main security goals:
+		in the literature, there are three main security goals:
 		\begin{itemize}
-			\item \textit{Unbreakability}: the attacker recovers the secret key from the verification key(or public key). It is denoted \textbf{UB}.
-			\item \textit{Universal Unforgeability}: the attacker can produce a valid signature of any message in the message space without necessarily recovering the verification key.
-It is denoted \textbf{UUF}.
-			\item \textit{Existential Unforgeability}: the attacker creates a message and produces a valid signature of it without necessarily having a choice for the message.
-			It is denoted \textbf{EUF}.
+			\item \textit{Unbreakability (\textbf{UB})}: the attacker recovers the secret key from the verification key (or public key).
+			\item \textit{Universal Unforgeability (\textbf{UUF})}: the attacker can produce a valid signature of any message in the message space without necessarily recovering the verification key.
+			\item \textit{Existential Unforgeability (\textbf{EUF})}: the attacker creates a message and produces a valid signature of it without necessarily having a choice for the message.
 		\end{itemize}
-		Finally, there are 3 main adversarial models:
+		Finally, there are three main adversarial models:
 		\begin{itemize}
-			\item \textit{Key-Only Attacks}: The adversary has only access to the verification keys, which is the simplest and unavoidable scenario. It is denoted \textbf{KOA}.
-			\item \textit{Known-Message Attacks}: The adversary has access to a set of messages and their valid signatures. It is denoted \textbf{KMA}. 
-			\item \textit{Chosen-Message Attacks}: The adversary can probe an oracle that can sign any message of his choice and he can request as much as he wants. It is denoted \textbf{CMA}.
+			\item \textit{Key-Only Attacks (\textbf{KOA})}: The adversary has only access to the verification keys, which is the simplest and unavoidable scenario.
+			\item \textit{Known-Message Attacks (\textbf{KMA})}: The adversary has access to a set of messages and their valid signatures. 
+			\item \textit{Chosen-Message Attacks (\textbf{CMA})}: The adversary can probe an oracle that can sign any message of his choice and he can request as much as he wants.
 		\end{itemize}
-		The definitions are inspired from Professor Damien Vergnaud's lectures at Sorbonne-Université.
 
-		The goal of any signature scheme is to be \textbf{EUF-CMA}, which is the strongest security proof. First we prove that the schema is \textbf{UB-KOA}, which is the weakest proof but the easiest.
+		When constructing a signature scheme, the goal is to acheive \textbf{EUF-CMA} security, which is the strongest security proof.
+First we prove that the schema is \textbf{UB-KOA}, which is the weakest proof but the easiest.
 
 		\begin{lemma}\label{UB-KOA}
-			Suppose that the SBC function $\text{SBC}: \mathbb{F}_{q^k}^{2n} \rightarrow \mathbb{F}_{q}^{2n}$ is ($\tau, \epsilon$)-hard, i.e.
-there are no $\mathcal{PPT}$ $\mathcal{A}$ with running time upper-bounded by $\tau$ such that the SBC problem is hard to solve for a certain parameter set $(q, n, m)$,
+			Suppose that the SBC function $\text{SBC}: \mathbb{F}_{q^k}^{2n} \rightarrow \mathbb{F}_{q}^{2n}$ is ($\tau, \epsilon_{OWF}$)-hard, i.e.
+there are no $\mathcal{PPT}$ $\mathcal{A}$ with running time upper-bounded by $\tau$ such that the SBC problem is hard to solve for a certain parameter set $(q, n, k)$,
 			\begin{align*}
@@
 
-			Finally, suppose that the SBC problem is hard, i.e. the complexity of the best attack of the SBC problem has complexity $2^{\lambda}$ using a $\mathcal{PPT}$.
+			Finally, suppose that the SBC problem is hard, i.e. the complexity of the best attack of the SBC problem is $2^{\lambda}$.
 			Then, the SBC signature scheme is \textbf{EUF-CMA}.
@@
 of degree $\leq d$), is $\binom{n}{d}$ (resp.
-$\sum_{i=0}^{d}\binom{n}{d}$) because of the frobenius identity that reduces their number.
+$\sum_{i=0}^{d}\binom{n}{d}$) because of the Frobenius identity that reduces their number.
 A good remark is that the number of monomials of system modelisation of SBC problems contain exactly $\frac{n^2}{4} + \frac{n}{2} + 1$ monomials.
@@
 			$$
-				memory\_footprint \leq \left(\left(\frac{n^2}{4} + \frac{n}{2} + 1\right) + 1\right) \times \left(\sum_{i=0}^{D}\binom{n}{i} - \sum_{i=0}^{d-2}\binom{k}{i}\right)m \times \lceil \lceil log\left(\sum_{i=0}^{D} \binom{n}{i}\right) \rceil \times \frac{1}{8} \rceil \text{ bytes}.
+				\left(\left(\frac{n^2}{4} + \frac{n}{2} + 1\right) + 1\right) \times \left(\sum_{i=0}^{D}\binom{n}{i} - \sum_{i=0}^{d-2}\binom{k}{i}\right)m \times \lceil \lceil log\left(\sum_{i=0}^{D} \binom{n}{i}\right) \rceil \times \frac{1}{8} \rceil \text{ bytes}.
 			$$
@@
 	\begin{lemma}
-		Suppose that the SBC function $\text{SBC}: \mathbb{F}_{q^k}^{2n} \rightarrow \mathbb{F}_{q}^{2n}$ is ($\tau, \epsilon$)-hard, i.e.
+		Suppose that the SBC function $\text{SBC}: \mathbb{F}_{q^k}^{2n} \rightarrow \mathbb{F}_{q}^{2n}$ is ($\tau, \epsilon_{OWF}$)-hard, i.e.
 there are no $\mathcal{PPT}$ $\mathcal{A}$ with running time upper-bounded by $\tau$ such that the SBC problem is hard to solve for a certain parameter set $(q, n, m)$,

### 2025-06-05 rapport.tex
premier batch de corrections

@@
-% !TeX spellcheck = <none>
+% !TeX spellcheck = en_US
 % !TXS template
@@
 What is remarkable is that the complexity does not implicate the number of polynomials in the system, $m$.
-There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, \textit{libFES-lite} \cite{Bouilla22}.
+There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, \textit{libFES-lite} \cite{Bouilla22} and there have been implementations on FPGA that broke real world cryptography \cite{DDVY20}..
 
@@
 
-		The last type is a hybrid algorithms using gröbner basis algorithms and exhaustive search to find solutions.
+		The last type is a hybrid algorithms using Gröbner basis algorithms and exhaustive search to find solutions.
 The most known and the best one is BooleanSolve hybrid by Bardet, Faugère, Salvy and Spaenlehauer \cite{BFSS13} where we guess a certain number of variables $k$ and apply F5 algorithm on the specialized systems.
@@
 
-		We wrote sage scripts that automated the search of admissible parameters using the bivariate series given in \cite{VID24}.
+		We wrote SageMath scripts that automated the search of admissible parameters using the bivariate series given in \cite{VID24}.
 We searched for different key sizes to have a rough estimate on the complexity.
@@
 		\subsection{Security proofs}
-		To prove the security of a signature scheme, one can simulate a game with assumptions and prove that the scheme is not breakable with assumptions on the adversary capabilities and the security goals.
+		The security of the signature scheme is demonstrated under standard assumptions regarding the adversary’s capabilities, ensuring that the scheme meets well-defined security goals.
 
-		in the literature, there are three main security goals:
+		In the literature, there are three main security goals:
 		\begin{itemize}
@@
 			\item \textit{Known-Message Attacks (\textbf{KMA})}: The adversary has access to a set of messages and their valid signatures. 
-			\item \textit{Chosen-Message Attacks (\textbf{CMA})}: The adversary can probe an oracle that can sign any message of his choice and he can request as much as he wants.
+			\item \textit{Chosen-Message Attacks (\textbf{CMA})}: The adversary can probe an oracle that can sign any message of his choice and he can request as many signatures as he wants.
 		\end{itemize}
 
-		When constructing a signature scheme, the goal is to acheive \textbf{EUF-CMA} security, which is the strongest security proof.
-First we prove that the schema is \textbf{UB-KOA}, which is the weakest proof but the easiest.
+		In constructing a signature scheme, the objective is to achieve \textbf{EUF-CMA} security, which represents the strongest and most desirable level of security.
+As a first step, we demonstrate that the scheme satisfies the \textbf{UB-KOA} property, which is a weaker but more tractable security notion.
 
 		\begin{lemma}\label{UB-KOA}
-			Suppose that the SBC function $\text{SBC}: \mathbb{F}_{q^k}^{2n} \rightarrow \mathbb{F}_{q}^{2n}$ is ($\tau, \epsilon_{OWF}$)-hard, i.e.
-there are no $\mathcal{PPT}$ $\mathcal{A}$ with running time upper-bounded by $\tau$ such that the SBC problem is hard to solve for a certain parameter set $(q, n, k)$,
+			Let $(q,n,k)$ be a set of parameters for the NSBC problem.
+Assume that the NSBC function $\text{NSBC} :\mathbb{F}_{q^k}^{2n} \rightarrow \mathbb{F}_{q}^{2n}$ is $(\tau, \epsilon_{OWF})$-hard, meaning that no probabilistic polynomial-time ($\mathcal{PPT}$) adversary $mathbb{A}$ running in time at most $\tau$ can solve the NSBC problem with probability greater than $\epsilon_{OWF}$.
 			\begin{align*}
@@
 				\end{array}
-				\right] < \epsilon,
+				\right] < \epsilon_{OWF}.
 			\end{align*}
-			then the signature scheme is \textbf{UB-KOA}.
+			Then the signature scheme is \textbf{UB-KOA}.
 		\end{lemma}
@@
 		\begin{theorem}\label{EUF-CMA}
-			Suppose that no adversary can distinguish the PRF used by the protocol from real random bits
+			Assume that no adversary can distinguish the output of the pseudorandom function (PRF) used in the protocol from truly random bits.
+The adversary outputs a bit $b$, where $b=0$ indicates that they believe the bits originate from the PRF, and $b=1$ indicates they believe the bits are uniformly random
 			$$
@@
 
-			Finally, suppose that the SBC problem is hard, i.e. the complexity of the best attack of the SBC problem is $2^{\lambda}$.
-			Then, the SBC signature scheme is \textbf{EUF-CMA}.
+			Finally, suppose that the NSBC problem is hard, i.e. the complexity of the best attack of the NSBC problem is $2^{\lambda}$.
+			Then, the NSBC signature scheme is \textbf{EUF-CMA}.
 		\end{theorem}
@@
 So we estimate the total number of arithmetic operations of the block-Wiedemann algorithm to be $(2(1 + \tilde{n}/\tilde{m})N(nnz + \tilde{m}N))$.
-In \cite{NNY17}, which describes the only and best implementation of XL using also a good implementation of the block-wiedemann algorithm, they suggest $\tilde{n} = \tilde{m} = 512$ for $\mathbb{F}_2$ and that is what we used for our estimations.
+In \cite{NNY17}, which describes the only and best implementation of XL using also a good implementation of the block-Wiedemann algorithm, they suggest $\tilde{n} = \tilde{m} = 512$ for $\mathbb{F}_2$ and that is what we used for our estimations.
 \textbf{A préciser et comprendre si possible}.
@@
 
-		The first attack is exhaustive search which can be done in $\mathcal{O}(ln(n)2^n)$ operations in the field with negligible space complexity \cite{BCC+10}.
-		There is very optimized implementations like \textit{libFES-lite} \cite{Bouilla22} by Charles Bouillaguet and there have been implementations on FPGA that broke real world cryptography \cite{DDVY20}.
-
-		There is also the Hybrid Solve algorithm by Bardet et al.
-\cite{BFSS13} with average running time $\tilde{\mathcal{O}}(2^{0.792n})$ under some algebraic assumptions like semi-regularity.
-We do not seem to have this assumption on our systems, but we know thanks to \cite{FSS11} that the degree of regularity upper-bound is lower than for semi-regular systems \ref{reg_bi}, but the cost of the bilinear criterion seems to be superior to that of the cost of solving \ref{complexCrit} so we will not consider this attack.
-
-		One type of attack is the probabilistic attacks. The best ones are Dinur's first and second algorithm \cite{Itai21A, Itai21B}.
-		The time complexity of Diner's first algorithm is estimated to be $2^{229.19}$.
-For the second algorithm it is $2^{221.78}$.
-Of course, these are asymptotic estimations and it can and is very probably an underestimation according to some authors like those of MQOM in their security analysis using these algorithms.
-We got these estimations from the CryptographicEstimator python library \cite{EVZ+24}.
-
 		Now we will estimate attacks based on the Crossbred algorithm. Even though we do not have tight complexity analysis of it, we can do estimations based on some algorithms and common assumptions.
 		We need to first compute the number of rows of $\mathcal{M}_{\leq D, \geq d, m}^k(\mathcal{F})$ and the number of nonzero elements per column for our systems.
-We focus on this matrix because it is the one on which we apply block-lancszos to find r vectors in its co-kernel.
+We focus on this matrix because it is the one on which we apply block-Lancszos to find r vectors in its co-kernel.
 		The number of rows is bounded by $$nb\_monomials\_\mathcal{M}_{\leq D, \geq d, m}^k = \sum_{i=d+1}^{D}\sum_{j=0}^{d-i} \binom{k}{i} \cdot \binom{n-k}{j}$$
@@
 		$$
-		The authors of \cite{MQOMv2} noticed that if $d \geq 1$, then $\mathcal{M}_{\leq D, \geq d, m}^k$ is a dense matrix, so it is better to just apply strassen to find these vectors.
+		The authors of \cite{MQOMv2} noticed that if $d \geq 1$, then $\mathcal{M}_{\leq D, \geq d, m}^k$ is a dense matrix, so it is better to just apply Strassen to find these vectors.
 We compute both complexities and take the minimum for our estimations.
 		After finding the vectors, comes the exhaustive search part.
-First, we compute the specialized system $\mathcal{F}*$ and compute $Mac_{\leq d, m}(\mathcal{F}*)$ and concatenate to the Macaulay matrix the polynomials $\{p_1,\dots,p_r\}$ evaluated and finally we check if the system is full rank by applying XL algorithm using block-wiedemann or just strassen or simply gaussian elimination which is done in \textit{hpXbred} \cite{BS23}.
-The most computationally intensive part here is the gaussian elimination that we apply on a matrix almost square matrix of size $\sum_{i=0}^{d}\binom{n-k}{i}$ with complexity $\mathcal{O}(n^3)$ or $\mathcal{O}(n^{2.81})$ using strassen or M4RI algorithms.
+First, we compute the specialized system $\mathcal{F}*$ and compute $Mac_{\leq d, m}(\mathcal{F}*)$ and concatenate to the Macaulay matrix the polynomials $\{p_1,\dots,p_r\}$ evaluated and finally we check if the system is full rank by applying XL algorithm using block-Wiedemann or just Strassen or simply Gaussian elimination which is done in \textit{hpXbred} \cite{BS23}.
+The most computationally intensive part here is the Gaussian elimination that we apply on a matrix almost square matrix of size $\sum_{i=0}^{d}\binom{n-k}{i}$ with complexity $\mathcal{O}(n^3)$ or $\mathcal{O}(n^{2.81})$ using Strassen or M4RI algorithms.
 We then apply these algorithms $2^{n-k}$ times to get an estimation for the time complexity of exhaustive search part.
@@
 		With all of these estimation, we get a broad estimation for the Crossbred algorithm for a admissible parameters.
-We wrote sage scripts to compute these estimation which are available in a public github repository
+We wrote SageMath scripts to compute these estimation which are available in a public github repository
 		\begin{center}
@@
 
-		Finally, the simplest and best attack to our knowledge is to do exhaustive search on all of the $y$ or $x$ variables, then the specialized systems $\mathcal{F}*$ are just linear systems and we could just apply gaussian elimination or system solving using M4RI \cite{M4RI} on it in $\mathcal{O}(n^{log(7)})$ operations.
+		Finally, the simplest and best attack to our knowledge is to do exhaustive search on all of the $y$ or $x$ variables, then the specialized systems $\mathcal{F}*$ are just linear systems and we could just apply Gaussian elimination or system solving using M4RI \cite{M4RI} on it in $\mathcal{O}(n^{log(7)})$ operations.
 The system is of size $257 \times 129$.
@@
 
-		One important thing to consider for practical implementations of attacks is the memory footprint.
-There exist a lot of algorithms that are either galactic, i.e.
-that have good asymptotic complexity but will never be used in practice because of big-Oh factors, famous examples of galactic algorithms are matrix multiplication in $\mathcal{O}(n^{2.371552})$ \cite{WXXZ23} or integer multiplication in $\mathcal{O}(nlog(n))$ \cite{HH21}.
-Other algorithms can be very practical but necessitate a huge amount of memory like birthday attacks or rainbow table attacks on hash functions thus making them impractical.
+		One important consideration for practical implementations of cryptographic attacks is the memory footprint.
+In the literature, many algorithms are classified as \emph{galactic}, meaning that although they exhibit favorable asymptotic complexity, they are unlikely to be used in practice due to large constant factors hidden in the big-O notation.
+Well-known examples of galactic algorithms include matrix multiplication in $\mathcal{O}(n^{2.371552})$~\cite{WXXZ23} and integer multiplication in $\mathcal{O}(n \log n)$~\cite{HH21}.
+
+		On the other hand, some algorithms are practically efficient in terms of computation but require an enormous amount of memory, such as birthday attacks or rainbow table attacks on hash functions, which often renders them impractical in real-world scenarios.
 
@@
 
-		First, we need to construct $Mac_{\leq D, \geq d, m}^k$ and $\mathcal{M}_{\leq D, \geq d, m}^k$, but the second is by definition a sub matrix of the first, so we would only need to construct the first.
-We can construct $\mathcal{M}_{\leq D, \geq d, m}^k$ by only manipulating pointers on $Mac_{\leq D, \geq d, m}^k$ for example.
-So, the important memory footprint that we need to consider in the first part is the one of $Mac_{\leq D, \geq d, m}^k$.
-Let us consider the following cases
+		We first need to construct $Mac_{\leq D, \geq d, m}^k$, and then consider its submatrix $\mathcal{M}_{\leq D, \geq d, m}^k$.
+In practice, $\mathcal{M}_{\leq D, \geq d, m}^k$ can be obtained by manipulating pointers to $Mac_{\leq D, \geq d, m}^k$, without duplicating data.
+Therefore, the primary memory footprint to consider at this stage is that of $Mac_{\leq D, \geq d, m}^k$.
+
+		Let us now analyze the following cases:
 		\begin{itemize}
-			\item We use the sparse representation.
-Then we can use the CSR representation which needs to store an array of column index of the non-zero $col\_index$ elements and an array containing the number of non-zero elements per row $row\_index$.
-We do not need a third array of coefficients because all the non-zero coefficients are equal to 1.
-			The size of $col\_index$ is $nnz$ and the size of $row\_index$ is the number of rows of the matrix.
+			\item We use a sparse matrix representation based on the Compressed Sparse Row (CSR) format, which requires storing two arrays: the array of column indices of non-zero elements, denoted as \texttt{col\_index}, and the array that indicates the number of non-zero elements per row, denoted as \texttt{row\_index}.
+A third array for storing the actual coefficients is not needed, since all non-zero entries are equal to 1.
+
+			The size of \texttt{col\_index} is $nnz$ and the size of \texttt{row\_index} is the number of rows of the matrix.
 			So, the memory footprint for the storage of $Mac_{\leq D, \geq d, m}^k$ is in our case
@@
 			$$
-			\begin{example}
-				If we have $D = 12, d = 10, n = 256, m = 257, k=176$, the memory\_footprint would be aboult
+			\begin{example}\label{exempleCSR}
+				If we have $D = 12, d = 10, n = 256, m = 257, k=176$, the memory\_footprint would be:
 				$$
@@
 			\begin{example}
-				If we apply the same last example, this time the footprint would be
+				For the parameter choice in Example \ref{exempleCSR}, the memory footprint of the dense representation would be:
 				$$
@@
 \section{Security proof}\label{appendix:proof}
-	\begin{lemma}
-		Suppose that the SBC function $\text{SBC}: \mathbb{F}_{q^k}^{2n} \rightarrow \mathbb{F}_{q}^{2n}$ is ($\tau, \epsilon_{OWF}$)-hard, i.e.
-there are no $\mathcal{PPT}$ $\mathcal{A}$ with running time upper-bounded by $\tau$ such that the SBC problem is hard to solve for a certain parameter set $(q, n, m)$,
-		\begin{align*}
-			\text{Succ}(\mathcal{A}) = \mathcal{P}\left[
-			\begin{array}{l}
-				(u, v) \leftarrow \mathbb{F}_{q^k}^{2n}, \\
-				(x, y) \leftarrow \mathcal{A}(u, v), \\
-				(x, y) \in \text{SBC}[u, v]
-			\end{array}
-			\right] < \epsilon,
-		\end{align*}
-		then the signature scheme is \textbf{UB-KOA}.
-	\end{lemma}
+	Proof of Lemma \ref{UB-KOA}:
 	\begin{proof}
@@
 
-	\begin{theorem}
-		Suppose that no adversary can distinguish the PRF used by the protocol from real random bits
-		$$
-		| \mathcal{P}_{seed \in \{0,1\}^n}[b=0 | p \leftarrow PRF(seed)] - \mathcal{P}[b=1 | p \overset{{\scriptscriptstyle\$}}{\leftarrow} \{0,1\}^{2n}] | \leq \frac{1}{2}.
-		$$
-		Suppose also that the protocol uses a cryptographic hash function which is one-way and does not have collision attacks with better complexity than $\frac{1}{\lambda}$ using a $\mathcal{PPT}$, where $\lambda$ is the number security bits needed for the signature scheme.
-		Finally, suppose that the SBC problem is hard, i.e. the complexity the best attack of the SBC problem has complexity $\frac{1}{\lambda}$ using a $\mathcal{PPT}$.
-		Then, the SBC signature scheme is \textbf{EUF-CMA}.
-	\end{theorem}
+	Proof of theorem \ref{EUF-CMA}:
 	\begin{proof}

### 2025-06-05 rapport.tex
minor changes

@@
 
-	\section{Introduction}
-		The MPC in the head paradigm is a new framework introduced in \cite{IKOS07} which is in vogue recently as we saw in NIST's second round competition for post-quantum signature schemes where 5 of the 14 candidates are based on this paradigm.
-		The general idea is to use a NP-relation $\mathcal{R}(x, w)$ to obtain a ZK-protocol in which a prover $\mathcal{P}$ convinces a verifier $\mathcal{V}$ that she knows a valid witness $w$ for a given (public) value of $x$ without revealing any information about $x$.
-		In \cite{HJ23}, the authors introduce a problem called subfield bilinear collision problem (SBC) which originates from the discrete logarithm problem and constructed a post-quantum signature scheme based on this problem using the MPCitH framework.
-		The goal of this internship is to perform a cryptanalysis of this signature scheme and try to improve it.
+	\begin{abstract}
+		The \emph{MPC-in-the-Head} (MPCitH) paradigm, first introduced in~\cite{IKOS07}, has recently gained significant attention—particularly in the second round of NIST’s post-quantum signature standardization process, where 5 out of 14 candidate schemes were based on this approach.
+
+		The core idea of this paradigm is to leverage an NP-relation $\mathcal{R}(x, w)$ to construct a zero-knowledge (ZK) protocol.
+In such a protocol, a prover $\mathcal{P}$ convinces a verifier $\mathcal{V}$ that they possess a valid witness $w$ for a given public input $x$, without revealing any additional information about $w$.
+
+		In~\cite{HJ23}, the authors introduced the \emph{Subfield Bilinear Collision} (SBC) problem, which is rooted in the hardness of the discrete logarithm problem \cite{Joux13}.
+Based on this assumption, they proposed a post-quantum signature scheme constructed using the MPCitH framework.
+
+		The objective of this internship is to conduct a cryptanalysis of this scheme, assess its security and efficiency, and explore potential improvements.
+	\end{abstract}
 

### 2025-06-05 rapport.tex
rearrangement des sections

@@
 
-	\section{Preliminaries and notations}
+	\section{MPCitH and SBC signature scheme}
 		In this section, we introduce some standard notations of the field and notions that we will use.
@@
 
-		\subsection{Gröbner basis theory}
+		\subsection{SBC Problem}
+		The problem is the following: let $q$ be a prime power and two positive integers $k$, $n$.
+		\newline
+		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent over $\mathbb{F}_q$, find two non-colinear vectors $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})$$
+		where $(\cdot)$ is the dot product between 2 vectors.
+		We denote an instance of the SBC problem given by two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$ by $\text{SBC}[\vec{u}, \vec{v}]$.
+If the vectors $\vec{x},\vec{y} \in (\mathbb{F}_q)^n$ are solution of $\text{SBC}[\vec{u},\vec{v}]$, we use the notation $(\vec{x}, \vec{y}) \in \text{SBC}[\vec{u}, \vec{v}]$.
+
+		\subsection{NSBC variant}
+
+		The authors also introduce a particular version of this problem called normalized SBC (NSBC) where $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ and
+		$\vec{x} = (\vec{x'}, 1, 0), \vec{y} = (\vec{y'}, 0, 1)$ to ensure that they are not colinear without needing to regenerate each time they are not.
+We also denote $(\vec{x}, \vec{y}) \in \text{NSBC}[\vec{u}, \vec{v}]$ if $(\vec{x}, \vec{y})$ are solutions to the NSBC problem on $(\vec{u}, \vec{v})$.
+
+		\subsection{Keys and parameters estimations}
+		The public key of this signature scheme is $(\vec{u},\vec{v})$ and the private key is $(\vec{x}, \vec{y})$.
+		The signature and verification protocol use the MPCitH paradigm, they are described in detail in \cite{HJ23} with proof of soundness, correctness and zero-knowledge.
+		The authors estimate that a good parameter for NSBC problem to be computationally secure are $q = 2, n = 130, k = 257$ using complexity analysis of what they think is the best known cryptanalysis of such problem in \cite{FSS11}.
+
+		The signature protocol is described in .. with the signature and verification algorithms.
+
+	\section{Polynomial system solving}
 		Gröbner basis theory was first introduced by Buchberger in his thesis \cite{Buc} and he gave an algorithm to compute the Gröbner basis of an ideal.
@@
 		\end{proposition}
-
-	\section{SBC Signature Scheme}
-		\subsection{SBC Problem}
-		The problem is the following: let $q$ be a prime power and two positive integers $k$, $n$.
-		\newline
-		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent over $\mathbb{F}_q$, find two non-colinear vectors $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})$$
-		where $(\cdot)$ is the dot product between 2 vectors.
-		We denote an instance of the SBC problem given by two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$ by $\text{SBC}[\vec{u}, \vec{v}]$.
-If the vectors $\vec{x},\vec{y} \in (\mathbb{F}_q)^n$ are solution of $\text{SBC}[\vec{u},\vec{v}]$, we use the notation $(\vec{x}, \vec{y}) \in \text{SBC}[\vec{u}, \vec{v}]$.
-
-		\subsection{NSBC variant}
-
-		The authors also introduce a particular version of this problem called normalized SBC (NSBC) where $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ and
-		$\vec{x} = (\vec{x'}, 1, 0), \vec{y} = (\vec{y'}, 0, 1)$ to ensure that they are not colinear without needing to regenerate each time they are not.
-We also denote $(\vec{x}, \vec{y}) \in \text{NSBC}[\vec{u}, \vec{v}]$ if $(\vec{x}, \vec{y})$ are solutions to the NSBC problem on $(\vec{u}, \vec{v})$.
-
- 		\subsection{Keys and parameters estimations}
-		The public key of this signature scheme is $(\vec{u},\vec{v})$ and the private key is $(\vec{x}, \vec{y})$.
-		The signature and verification protocol use the MPCitH paradigm, they are described in detail in \cite{HJ23} with proof of soundness, correctness and zero-knowledge.
-		The authors estimate that a good parameter for NSBC problem to be computationally secure are $q = 2, n = 130, k = 257$ using complexity analysis of what they think is the best known cryptanalysis of such problem in \cite{FSS11}.
-
-		The signature protocol is described in .. with the signature and verification algorithms.
-
-		\section{Cryptanalysis of the scheme}
-		One of the known attacks given by the authors is through algebraic cryptanalysis.
-		One can model the attack by taking $g$ as the following polynomial
-		$$
-		g(x_1,...,x_n,y_1,...,y_n) := (\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) - (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x}).
-		$$
-		We consider
-		$$
-		g(x_1,..,x_{n-2},y_1,..,y_{n-2}) := (\sum_{i = 1}^{n-2}u_{i}x_{i} + u_{n-1})(\sum_{i = 1}^{n-2}v_{i}y_{i} + v_{n}) - (\sum_{i = 1}^{n-2}u_{i}y_{i} + u_{n})(\sum_{i = 1}^{n-2}v_{i}x_{i} + v_{n-1}).
-		$$
-
-		Then, one can perform a Weil descent on the polynomial $g$ because $\mathbb{F}_{q^k}$ is a $\mathbb{F}_q$-vector space and obtain the polynomial system
-
-		\begin{align}
-			g_1(x_1, \dots, x_{n-2},& y_1, \dots, y_{n-2}) = 0 \notag \\
-			&\vdots \notag \\
-			g_k(x_1, \dots, x_{n-2},& y_1, \dots, y_{n-2}) = 0 \notag
-		\end{align}
-
-		where $g_i$ are polynomials with coefficients in $\mathbb{F}_q$.
-
-		The most straightforward cryptanalysis of this type of systems is to perform a Gröbner basis algorithm like Buchberger's algorithm \cite{Buc}, Faugère's F4 \cite{F99}, F5 \cite{F02} or their multiple variants \cite{EF19}.
-
-		Huth and Joux cite a complexity result in \cite{FSS11} to argue their parameters choice, we will dive deeper into this article  results in a later section.
-
-		\subsection{Algebraic cryptanalysis}
-
-		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $H(I) - 1$, where $I$ is the ideal generated by the polynomials of the system.
-
-		\begin{theorem}[Theorem 3.2.10 in \cite{Bardet04}]
-			Let $f_1,...,f_m$ a homogeneous sequence, such that $\langle f_1,...,f_m \rangle$ is 0-dimensional and
-			$\prec$ an admissible graded monomial ordering. We have,
-			\begin{itemize}
-				\item[-] If the sequence is semi-regular, then no reductions to 0 are performed during the F5-matrix algorithm until the degree $d = H(I) - 1$
-				\item[-] If there are no reductions to 0 during the F5-matrix algorithm until the degree $D - 1$, and if the matrix of degree D is full rank and is the first matrix to have more rows than columns, then the sequence is semi-regular and its index of regularity is $H(I) = D$.
-			\end{itemize}
-		\end{theorem}
-
-		In \cite{BFS15}, the authors show that the complexity of computing the Gröbner basis of a semi-regular system $f_1,...,f_m \in \mathbb{K}[x_1,...,x_n]$ is $O\left(mD\binom{n+D-1}{D}^\omega \right)$.
-
-		We will also need the next definition of $\gamma$-strong semi-regularity.
-
-		 \begin{definition}[Definition 4 of \cite{VID24}]
-			Let $\mathcal{F} = {f_1,...,f_m}$ be a semi-regular sequence of polynomials in $\mathbb{F}_2[x_1,...,x_n]$ and let $0 \leq \gamma \leq 1$ such that $k = (1 - \gamma)n$.
-We say that this sequence is $\gamma$-strong semi-regular if
-			$$
-			\mathcal{S}(I) = \{(a_{k+1},...,a_n) \in \mathbb{F}_{2}^{n-k} |
-			$$
-			$$
-			\{f_1(x_1,...,x_k,a_{k+1},...,a_n),...,f_m(x_1,...,x_k,a_{k+1},...,a_n)\} \text{ is not semi-regular}\}
-			$$
-			has cardinality $\mathcal{O}(2^{-\gamma n})$.
-		\end{definition}
 
@@
 
-	\section{Boolean systems}
+		\subsection{Boolean systems}
 		Sometimes, the finite field used is $\mathbb{F}_2$ because it is still supposedly hard to solve but is very easy to use and manipulate on computers.
@@
 
+	\section{Contributions}
+
+		\subsection{System Modeling}
+		One of the known attacks given by the authors is through algebraic cryptanalysis.
+		One can model the attack by taking $g$ as the following polynomial
+		$$
+		g(x_1,...,x_n,y_1,...,y_n) := (\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) - (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x}).
+		$$
+		We consider
+		$$
+		g(x_1,..,x_{n-2},y_1,..,y_{n-2}) := (\sum_{i = 1}^{n-2}u_{i}x_{i} + u_{n-1})(\sum_{i = 1}^{n-2}v_{i}y_{i} + v_{n}) - (\sum_{i = 1}^{n-2}u_{i}y_{i} + u_{n})(\sum_{i = 1}^{n-2}v_{i}x_{i} + v_{n-1}).
+		$$
+
+		Then, one can perform a Weil descent on the polynomial $g$ because $\mathbb{F}_{q^k}$ is a $\mathbb{F}_q$-vector space and obtain the polynomial system
+
+		\begin{align}
+			g_1(x_1, \dots, x_{n-2},& y_1, \dots, y_{n-2}) = 0 \notag \\
+			&\vdots \notag \\
+			g_k(x_1, \dots, x_{n-2},& y_1, \dots, y_{n-2}) = 0 \notag
+		\end{align}
+
+		where $g_i$ are polynomials with coefficients in $\mathbb{F}_q$.
+
+		The most straightforward cryptanalysis of this type of systems is to perform a Gröbner basis algorithm like Buchberger's algorithm \cite{Buc}, Faugère's F4 \cite{F99}, F5 \cite{F02} or their multiple variants \cite{EF19}.
+
+		Huth and Joux cite a complexity result in \cite{FSS11} to argue their parameters choice, we will dive deeper into this article  results in a later section.
+
+		\subsection{Algebraic cryptanalysis}
+
+		Bardet showed in \cite{Bardet04} that F5 criterion removed reductions to 0 for semi-regular sequences until the degree $H(I) - 1$, where $I$ is the ideal generated by the polynomials of the system.
+
+		\begin{theorem}[Theorem 3.2.10 in \cite{Bardet04}]
+			Let $f_1,...,f_m$ a homogeneous sequence, such that $\langle f_1,...,f_m \rangle$ is 0-dimensional and
+			$\prec$ an admissible graded monomial ordering. We have,
+			\begin{itemize}
+				\item[-] If the sequence is semi-regular, then no reductions to 0 are performed during the F5-matrix algorithm until the degree $d = H(I) - 1$
+				\item[-] If there are no reductions to 0 during the F5-matrix algorithm until the degree $D - 1$, and if the matrix of degree D is full rank and is the first matrix to have more rows than columns, then the sequence is semi-regular and its index of regularity is $H(I) = D$.
+			\end{itemize}
+		\end{theorem}
+
+		In \cite{BFS15}, the authors show that the complexity of computing the Gröbner basis of a semi-regular system $f_1,...,f_m \in \mathbb{K}[x_1,...,x_n]$ is $O\left(mD\binom{n+D-1}{D}^\omega \right)$.
+
+		We will also need the next definition of $\gamma$-strong semi-regularity.
+
+		 \begin{definition}[Definition 4 of \cite{VID24}]
+			Let $\mathcal{F} = {f_1,...,f_m}$ be a semi-regular sequence of polynomials in $\mathbb{F}_2[x_1,...,x_n]$ and let $0 \leq \gamma \leq 1$ such that $k = (1 - \gamma)n$.
+We say that this sequence is $\gamma$-strong semi-regular if
+			$$
+			\mathcal{S}(I) = \{(a_{k+1},...,a_n) \in \mathbb{F}_{2}^{n-k} |
+			$$
+			$$
+			\{f_1(x_1,...,x_k,a_{k+1},...,a_n),...,f_m(x_1,...,x_k,a_{k+1},...,a_n)\} \text{ is not semi-regular}\}
+			$$
+			has cardinality $\mathcal{O}(2^{-\gamma n})$.
+		\end{definition}
+
 	\subsection{Admissible parameters for Crossbred}
@@
 		We discuss more the attacks complexity using Crossbred and different algorithms in the next section.
-
-	\section{Security analysis}
-		This section is heavily inspired from the security analysis of the MQOM signature scheme \cite{MQOMv2}.
 
@@
 
-		\subsection{Attacks against SBC signatures}
+		\subsection{Security analysis}
+		This section is heavily inspired from the security analysis of the MQOM signature scheme \cite{MQOMv2}.
+
 		First we need to enumerate the parameters for the signature scheme.
@@
 
-		\subsection{Memory footprint of the attacks}
+		\subsection{Space complexity of Crossbred}
 

### 2025-06-05 rapport.tex
ajout cars et schema et nouvelles parties

@@
 \usepackage{stmaryrd}
+\usepackage{tikz}
+\usepackage{graphicx}
+\usepackage{caption}
+\usetikzlibrary{arrows.meta, positioning}
 
 \addbibresource{biblio.bib}
+
+\newlength{\participantimgsize}
+\setlength{\participantimgsize}{1.8cm}
 
@@
 	\begin{abstract}
-		The \emph{MPC-in-the-Head} (MPCitH) paradigm, first introduced in~\cite{IKOS07}, has recently gained significant attention—particularly in the second round of NIST’s post-quantum signature standardization process, where 5 out of 14 candidate schemes were based on this approach.
+		The \emph{MPC-in-the-Head} (MPCitH) paradigm, first introduced in \cite{IKOS07}, has recently gained significant attention—particularly in the second round of NIST’s post-quantum signature standardization process, where 5 out of 14 candidate schemes were based on this approach.
 
@@
 	\section{MPCitH and SBC signature scheme}
-		In this section, we introduce some standard notations of the field and notions that we will use.
+		This section presents the MPCitH framework, the Subfield Bilinear Collision (SBC) problem, and the associated signature scheme.
 
-		\subsubsection{Notations}
+		\subsection{Multiparty Computation}
 
-		\begin{itemize}
-			\item Let $\mathbb{F}$ be a finite field and $\mathbb{K}$ a field not necessarily finite. 
-			\item We denote the multivariate polynomial ring defined over a field in $n$ variables $\mathbb{K}[x_1,\dots,x_n] = R$ where $n$ is a positive integer.
-			\item We will use the standard notation for matrix multiplication complexity for a square matrix of size $n$, $\mathcal{O}(n^\omega)$,
-			where $2 \leq \omega \leq 3$.
-			\item Let $R_{s}$ be the polynomials of total degree $s$. $I_{s} = I \cap R_{s}$ where $I$ is an ideal.
-		\end{itemize}
+		A standard example is to make construct a $(N-1)$-private arithmetic circuit among $N$ participants and they all broadcast their results. See figure \ref{fig:mpc}.
+
+		\begin{figure}[h]
+			\centering
+			\begin{tikzpicture}[
+				participant/.style={
+					thick, 
+					minimum size=2cm, 
+					inner sep=0pt
+				},
+				image/.style={
+					anchor=center
+				},
+				arrow/.style={
+					-{Latex[length=3mm, width=2mm]}, 
+					thick
+				}
+				]
+
+				% Nœuds des participants avec images
+				\node[participant] (A) at (0,0) {\includegraphics[width=\participantimgsize]{flash.png}};
+				\node[participant] (B) at (2,-3) {\includegraphics[width=\participantimgsize]{hudson.png}};
+				\node[participant] (C) at (5,-3) {\includegraphics[width=\participantimgsize]{martin.png}};
+				\node[participant] (D) at (7,0) {\includegraphics[width=\participantimgsize]{king.png}};
+				\node[participant] (E) at (2,3) {\includegraphics[width=\participantimgsize]{sally.png}};
+				\node[participant] (F) at (5,3) {\includegraphics[width=\participantimgsize]{sheriff.png}};
+
+				% Communications (flèches)
+				\draw[<->] (A) -- (B);
+				\draw[<->] (A) -- (C);
+				\draw[<->] (A) -- (D);
+				\draw[<->] (A) -- (E);
+				\draw[<->] (A) -- (F);
+				\draw[<->] (B) -- (C);
+				\draw[<->] (B) -- (D);
+				\draw[<->] (B) -- (E);
+				\draw[<->] (B) -- (F);
+				\draw[<->] (C) -- (D);
+				\draw[<->] (C) -- (E);
+				\draw[<->] (C) -- (F);
+				\draw[<->] (D) -- (E);
+				\draw[<->] (D) -- (F);
+				\draw[<->] (E) -- (F);
+
+			\end{tikzpicture}
+			\caption{MPC protocol with broadcast communication}
+			\label{fig:mpc}
+		\end{figure}
+
+		To evaluate the circuit that depends on the secret witness $w$, each party holds a secret share $[w]$.
+They perform local computations on their respective shares, then broadcast the results, which are subsequently combined to obtain the final output.
+
+		\subsection{MPCitH}
+
+		The \emph{MPC-in-the-Head} (MPCitH) framework, introduced by~\cite{IKOS07}, is a technique for constructing zero-knowledge proofs.
+It enables a prover to convince a verifier that they possess a secret witness $w$ for a given public input $x$, without revealing any information about the witness itself.
+
+		The main idea is to simulate a secure multiparty computation (MPC) protocol internally—\textit{in the head}—of the prover $\mathcal{P}$.
+This simulation involves a set of fictitious parties who jointly evaluate a circuit that depends on the witness $w$.
+The verifier $\mathcal{V}$ is then allowed to inspect a subset of the simulated views to ensure the prover is behaving honestly, while maintaining the zero-knowledge property.
+
+		To represent the secret witness $w$, a linear secret sharing scheme is used among $N$ simulated parties.
+Each party $i$ receives a random share denoted by $[w]^{\llbracket i \rrbracket}$, and the prover computes an additional correction term defined as:
+		$$
+		\delta_w = w - \sum_{i=1}^{N} [w]^{\llbracket i \rrbracket}.
+		$$
+		This ensures that the full witness can be correctly reconstructed from the shares and the correction term during the simulated execution.
+
+		Next, the prover $\mathcal{P}$ commits to the views of the $N$ fictitious parties, denoted as $View_i$, and sends these commitments to the verifier $\mathcal{V}$.
+Each view contains all messages sent and received by the corresponding party during the simulated execution of the MPC protocol.
+
+		The verifier $\mathcal{V}$ then selects a random index $i \in [1, N]$ and sends it to $\mathcal{P}$.
+In response, $\mathcal{P}$ opens the views of all parties except the $i$-th one, i.e., for all $i' \in [1, N]$ such that $i' \neq i$.
+
+		In order for this protocol to satisfy the zero-knowledge property, the underlying MPC protocol must be $(N-1)$-private, meaning that the joint view of any $N-1$ parties reveals no information about the secret witness $w$.
+
+		Since the introduction of the framework, many optimizations have been published to reduce the communication or the computation cost. We will discuss some of them.
+
+		\subsubsection{GGM tree}
 

### 2025-06-06 rapport.tex
minor changes

@@
 
-		The \emph{MPC-in-the-Head} (MPCitH) framework, introduced by~\cite{IKOS07}, is a technique for constructing zero-knowledge proofs.
+		The \emph{MPC-in-the-Head} (MPCitH) framework, introduced by \cite{IKOS07}, is a technique for constructing zero-knowledge proofs.
 It enables a prover to convince a verifier that they possess a secret witness $w$ for a given public input $x$, without revealing any information about the witness itself.
 
-		The main idea is to simulate a secure multiparty computation (MPC) protocol internally—\textit{in the head}—of the prover $\mathcal{P}$.
+		Given $\mathcal{L}$ a language in NP (i.e.
+$x \in \mathcal{L} \Longleftrightarrow \exists w, \mathcal{R}(x, w) = 1$).
+Given a public $x$, $\mathcal{P}$ wants to prove that he knows $w$ such that: $\mathcal{R}(x, w)$.
+
+		The main idea is to simulate a secure multiparty computation (MPC) protocol internally \textit{in the head} of the prover $\mathcal{P}$.
 This simulation involves a set of fictitious parties who jointly evaluate a circuit that depends on the witness $w$.

### 2025-06-11 rapport.tex
minor fix

@@
 		\begin{definition}
-			A homogeneous sequence $f_1,\dots,f_m \subset R$ is semi-regular if the following conditions are verified:
+			A zero-dimensional homogeneous sequence $f_1,\dots,f_m \subset R$ is semi-regular if the following conditions are verified:
 			\begin{itemize}
@@
 		\begin{theorem}\label{reg_bi}
-			If the system $\mathcal{F}$ is generic, then the degree of regularity of $\mathcal{V}(I)$ is upper bounded by $d_{reg} \leq min(n_x, n_y)$.
+			If the system $\mathcal{F}$ is a generic bilinear system, then the degree of regularity of $\mathcal{V}(I)$ is upper bounded by $d_{reg} \leq min(n_x, n_y)$.
 		\end{theorem}
@@
 			$$
-				| \mathcal{P}_{seed \in \{0,1\}^n}[b=0 | p \leftarrow PRF(seed)] - \mathcal{P}[b=1 | p \overset{{\scriptscriptstyle\$}}{\leftarrow} \{0,1\}^{2n}] | \leq \frac{1}{2}.
+				| \mathcal{P}_{seed \in \{0,1\}^n}[b=0 \mid p \leftarrow PRF(seed)] - \mathcal{P}[b=1 \mid p \overset{{\scriptscriptstyle\$}}{\leftarrow} \{0,1\}^{2n}] | \leq \frac{1}{2}.
 			$$

### 2025-06-12 rapport.tex
minor changes GGM

@@
 		\subsubsection{GGM tree}
+		Since most of the values used in the MPC protocol are randomly generated, a technique known as the GGM tree can be employed to transmit only a seed, rather than all the individual random values.
+
+		A GGM tree is essentially a binary tree rooted at a seed, on which a pseudorandom function (PRF) is applied.
+The PRF must produce an output of twice the bit-length of its input.
+The first half of the output serves as the root of the left subtree, and the second half as the root of the right subtree.
+This process is applied recursively until enough leaves are generated, which correspond to the required random values.
+
+		In our case, we do not transmit the seed due to the use of linear secret sharing.
+Since we typically employ an $(N-1)$-private MPC protocol, we only share the co-path of the $i$-th leaf, which remains hidden from $\mathcal{V}$.
+As a result, instead of sharing $N$ values, we only need to share $log_2(N)$ values.
+
+		\subsubsection{HyperCube technique}
 
@@
 		\begin{corollary}[Corollary 3 in \cite{FSS11}]\label{complex_bi}
-			The arithmetic complexity of computing a Gröbner basis of a generic bilinear system $f_1,...,f_{n_x + n_y}$ with the F5 algorithm is bounded by
+			The arithmetic complexity of computing a Gröbner basis of a generic affine bilinear system $f_1,...,f_{n_x + n_y}$ with the F5 algorithm is bounded by
 			$$\mathcal{O}\left(\binom{n_x + n_y + min(n_x + 1, n_y + 1)}{min(n_x + 1, n_y + 1)}^\omega\right).$$
+		\end{corollary}
+
+		\begin{corollary}
+
 		\end{corollary}
@@
 
-		\subsubsection{Space complexity of other attacks}
+	\subsection{Attacks summary}
+		\begin{table}[h!]
+		\begin{center}
+			\begin{tabular}{|c|c|c|c|}
+				\hline
+				Algorithm & Parameters & Time & Memory \\
+				\hline
+				Exhaustive search & / & 260 & 24 \\
+				\hline
+				Dinur First & / & 229 & 200 \\
+				\hline
+				Dinur Second & / & 221 & 172 \\
+				\hline
+				Crossbred & D=12, d=10, k=130 & 159 & 90 \\
+				\hline
+				Naive & / & 150 & 15 \\
+				\hline
+			\end{tabular}
+			\caption{Time and memory complexity of given algorithms in powers of 2.
+				For Exhaustive search, Dinur First and Dinur second, the estimations were obtained using CryptographicEstimator \cite{EVZ+24}}
+		\end{center}
+		\end{table}
 

### 2025-06-16 rapport.tex
minor changes

@@
 	\section{MPCitH and SBC signature scheme}
-		This section presents the MPCitH framework, the Subfield Bilinear Collision (SBC) problem, and the associated signature scheme.
-
-		\subsection{Multiparty Computation}
-
-		A standard example is to make construct a $(N-1)$-private arithmetic circuit among $N$ participants and they all broadcast their results. See figure \ref{fig:mpc}.
-
-		\begin{figure}[h]
-			\centering
-			\begin{tikzpicture}[
-				participant/.style={
-					thick, 
-					minimum size=2cm, 
-					inner sep=0pt
-				},
-				image/.style={
-					anchor=center
-				},
-				arrow/.style={
-					-{Latex[length=3mm, width=2mm]}, 
-					thick
-				}
-				]
-
-				% Nœuds des participants avec images
-				\node[participant] (A) at (0,0) {\includegraphics[width=\participantimgsize]{flash.png}};
-				\node[participant] (B) at (2,-3) {\includegraphics[width=\participantimgsize]{hudson.png}};
-				\node[participant] (C) at (5,-3) {\includegraphics[width=\participantimgsize]{martin.png}};
-				\node[participant] (D) at (7,0) {\includegraphics[width=\participantimgsize]{king.png}};
-				\node[participant] (E) at (2,3) {\includegraphics[width=\participantimgsize]{sally.png}};
-				\node[participant] (F) at (5,3) {\includegraphics[width=\participantimgsize]{sheriff.png}};
-
-				% Communications (flèches)
-				\draw[<->] (A) -- (B);
-				\draw[<->] (A) -- (C);
-				\draw[<->] (A) -- (D);
-				\draw[<->] (A) -- (E);
-				\draw[<->] (A) -- (F);
-				\draw[<->] (B) -- (C);
-				\draw[<->] (B) -- (D);
-				\draw[<->] (B) -- (E);
-				\draw[<->] (B) -- (F);
-				\draw[<->] (C) -- (D);
-				\draw[<->] (C) -- (E);
-				\draw[<->] (C) -- (F);
-				\draw[<->] (D) -- (E);
-				\draw[<->] (D) -- (F);
-				\draw[<->] (E) -- (F);
-
-			\end{tikzpicture}
-			\caption{MPC protocol with broadcast communication}
-			\label{fig:mpc}
-		\end{figure}
-
-		To evaluate the circuit that depends on the secret witness $w$, each party holds a secret share $[w]$.
-They perform local computations on their respective shares, then broadcast the results, which are subsequently combined to obtain the final output.
 
@@
 
-		Since the introduction of the framework, many optimizations have been published to reduce the communication or the computation cost. We will discuss some of them.
+		We introduce some standard notations in MPCitH protocols. 
+		\begin{itemize}
+			\item $[w]$, $w$ is a shared value.
+			\item $w^{\llbracket i \rrbracket}$ is the $i$-th share of w.
+			\item $w^[e]$ is a value used in the $e$-th round of the protocol.
+			\item $\delta_x$ is the offset for the additive sharing of $x$ s.t. $x = \delta_x + \sum_i x^{\llbracket i \rrbracket}$
+		\end{itemize}
+
+		Finally, we discuss several optimizations that have been proposed in the literature since and are now considered standard and widely used.
 
@@
 Since we typically employ an $(N-1)$-private MPC protocol, we only share the co-path of the $i$-th leaf, which remains hidden from $\mathcal{V}$.
-As a result, instead of sharing $N$ values, we only need to share $log_2(N)$ values.
+As a result, instead of sharing $N$ values, we only need to share $log(N)$ values.
 
 		\subsubsection{HyperCube technique}
+ 		This optimization technique was introduced in \cite{AGHH+22} and is commonly used in MPCitH schemes.
+This geometric approach models the $N$ shares as forming a $d$-dimensional hypercube of side length $k$.
+The often used parameters are $k = 2$ and $N = 2^d$, so the number party counts reduces to $log(N) + 1$.
+This technique often leads to MPC computations less expensive, smaller proof sizes and larger party sizes.
+However, these benefits come at the cost of increased reliance on GGM tree generation, which can become computationally expensive.
 
@@
 
-		\subsection{NSBC variant}
-
+		\textbf{NSBC variant}.
 		The authors also introduce a particular version of this problem called normalized SBC (NSBC) where $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ and
@@
 
-		The signature protocol is described in .. with the signature and verification algorithms.
+		\subsection{Multiparty protocol for the SBC problem}
+
+		\subsection{Identification Protocol}
+		We present the identification protocol, which serves as a basis for constructing the signature scheme using the Fiat-Shamir transformation.
 
@@
 				For Exhaustive search, Dinur First and Dinur second, the estimations were obtained using CryptographicEstimator \cite{EVZ+24}}
+			\label{tab_attacks}
 		\end{center}

### 2025-06-17 rapport.tex
preuve EUF-CMA

@@
 
-		\begin{tcolorbox}[title=Protocol 15: Signing Algorithm, breakable, enhanced, colback=white, colframe=black]
+		\begin{tcolorbox}[title=Protocol 15: Signing Algorithm, breakable, enhanced, colback=white, colframe=black]\label{Protocol_sig}
 			\textbf{Inputs:} A public key $x = (\vec{u}, \vec{v})$, a private key $w = (\vec{x}, \vec{y}) \in \text{NSBC}[\vec{u}, \vec{v}]$ and a message $m \in \{0,1\}^*$.
@@
 
-	\begin{theorem}[\cite{feneu2023}]
+	\begin{theorem}[\cite{feneu2023}]\label{theorem_feneu}
 		Let $\text{Hash}_0, \text{Hash}_1, \text{Hash}_m$ and $\text{Expand}$ be modeled as random oracles, and let $(N, \tau, \lambda, p)$ be parameters of the signature scheme.
 Let $\mathcal{A}$ be an adversary against the \textbf{EUF-CMA} security of the scheme running in time $t_{\mathcal{A}}$ and making a total of $Q_{RO}$ random oracle queries and $Q_{Sig}$ signing queries.
-Assuming that $F$ is an $(t_{\mathcal{A}}, \epsilon_{OWF})$-hard one-way function and that PRG is a $(t_{\mathcal{A}}, \epsilon_{PRG})$-secure psuedorandom generator then $\mathcal{A}$'s advantage in the \textbf{EUF-CMA} game is
+Assuming that $F$ is a $(t_{\mathcal{A}}, \epsilon_{OWF})$-hard one-way function and that PRG is a $(t_{\mathcal{A}}, \epsilon_{PRG})$-secure psuedorandom generator then $\mathcal{A}$'s advantage in the \textbf{EUF-CMA} game is
 		$$
@@
 	\begin{proof}
-		\textbf{Il suffit de modeler le schema avec les énoncés de \cite{feneu2023} et hop on peut simplement utiliser ses preuves qui sont assez gnérales pour tout système de signature MPCitH avec une oneway function et une PRF.}
-
+		We model the $SBC$ signature scheme \cite{HJ23} as a 5-round protocol (see Protocol~\ref{Protocol_sig}).
+By applying Theorem~\ref{theorem_feneu} from \cite{feneu2023}, and under the assumption that the underlying PRG is a $(t_{\mathcal{A}}, \epsilon_{\mathrm{PRG}})$-secure pseudorandom generator, and as established in Lemma~\ref{UB-KOA} that the $SBC$ function is a $(t_{\mathcal{A}}, \epsilon_{\mathrm{OWF}})$-secure one-way function, we conclude that the $SBC$ signature scheme achieves $EUF$-$CMA$ security.
 	\end{proof}

### 2025-06-19 rapport.tex
minor changes

@@
 \newcommand{\Mact}[3]{{\widetilde{Mac}_{#1, #2}(\mathcal{#3})}}
+\newcommand{\lir}{\llbracket i \rrbracket}
 
@@
 		To represent the secret witness $w$, a linear secret sharing scheme is used among $N$ simulated parties.
-Each party $i$ receives a random share denoted by $[w]^{\llbracket i \rrbracket}$, and the prover computes an additional correction term defined as:
+Each party $i$ receives a random share denoted by $[w]^{\lir}$, and the prover computes an additional correction term defined as:
 		$$
-		\delta_w = w - \sum_{i=1}^{N} [w]^{\llbracket i \rrbracket}.
+		\delta_w = w - \sum_{i=1}^{N} [w]^{\lir}.
 		$$
@@
 			\item $[w]$, $w$ is a shared value.
-			\item $w^{\llbracket i \rrbracket}$ is the $i$-th share of w.
+			\item $w^{\lir}$ is the $i$-th share of w.
 			\item $w^[e]$ is a value used in the $e$-th round of the protocol.
-			\item $\delta_x$ is the offset for the additive sharing of $x$ s.t. $x = \delta_x + \sum_i x^{\llbracket i \rrbracket}$
+			\item $\delta_x$ is the offset for the additive sharing of $x$ s.t. $x = \delta_x + \sum_i x^{\lir}$
 		\end{itemize}
@@
 				\item Compute $X_1^{[e]}, X_2^{[e]}, Y_1^{[e]}, Y_2^{[e]}$ \\
-				\quad $X_1^{[e]} \leftarrow \sum_{i=1}^{N} [X_1^{[e]}]^{\llbracket i \rrbracket}, X_2^{[e]} \leftarrow \sum_{i=1}^{N} [X_2^{[e]}]^{\llbracket i \rrbracket}, Y_1^{[e]} \leftarrow \sum_{i=1}^{N} [Y_1^{[e]}]^{\llbracket i \rrbracket}, Y_2^{[e]} \leftarrow \sum_{i=1}^{N} [Y_2^{[e]}]^{\llbracket i \rrbracket}$
+				\quad $X_1^{[e]} \leftarrow \sum_{i=1}^{N} [X_1^{[e]}]^{\lir}$. \\
+				\quad $X_2^{[e]} \leftarrow \sum_{i=1}^{N} [X_2^{[e]}]^{\lir}$. \\
+				\quad $Y_1^{[e]} \leftarrow \sum_{i=1}^{N} [Y_1^{[e]}]^{\lir}$. \\
+				\quad $Y_2^{[e]} \leftarrow \sum_{i=1}^{N} [Y_2^{[e]}]^{\lir}$.
 
@@
 				\quad $B^{[e]} \leftarrow X_1(\vec{v} \cdot \vec{y}) + Y_1(\vec{u} \cdot \vec{x}) - X_2(\vec{u} \cdot \vec{y}) - Y_2(\vec{v} \cdot \vec{x})$. \\
-				\quad $\delta_A^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [R_A^{[e]}]^{\llbracket i \rrbracket}$. \\
-				\quad $\delta_B^{[e]} \leftarrow B^{[e]} - \sum_{i=1}^{N} [R_B^{[e]}]^{\llbracket i \rrbracket}$. \\
-				\quad $\vec{\delta}_x^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [\vec{R}_x^{[e]}]^{\llbracket i \rrbracket}$. \\
-				\quad $\vec{\delta}_y^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [\vec{R}_y^{[e]}]^{\llbracket i \rrbracket}$.
+				\quad $\delta_A^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [R_A^{[e]}]^{\lir}$. \\
+				\quad $\delta_B^{[e]} \leftarrow B^{[e]} - \sum_{i=1}^{N} [R_B^{[e]}]^{\lir}$. \\
+				\quad $\vec{\delta}_x^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [\vec{R}_x^{[e]}]^{\lir}$. \\
+				\quad $\vec{\delta}_y^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [\vec{R}_y^{[e]}]^{\lir}$.
 
 				\item Compute random value $t_0^{[e]}$ and share it. \\
-				\quad $[t_0^{[e]}]^{\llbracket i \rrbracket} \leftarrow Hash_0(i || [vec{R}_x^{[e]}]^{\llbracket i \rrbracket}, [\vec{R}_y^{[e]}]^{\llbracket i \rrbracket}, [X_1^{[e]}]^{\llbracket i \rrbracket}, [X_2^{[e]}]^{\llbracket i \rrbracket}, [Y_1^{[e]}]^{\llbracket i \rrbracket}, [Y_2^{[e]}]^{\llbracket i \rrbracket},$ \\ $[R_A^{[e]}]^{\llbracket i \rrbracket}, [R_B^{[e]}]^{\llbracket i \rrbracket} || \vec{\delta}_x^{[e]}, \vec{\delta}_y^{[e]}, \delta_A^{[e]}, \delta_B^{[e]})$.
+				\quad $[t_0^{[e]}]^{\lir} \leftarrow Hash_0(i || [\vec{R}_x^{[e]}]^{\lir}, [\vec{R}_y^{[e]}]^{\lir}, [X_1^{[e]}]^{\lir}, [X_2^{[e]}]^{\lir}, [Y_1^{[e]}]^{\lir}, [Y_2^{[e]}]^{\lir},$ \\ $[R_A^{[e]}]^{\lir}, [R_B^{[e]}]^{\lir} || \vec{\delta}_x^{[e]}, \vec{\delta}_y^{[e]}, \delta_A^{[e]}, \delta_B^{[e]})$.
 \\
-				\quad $t_0^{[e]} \leftarrow \sum_{i=1}^{N} [t_0^{[e]}]^{\llbracket i \rrbracket}$.
+				\quad $t_0^{[e]} \leftarrow \sum_{i=1}^{N} [t_0^{[e]}]^{\lir}$.
 
 				\item Compute the commitments \\
-				\quad $\texttt{com}^{[e]}_i := \text{Hash}_0(\texttt{salt}, e, i, [w^{[e]}_i], [g^{[e]}_i])$ \\
-				for all $i \in [1:N]$, and compute the Merkle root \\
-				\quad $\tilde{h}^{[e]} := \text{MerkleTree}(\texttt{com}^{[e]}_1, \dots, \texttt{com}^{[e]}_N)$.
+				\quad $View_i \leftarrow (
+				[t_0^{[e]}]^{\lir},
+				[(X_1 + t_0(\vec{u} \cdot \vec{x}))^{[e]}]^{\lir},
+				[(X_2 + t_0(\vec{v} \cdot \vec{x}))^{[e]}]^{\lir},
+				[(Y_1 + t_0(\vec{v} \cdot \vec{y}))^{[e]}]^{\lir},
+				[(Y_2 + t_0(\vec{u} \cdot \vec{y}))^{[e]}]^{\lir},
+				[(A + t_0 B)^{[e]}]^{\lir}
+				)$. \\
+				\quad $\texttt{com}^{[e]} := \text{Hash}_0(\texttt{salt}, e, View_1, \dots, View_N, \vec{\delta}_x, \vec{\delta}_y, \delta_A, \delta_B)$
 			\end{enumerate}
@@
 			\begin{enumerate}[label=\arabic*.]
-				\item Compute $h_1 := \text{Hash}(m, \texttt{salt}, \tilde{h}^{[1]}, \dots, \tilde{h}^{[\tau]})$.
+				\item Compute $h_1 := \text{Hash}(m, \texttt{salt}, \texttt{comm}^{[1]}, \dots, \texttt{comm}^{[\tau]})$.
 				\item Expand $h_1$ as $(e^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h_1)$.
@@
 				and recomposes $\alpha^{[e]}$. \\
-
-				\textit{This step is repeated as many times as in the MPC protocol (cf Protocol 11).}
 			\end{enumerate}
@@
 
-	The following proofs have been inspired by the proof of \textbf{EUF-CMA} of a generic 5-round protocols by Thibauld Feneuil \cite{feneu2023}.
+	We rely on the following theorem from Thibault Feneuil's PhD thesis \cite{feneu2023}, which states that any 5-round signing protocol instantiated within the MPC-in-the-Head (MPCitH) framework is \textbf{EUF-CMA} secure, under certain assumptions specified in the theorem.
 

### 2025-06-20 rapport.tex
protocol terminé

@@
 
-				\item Share the witness $w$ into an $(N, N)$-threshold linear secret sharing $[w^{[e]}_i]$: \\
+				\item Share the witness $w$ into an $(N-1)$-private linear secret sharing $[w^{[e]}_i]$: \\
 				\quad $[X_1^{[e]}], [X_2^{[e]}], [Y_1^{[e]}], [Y_2^{[e]}], [R_A^{[e]}], [R_B^{[e]}] \leftarrow \text{Share}(w; r^{[e]}_w)$ \#The shares are in $\mathbb{F}_{q^k}$. \\
@@
 				\quad $Y_2^{[e]} \leftarrow \sum_{i=1}^{N} [Y_2^{[e]}]^{\lir}$.
+			\end{enumerate}
 
+			\vspace{0.5em}
+			\textbf{Phase 2: First challenge (randomness for the MPC protocol).}
+			\begin{enumerate}[label=\arabic*.]
+				\item Expand $h_1$ as $(e^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h_1)$.
+
+				\item Compute random value $t_0^{[e]}$ and share it. \\
+				\quad $[t_0^{[e]}]^{\lir} \leftarrow Hash_0(i || [\vec{R}_x^{[e]}]^{\lir}, [\vec{R}_y^{[e]}]^{\lir}, [X_1^{[e]}]^{\lir}, [X_2^{[e]}]^{\lir}, [Y_1^{[e]}]^{\lir}, [Y_2^{[e]}]^{\lir},$ \\ $[R_A^{[e]}]^{\lir}, [R_B^{[e]}]^{\lir} || \vec{\delta}_x^{[e]}, \vec{\delta}_y^{[e]}, \delta_A^{[e]}, \delta_B^{[e]})$.
+\\
+				\quad $t_0^{[e]} \leftarrow \sum_{i=1}^{N} [t_0^{[e]}]^{\lir}$.
+			\end{enumerate}
+
+			\vspace{0.5em}
+			\textbf{Phase 3: Simulation of the MPC protocol.} For each $e \in [1:\tau]$:
+			\begin{enumerate}[label=\arabic*.]
 				\item Compute $A^{[e]}$ and $B^{[e]}$ and share them: \\
@@
 				\quad $\delta_B^{[e]} \leftarrow B^{[e]} - \sum_{i=1}^{N} [R_B^{[e]}]^{\lir}$. \\
-				\quad $\vec{\delta}_x^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [\vec{R}_x^{[e]}]^{\lir}$. \\
-				\quad $\vec{\delta}_y^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [\vec{R}_y^{[e]}]^{\lir}$.
-
-				\item Compute random value $t_0^{[e]}$ and share it. \\
-				\quad $[t_0^{[e]}]^{\lir} \leftarrow Hash_0(i || [\vec{R}_x^{[e]}]^{\lir}, [\vec{R}_y^{[e]}]^{\lir}, [X_1^{[e]}]^{\lir}, [X_2^{[e]}]^{\lir}, [Y_1^{[e]}]^{\lir}, [Y_2^{[e]}]^{\lir},$ \\ $[R_A^{[e]}]^{\lir}, [R_B^{[e]}]^{\lir} || \vec{\delta}_x^{[e]}, \vec{\delta}_y^{[e]}, \delta_A^{[e]}, \delta_B^{[e]})$.
-\\
-				\quad $t_0^{[e]} \leftarrow \sum_{i=1}^{N} [t_0^{[e]}]^{\lir}$.
+				\quad $\vec{\delta}_x^{[e]} \leftarrow \vec{x}^{[e]} - \sum_{i=1}^{N} [\vec{R}_x^{[e]}]^{\lir}$. \\
+				\quad $\vec{\delta}_y^{[e]} \leftarrow \vec{y}^{[e]} - \sum_{i=1}^{N} [\vec{R}_y^{[e]}]^{\lir}$.
 
@@
 				)$. \\
-				\quad $\texttt{com}^{[e]} := \text{Hash}_0(\texttt{salt}, e, View_1, \dots, View_N, \vec{\delta}_x, \vec{\delta}_y, \delta_A, \delta_B)$
-			\end{enumerate}
+				\quad $\texttt{com}^{[e]} := \text{Hash}_0(e, View_1, \dots, View_N, \vec{\delta}_x, \vec{\delta}_y, \delta_A, \delta_B)$
 
-			\vspace{0.5em}
-			\textbf{Phase 2: First challenge (randomness for the MPC protocol).}
-			\begin{enumerate}[label=\arabic*.]
 				\item Compute $h_1 := \text{Hash}(m, \texttt{salt}, \texttt{comm}^{[1]}, \dots, \texttt{comm}^{[\tau]})$.
-				\item Expand $h_1$ as $(e^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h_1)$.
-			\end{enumerate}
-
-			\vspace{0.5em}
-			\textbf{Phase 3: Simulation of the MPC protocol.} For each $e \in [1:\tau]$:
-			\begin{enumerate}[label=\arabic*.]
-				\item Computes, for $i \in S$, \\
-				\quad $[\alpha^{[e]}_i] := \varphi_{x,e,f}([w^{[e]}_i], [g^{[e]}_i])$ \\
-				and recomposes $\alpha^{[e]}$. \\
 			\end{enumerate}
@@
 				\item Compute $h_2 := \text{Hash}(\texttt{salt}, h_1, [\alpha^{[1]}]_{S}, \dots, [\alpha^{[\tau]}]_{S})$.
-				\item Expand $h_2$ as $(r^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h_2)$.
+				\item Expand $h_2$ as $(I^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h_2)$ where, for every $e, I^{[e]} \in [1:N]$ the party that will not be opened.
 			\end{enumerate}
@@
 
-			\[
-			\sigma := \texttt{salt} \parallel h_1 \parallel h_2 \parallel (([w^{[e]}_i], [g^{[e]}_i])_{i \in r^{[e]}}, \texttt{auth}^{[e]}, \alpha^{[e]})_{e \in [1:\tau]}
-			\]
-
-			where $\texttt{auth}^{[e]}$ is the authentication path for $\{\texttt{com}^{[e]}_i\}_{i \in r^{[e]}}$ w.r.t. Merkle root $\tilde{h}^{[e]}$ and $i^{[e]} \notin r^{[e]}$.
+			$$
+			\sigma := \texttt{salt} \parallel
+			h_1 \parallel
+			punctured\_keys \parallel
+			[\delta_A^{[e]}]_{e\in[1,\tau]} \parallel
+			[\delta_B^{[e]}]_{e\in[1,\tau]} \parallel
+			$$
+			$$
+			[\vec{\delta}_x^{[e]}]_{e\in[1,\tau]} \parallel [\vec{\delta}_y^{[e]}]_{e\in[1,\tau]} \parallel
+			[(X_1 + t_0(\vec{u} \cdot \vec{x}))^{[e]}]_{e\in[1,\tau]} \parallel
+			[(X_2 + t_0(\vec{v} \cdot \vec{x}))^{[e]}]_{e\in[1,\tau]} \parallel
+			$$
+			$$
+			[(Y_1 + t_0(\vec{v} \cdot \vec{y}))^{[e]}]_{e\in[1,\tau]} \parallel
+			[(Y_2 + t_0(\vec{u} \cdot \vec{y}))^{[e]}]_{e\in[1,\tau]}
+			$$
 		\end{tcolorbox}

### 2025-06-20 rapport.tex
doucement mais sûrement

@@
 
+	To lighten the notation, we will not write the index of the round, but all the sharing elements are per round $e \in [1, \tau]$.
+
 		\begin{tcolorbox}[title=Protocol 15: Signing Algorithm, breakable, enhanced, colback=white, colframe=black]\label{Protocol_sig}
@@
 			\begin{enumerate}[label=\arabic*.]				
-				\item Derive randomness $r^{[e]}_{w}, r^{[e]}_r, r^{[e]}_{\psi}$ from the root seed $\rho$: \\
-				\quad $(r^{[e]}_w, r^{[e]}_r, r^{[e]}_{\psi}) \leftarrow \text{PRG}(\texttt{salt}, e, \rho)$.
+				\item Share the witness $(\vec{x},\vec{y})$ into an $(N-1)$-private linear secret sharing $\vec{x}^{\lir}, \vec{y}^{\lir}$: \\
+				\quad $\vec{R}_x^{\lir} = \text{PRG}(\texttt{"$R_x$ i"} \parallel salt \parallel e \parallel \rho)$ \\
+				\quad $\vec{R}_y^{\lir} = \text{PRG}(\texttt{"$R_y$ i}\parallel salt \parallel e \parallel \rho)$ \\
+				\quad $\vec{\delta}_x \leftarrow \vec{x} - \sum_{i=1}^{N} \vec{R}_x^{\lir}$. \\
+				\quad $\vec{\delta}_y \leftarrow \vec{y} - \sum_{i=1}^{N} \vec{R}_y^{\lir}$.
 
-				\item Share the witness $w$ into an $(N-1)$-private linear secret sharing $[w^{[e]}_i]$: \\
-				\quad $[X_1^{[e]}], [X_2^{[e]}], [Y_1^{[e]}], [Y_2^{[e]}], [R_A^{[e]}], [R_B^{[e]}] \leftarrow \text{Share}(w; r^{[e]}_w)$ \#The shares are in $\mathbb{F}_{q^k}$. \\
-				\quad $[\vec{R}_y^{[e]}], [\vec{R}_y^{[e]}] \leftarrow \text{Share}(w; r^{[e]}_w)$ \#The shares are in $(\mathbb{F}_{q})^n$.
+				\item Prepare MPC protocol inputs\\
+				\quad $X_1^{\lir} = \text{PRG}(\texttt{"$X_1$ i}\parallel salt \parallel e \parallel \rho)$ \\
+				\quad $X_2^{\lir} = \text{PRG}(\texttt{"$X_2$ i}\parallel salt \parallel e \parallel \rho)$ \\
+				\quad $Y_1^{\lir} = \text{PRG}(\texttt{"$Y_1$ i}\parallel salt \parallel e \parallel \rho)$ \\
+				\quad $Y_2^{\lir} = \text{PRG}(\texttt{"$Y_2$ i}\parallel salt \parallel e \parallel \rho)$ \\
+				\quad $R_A^{\lir} = \text{PRG}(\texttt{"$R_A$ i}\parallel salt \parallel e \parallel \rho)$ \\
+				\quad $R_B^{\lir} = \text{PRG}(\texttt{"$R_B$ i}\parallel salt \parallel e \parallel \rho)$ \\
 
-				\item Compute $X_1^{[e]}, X_2^{[e]}, Y_1^{[e]}, Y_2^{[e]}$ \\
-				\quad $X_1^{[e]} \leftarrow \sum_{i=1}^{N} [X_1^{[e]}]^{\lir}$. \\
-				\quad $X_2^{[e]} \leftarrow \sum_{i=1}^{N} [X_2^{[e]}]^{\lir}$. \\
-				\quad $Y_1^{[e]} \leftarrow \sum_{i=1}^{N} [Y_1^{[e]}]^{\lir}$. \\
-				\quad $Y_2^{[e]} \leftarrow \sum_{i=1}^{N} [Y_2^{[e]}]^{\lir}$.
+				\item Compute $X_1, X_2, Y_1, Y_2$ \\
+				\quad $X_1 \leftarrow \sum_{i=1}^{N} X_1^{\lir}$. \\
+				\quad $X_2 \leftarrow \sum_{i=1}^{N} X_2^{\lir}$. \\
+				\quad $Y_1 \leftarrow \sum_{i=1}^{N} Y_1^{\lir}$. \\
+				\quad $Y_2 \leftarrow \sum_{i=1}^{N} Y_2^{\lir}$.
 			\end{enumerate}
@@
 			\begin{enumerate}[label=\arabic*.]
-				\item Expand $h_1$ as $(e^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h_1)$.
+				\item Compute random value $t_0^{\lir}$ for each round and each party. \\
+				\quad $t_0^{\lir} \leftarrow Hash_0(
+				i \parallel
+				\vec{R}_x^{\lir},
+				\vec{R}_y^{\lir},
+				R_A^{\lir},
+				R_B^{\lir},
+				X_1^{\lir},
+				X_2^{\lir},
+				Y_1^{\lir},
+				Y_2^{\lir} \parallel
+				\vec{\delta}_x^{\lir},
+				\vec{\delta}_y^{\lir},
+				\delta_A^{\lir},
+				\delta_B^{\lir}
+				)$.
 
-				\item Compute random value $t_0^{[e]}$ and share it. \\
-				\quad $[t_0^{[e]}]^{\lir} \leftarrow Hash_0(i || [\vec{R}_x^{[e]}]^{\lir}, [\vec{R}_y^{[e]}]^{\lir}, [X_1^{[e]}]^{\lir}, [X_2^{[e]}]^{\lir}, [Y_1^{[e]}]^{\lir}, [Y_2^{[e]}]^{\lir},$ \\ $[R_A^{[e]}]^{\lir}, [R_B^{[e]}]^{\lir} || \vec{\delta}_x^{[e]}, \vec{\delta}_y^{[e]}, \delta_A^{[e]}, \delta_B^{[e]})$.
-\\
-				\quad $t_0^{[e]} \leftarrow \sum_{i=1}^{N} [t_0^{[e]}]^{\lir}$.
+				\item Compute $t_0 \leftarrow \sum_{i=1}^{N} t_0^{\lir}$.
 			\end{enumerate}
@@
 			\begin{enumerate}[label=\arabic*.]
-				\item Compute $A^{[e]}$ and $B^{[e]}$ and share them: \\
-				\quad $A^{[e]} \leftarrow X_1Y_1 - X_2Y_2$. \\
-				\quad $B^{[e]} \leftarrow X_1(\vec{v} \cdot \vec{y}) + Y_1(\vec{u} \cdot \vec{x}) - X_2(\vec{u} \cdot \vec{y}) - Y_2(\vec{v} \cdot \vec{x})$. \\
-				\quad $\delta_A^{[e]} \leftarrow A^{[e]} - \sum_{i=1}^{N} [R_A^{[e]}]^{\lir}$. \\
-				\quad $\delta_B^{[e]} \leftarrow B^{[e]} - \sum_{i=1}^{N} [R_B^{[e]}]^{\lir}$. \\
-				\quad $\vec{\delta}_x^{[e]} \leftarrow \vec{x}^{[e]} - \sum_{i=1}^{N} [\vec{R}_x^{[e]}]^{\lir}$. \\
-				\quad $\vec{\delta}_y^{[e]} \leftarrow \vec{y}^{[e]} - \sum_{i=1}^{N} [\vec{R}_y^{[e]}]^{\lir}$.
+				\item Compute $A$ and $B$ and share them: \\
+				\quad $A \leftarrow X_1Y_1 - X_2Y_2$. \\
+				\quad $B \leftarrow X_1(\vec{v} \cdot \vec{y}) + Y_1(\vec{u} \cdot \vec{x}) - X_2(\vec{u} \cdot \vec{y}) - Y_2(\vec{v} \cdot \vec{x})$. \\
+				\quad $\delta_A \leftarrow A - \sum_{i=1}^{N} R_A^{\lir}$. \\
+				\quad $\delta_B \leftarrow B - \sum_{i=1}^{N} R_B^{\lir}$. \\
 
@@
 				\quad $View_i \leftarrow (
-				[t_0^{[e]}]^{\lir},
-				[(X_1 + t_0(\vec{u} \cdot \vec{x}))^{[e]}]^{\lir},
-				[(X_2 + t_0(\vec{v} \cdot \vec{x}))^{[e]}]^{\lir},
-				[(Y_1 + t_0(\vec{v} \cdot \vec{y}))^{[e]}]^{\lir},
-				[(Y_2 + t_0(\vec{u} \cdot \vec{y}))^{[e]}]^{\lir},
-				[(A + t_0 B)^{[e]}]^{\lir}
+				t_0^{\lir},
+				(X_1 + t_0(\vec{u} \cdot \vec{x}))^{\lir},
+				(X_2 + t_0(\vec{v} \cdot \vec{x}))^{\lir},
+				(Y_1 + t_0(\vec{v} \cdot \vec{y}))^{\lir},
+				(Y_2 + t_0(\vec{u} \cdot \vec{y}))^{\lir},
+				(A + t_0 B)^{\lir}
 				)$. \\
 				\quad $\texttt{com}^{[e]} := \text{Hash}_0(e, View_1, \dots, View_N, \vec{\delta}_x, \vec{\delta}_y, \delta_A, \delta_B)$
-
-				\item Compute $h_1 := \text{Hash}(m, \texttt{salt}, \texttt{comm}^{[1]}, \dots, \texttt{comm}^{[\tau]})$.
 			\end{enumerate}
@@
 			\begin{enumerate}[label=\arabic*.]
-				\item Compute $h_2 := \text{Hash}(\texttt{salt}, h_1, [\alpha^{[1]}]_{S}, \dots, [\alpha^{[\tau]}]_{S})$.
+				\item Compute $h_2 := \text{Hash}(m, \texttt{salt}, \texttt{comm}^{[1]}, \dots, \texttt{comm}^{[\tau]})$.
 				\item Expand $h_2$ as $(I^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h_2)$ where, for every $e, I^{[e]} \in [1:N]$ the party that will not be opened.

### 2025-06-20 rapport.tex
manque expand et punctured keys
minor changes

@@
 
-	To lighten the notation, we will not write the index of the round, but all the sharing elements are per round $e \in [1, \tau]$.
+	To simplify the notation, we omit the round index when it is clear from context. Unless stated otherwise, all sharing-related elements are defined per round $e \in [1, \tau]$.
+
+	We denote by $\mathcal{K}^e$ the punctured key corresponding to round $e$. This punctured key represents the co-path in the GGM tree used to derive the associated random values.
+
+	The function Expand is a generic function, that expands the hash to extract different psuedo-random values from it.
 
@@
 				\quad $Y_2 \leftarrow \sum_{i=1}^{N} Y_2^{\lir}$.
+
+				\item Compute $A$ and $B$ and share them: \\
+				\quad $A \leftarrow X_1Y_1 - X_2Y_2$. \\
+				\quad $B \leftarrow X_1(\vec{v} \cdot \vec{y}) + Y_1(\vec{u} \cdot \vec{x}) - X_2(\vec{u} \cdot \vec{y}) - Y_2(\vec{v} \cdot \vec{x})$. \\
+				\quad $\delta_A \leftarrow A - \sum_{i=1}^{N} R_A^{\lir}$. \\
+				\quad $\delta_B \leftarrow B - \sum_{i=1}^{N} R_B^{\lir}$.
 			\end{enumerate}
@@
 			\begin{enumerate}[label=\arabic*.]
-				\item Compute $A$ and $B$ and share them: \\
-				\quad $A \leftarrow X_1Y_1 - X_2Y_2$. \\
-				\quad $B \leftarrow X_1(\vec{v} \cdot \vec{y}) + Y_1(\vec{u} \cdot \vec{x}) - X_2(\vec{u} \cdot \vec{y}) - Y_2(\vec{v} \cdot \vec{x})$. \\
-				\quad $\delta_A \leftarrow A - \sum_{i=1}^{N} R_A^{\lir}$. \\
-				\quad $\delta_B \leftarrow B - \sum_{i=1}^{N} R_B^{\lir}$. \\
-
 				\item Compute the commitments \\

### 2025-06-27 rapport.tex
intro + cGGM

@@
 \usepackage{mathtools}
-\usepackage[backend=bibtex]{biblatex}
+\usepackage[backend=bibtex,style=alphabetic]{biblatex}
 \usepackage{csquotes}
@@
 
-	\begin{abstract}
+	\section{Introduction}
+		Since the emergence of quantum computing in the late 1990s, several breakthroughs have been achieved in the field of quantum algorithms.
+Some of these advancements have found applications in areas such as medicine and physics, while others have demonstrated the ability to break cryptographic problems that are currently assumed to be hard for classical computers, such as RSA and ECDH.
+
+		In particular, quantum algorithms like Grover's \cite{Grover} and Shor's \cite{Shor} have shown that existing public-key cryptosystems are vulnerable in a post-quantum setting.
+This has led to an urgent need for new hard problems on which to base quantum-resistant public-key cryptography.
+
+		Post-quantum cryptographic schemes are generally built upon four main problem classes: isogeny-based, multivariate, code-based, and lattice-based cryptography.
+In recent years, new paradigms have also been explored to improve the efficiency and security of schemes within these classes.
+One such paradigm is the \emph{MPC-in-the-Head} approach.
+
+		Although these underlying problems are believed to be hard even for quantum computers, some have been shown to be vulnerable to classical attacks due to hidden structures.
+A notable example is the NIST finalist Rainbow \cite{JD05}, a multivariate signature scheme that was broken in 2022 by Beullens \cite{Beu22} through exploitation of structural weaknesses.
+
+		One might argue that post-quantum cryptography is not yet necessary, given that large-scale quantum computers capable of breaking classical cryptography do not currently exist.
+However, adversaries could perform so-called \emph{harvest-and-decrypt-later} attacks, in which encrypted data is collected today and decrypted in the future when quantum capabilities become available.
+
+		In response to these concerns, the National Institute of Standards and Technology (NIST) launched an international competition in 2016 to standardize post-quantum cryptographic primitives.
+
 		The \emph{MPC-in-the-Head} (MPCitH) paradigm, first introduced in \cite{IKOS07}, has recently gained significant attention—particularly in the second round of NIST’s post-quantum signature standardization process, where 5 out of 14 candidate schemes were based on this approach.
@@
 
+		The main appeal of the MPCitH paradigm lies in the fact that it enables the construction of efficient, non-interactive, and quantum-resistant signature schemes with relatively small signature sizes and low verification cost and extremely small keys size.
+
 		In~\cite{HJ23}, the authors introduced the \emph{Subfield Bilinear Collision} (SBC) problem, which is rooted in the hardness of the discrete logarithm problem \cite{Joux13}.
@@
 		The objective of this internship is to conduct a cryptanalysis of this scheme, assess its security and efficiency, and explore potential improvements.
-	\end{abstract}
-
-	\section{MPCitH and SBC signature scheme}
+		This includes evaluating its concrete security parameters and identifying potential structural weaknesses.
+
+		In Section~\ref{sec2}, we introduce the MPCitH paradigm and provide an overview of the SBC problem, including its variants as well as the identification and signature schemes built upon it.
+
+		In Section~\ref{sec3}, we present key results in polynomial system solving and describe the main algorithms and results used in our cryptanalysis.
+
+		Finally, in Section~\ref{sec4}, we detail our contributions, including the cryptanalysis and formal security evaluation of the SBC-based signature scheme, and summarize the most efficient known attacks against it.
+
+	\newpage
+	\tableofcontents
+	\newpage
+	\section{MPCitH and SBC signature scheme}\label{sec2}
 
@@
 		Since most of the values used in the MPC protocol are randomly generated, a technique known as the GGM tree can be employed to transmit only a seed, rather than all the individual random values.
+This technique has been introduced in \cite{GGM86}.
 
@@
 As a result, instead of sharing $N$ values, we only need to share $log(N)$ values.
+
+		Usually, the linear private share of a value $x$ is done this way:
+		\begin{enumerate}
+			\item we create $n$ random shares $x^{\lir} \overset{{\scriptscriptstyle\$}}\gets \mathbb{F}_q, \forall i \in [1,N]$.
+			\item we compute the offset $\delta_x = x - \sum_{i=1}^{N} x^{\lir} \text{ mod } \mathbb{F}_q$.
+		\end{enumerate}
+
+		One challenge in this approach is the need to communicate the offset to the verifier so that they can verify the computations.
+However, a recent technique called \emph{Correlated GGM} (cGGM) has been introduced in \cite{GYWZ+22}, which allows the prover to generate random values $x^{\lir}$ using a modified GGM tree, such that $\sum_{i=1}^{N} x^{\lir} = x$, with minimal computational overhead.
 
@@
 
-	\section{Polynomial system solving}
+	\section{Polynomial system solving}\label{sec3}
 		Gröbner basis theory was first introduced by Buchberger in his thesis \cite{Buc} and he gave an algorithm to compute the Gröbner basis of an ideal.
@@
 
-	\section{Contributions}
+	\section{Contributions}\label{sec4}
 
@@
 
-		\begin{tcolorbox}[title=Protocol 15: Signing Algorithm, breakable, enhanced, colback=white, colframe=black]\label{Protocol_sig}
-			\textbf{Inputs:} A public key $x = (\vec{u}, \vec{v})$, a private key $w = (\vec{x}, \vec{y}) \in \text{NSBC}[\vec{u}, \vec{v}]$ and a message $m \in \{0,1\}^*$.
+		\begin{tcolorbox}[title=Protocol 1: Signature scheme - Signing Algorithm, breakable, enhanced, colback=white, colframe=black]\label{Protocol_sig}
+			\textbf{Inputs:} A public key $(\vec{u}, \vec{v})$, a private key $(\vec{x}, \vec{y}) \in \text{NSBC}[\vec{u}, \vec{v}]$ and a message $m \in \{0,1\}^*$.
 
@@
 			\begin{enumerate}[label=\arabic*.]
-				\item Compute $h_2 := \text{Hash}(m, \texttt{salt}, \texttt{comm}^{[1]}, \dots, \texttt{comm}^{[\tau]})$.
-				\item Expand $h_2$ as $(I^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h_2)$ where, for every $e, I^{[e]} \in [1:N]$ the party that will not be opened.
+				\item Compute $h := \text{Hash}(m, \texttt{salt}, \texttt{comm}^{[1]}, \dots, \texttt{comm}^{[\tau]})$.
+				\item Expand $h$ as $(I^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h)$ where, for every $e, I^{[e]} \in [1:N]$ the party that will not be opened.
 			\end{enumerate}
@@
 			\sigma := \texttt{salt} \parallel
-			h_1 \parallel
+			h \parallel
 			punctured\_keys \parallel
@@
 			$$
+		\end{tcolorbox}
+
+		\begin{tcolorbox}[title=Protocol 2: Signature scheme - Verification Algorithm, breakable, enhanced, colback=white, colframe=black]\label{Protocol_verif}
+			\textbf{Inputs:} A public key $(\vec{u}, \vec{v})$, a message $m \in \{0,1\}^*$ and a signature $\sigma$.
+
+			\begin{enumerate} 
+			\item Parse the signature $\sigma$ as
+			$$
+			\texttt{salt} \parallel
+			h \parallel
+			punctured\_keys \parallel
+			[\delta_A^{[e]}]_{e\in[1,\tau]} \parallel
+			[\delta_B^{[e]}]_{e\in[1,\tau]} \parallel
+			$$
+			$$
+			[\vec{\delta}_x^{[e]}]_{e\in[1,\tau]} \parallel [\vec{\delta}_y^{[e]}]_{e\in[1,\tau]} \parallel
+			[(X_1 + t_0(\vec{u} \cdot \vec{x}))^{[e]}]_{e\in[1,\tau]} \parallel
+			[(X_2 + t_0(\vec{v} \cdot \vec{x}))^{[e]}]_{e\in[1,\tau]} \parallel
+			$$
+			$$
+			[(Y_1 + t_0(\vec{v} \cdot \vec{y}))^{[e]}]_{e\in[1,\tau]} \parallel
+			[(Y_2 + t_0(\vec{u} \cdot \vec{y}))^{[e]}]_{e\in[1,\tau]}
+			$$
+
+			\item Expand $h$ as $(I^{[e]})_{e \in [1:\tau]} \leftarrow \text{Expand}(h)$ where, for every $e, I^{[e]} \in [1:N]$ the party that we do not have.
+			\end{enumerate}
 		\end{tcolorbox}
@@
 
+	This type of proof typically proceeds by simulating a game in which an adversary interacts with a signing oracle that responds to arbitrary signing queries.
+The adversary's goal is to produce a valid signature on a message that was not previously queried to the oracle.
+A signature scheme is said to be \textbf{EUF-CMA} (Existential Unforgeability under Chosen Message Attack) if the probability that the adversary succeeds in forging such a signature is at most $2^{-\lambda}$, where $\lambda$ denotes the security parameter.
+This probability is referred to as the adversary's advantage.
+
 	\begin{theorem}[\cite{feneu2023}]\label{theorem_feneu}

### 2025-06-27 rapport.tex
cGGM construction

@@
 		One challenge in this approach is the need to communicate the offset to the verifier so that they can verify the computations.
-However, a recent technique called \emph{Correlated GGM} (cGGM) has been introduced in \cite{GYWZ+22}, which allows the prover to generate random values $x^{\lir}$ using a modified GGM tree, such that $\sum_{i=1}^{N} x^{\lir} = x$, with minimal computational overhead.
+However, a recent technique called \emph{Correlated GGM} (cGGM) has been introduced in \cite{GYWZ+22}, which allows the prover to generate random values $x^{\lir}$ using a modified GGM tree, such that $\sum_{i=1}^{N} x^{\lir} = x$, with no computational.
+		This is achieved by enforcing an invariant over the cGGM tree: at each level, the sum of the nodes remains constant.
+More precisely, given a parent node $x$, the left child is defined as $H(x)$ and the right child as $x - H(x)$, where $H$ is a cryptographic hash function.
+As a result, only approximately half of the tree nodes require hash evaluations, compared to all nodes in the classical GGM construction.
+		This technique not only reduces the computational cost but also eliminates the need to transmit the offset to the verifier, thereby improving the overall communication efficiency.
 

### 2025-06-27 rapport.tex
schema cGGM et GGM etc

@@
 \usepackage{caption}
+
 \usetikzlibrary{arrows.meta, positioning}
@@
 
+		\begin{figure}[H]
+			\begin{center}
+				\begin{tikzpicture}
+					\node {$F(\rho)$} [sibling distance = 3cm]
+					child {node {$F\lceil(F(\rho)\rceil)$} 
+						child {node {$F(\lceil F(\lceil F(\rho)\rceil)\rceil)$}}
+						child {node [yshift = 0.5cm, xshift = -0.8cm]{$F(\lfloor F\lceil(F(\rho)\rceil)\rfloor)$}}}
+					child {node {$F(\lfloor F(\rho)\rfloor)$}
+						child {node {$F(\lceil F(\lfloor F(\rho)\rfloor)\rceil)$}}
+						child {node {$F(\lfloor F(\lfloor F(\rho)\rfloor)\rfloor)$}
+						}
+					};
+				\end{tikzpicture}
+				\caption{GGM tree example to secretly share a value $x$ using a PRF $F$. $\lceil F(\rho)\rceil$ (resp. $\lceil F(\rho)\rceil$) denotes the first (resp. second) half of the output of $F(\rho)$.}
+				\label{fig:GGMtree}
+			\end{center}
+		\end{figure}
+
 		One challenge in this approach is the need to communicate the offset to the verifier so that they can verify the computations.
@@
 		This is achieved by enforcing an invariant over the cGGM tree: at each level, the sum of the nodes remains constant.
-More precisely, given a parent node $x$, the left child is defined as $H(x)$ and the right child as $x - H(x)$, where $H$ is a cryptographic hash function.
-As a result, only approximately half of the tree nodes require hash evaluations, compared to all nodes in the classical GGM construction.
+More precisely, given a parent node $x$, the left child is defined as $F(x)$ and the right child as $x - F(x)$, where $F$ is a cryptographic psuedorandom function.
+As a result, only approximately half of the tree nodes require PRF evaluations, compared to all nodes in the classical GGM construction.
 		This technique not only reduces the computational cost but also eliminates the need to transmit the offset to the verifier, thereby improving the overall communication efficiency.
+
+		\begin{figure}[H]
+		\begin{center}
+			\begin{tikzpicture}
+				\node {$x$} [sibling distance = 3cm]
+				child {node {$F(x)$} 
+					child {node {$F(F(x))$}}
+					child {node [yshift = 0.5cm, xshift = -0.8cm]{$F(x) - F(F(x))$}}}
+				child {node {$x - F(x)$}
+					child {node {$F(x - F(x))$}}
+					child {node {$x - F(x) - F(x -F(x))$}
+					}
+				};
+			\end{tikzpicture}
+			\caption{cGGM tree example. Here, it is easy to verify that the sum at each level is equal to $x$.}
+			\label{fig:cGGMtree}
+		\end{center}
+		\end{figure}
+
+		In practice, the PRF used is AES specially because of it's hardware implementation which enables much faster tree generation than other PRF like SHA256 or other hash functions.
+The security of AES as a cryptographic PRF in a GGM tree generation has been studied in \cite{BCCG+24}.
 

### 2025-06-27 rapport.tex
ajout ptite photo

@@
 \usepackage{caption}
+\usepackage[affil-it]{authblk}
 
@@
 \addbibresource{biblio.bib}
+
+\graphicspath{./}
 
@@
 
-\author{Paul Mekhail\\ \textbf{Advisor:} Sorina Ionica}
+\author{Paul Mekhail
+	\thanks{\texttt{paul.mekhail4@pm.me}}}
+\affil{Laboratoire MIS, Université de Picardie Jules Verne, Amiens, France\\Sorbonne Université, Paris, France}
+
+\author{\textbf{Advisor}: Sorina Ionica
+	\thanks{\texttt{sorina.ionica@u-picardie.fr}}}
+\affil{Laboratoire MIS, Université de Picardie Jules Verne, Amiens, France}
 
@@
 		Finally, in Section~\ref{sec4}, we detail our contributions, including the cryptanalysis and formal security evaluation of the SBC-based signature scheme, and summarize the most efficient known attacks against it.
+
+		\begin{figure}[H]
+			\begin{center}
+				\includegraphics[scale=0.04]{unilasalle.jpg}
+				\caption{UniLaSalle Engineering School, where the MIS offices are located(for now). Observant readers may notice the unusual UFO-like structure situated in front of the building.}
+			\end{center}	
+		\end{figure}
 

### 2025-06-28 rapport.tex
minor changes

@@
 
-\title{Cryptanalysis of SBC based post-quantum signature scheme}
+\title{Cryptanalysis of SBC based post-quantum signature scheme \\\textit{Graduation internship report}}
 
@@
 		Since the emergence of quantum computing in the late 1990s, several breakthroughs have been achieved in the field of quantum algorithms.
-Some of these advancements have found applications in areas such as medicine and physics, while others have demonstrated the ability to break cryptographic problems that are currently assumed to be hard for classical computers, such as RSA and ECDH.
+Some of these advancements have found applications in areas such as medicine and physics, while others have demonstrated the ability to break cryptographic problems that are currently assumed to be hard for classical computers, such as RSA \cite{RSA78} and ECDH.
 
@@
 
+		We can notice that the output of algorithm \ref{alg:reduc} is polynomials of degree $q$.
+Here $q$ is always $n_x + 1$ or $n_y + 1$.
+We deduce that the only non trivial syzygies are of degree $n_x + 1$ or $n_y + 1$.
+\textbf{c'est pas $n_x + 2$ ou $n_y + 2$ du coup ?}
+
 		The following theorem, ensures that there are no reductions to 0 with extended the F5 criterion.

### 2025-06-30 rapport.tex
minor chages

@@
 
-		In practice, the PRF used is AES specially because of it's hardware implementation which enables much faster tree generation than other PRF like SHA256 or other hash functions.
+		In practice, most MPCitH schemes use hash functions with a salt value but the size of the entry must be $2\lambda$ to achieve $2^{\lambda}$ security bits.
+A recent work by Bui et al.
+studied the security of using the AES as the PRF.
+specially because of it's hardware implementation which enables much faster tree generation than other PRF like SHA256 or other hash functions.
 The security of AES as a cryptographic PRF in a GGM tree generation has been studied in \cite{BCCG+24}.
+Thanks to AES hardware implementation like Intel's AES-NI instruction set, schemes can be up to 50x times faster than versions using hash functions like SHA256.
 

### 2025-06-30 rapport.tex
ajout des tailles

@@
 
+		\textbf{Key generation}.
+The private key $(\vec{x}, \vec{y})$ is constructed as such: uniformly at random choose $\vec{x'}, \vec{y'} \in (\mathbb{F}_q)^{n-2}$ and set $\vec{x} = (\vec{x'}, 1, 0)$ and $\vec{y} = (\vec{y'}, 0, 1)$.
+		For the public key, randomly choose $\vec{u} \in (\mathbb{F}_{q^k})^{n}$ and $v_1,\dots,v_{n-1} \in \mathbb{F}_{q^k}$.
+		Finally, set $v_n$ as:
+		$$
+			v_n = \frac{(\vec{u} \cdot \vec{y})(\sum_{i=1}^{n-1}v_ix_i) - (\vec{u} \cdot \vec{x})(\sum_{i=1}^{n-1}v_iy_i)}{y_n(\vec{u} \cdot \vec{x})}.
+		$$
+
 		\subsection{Keys and parameters estimations}
@@
 		The signature and verification protocol use the MPCitH paradigm, they are described in detail in \cite{HJ23} with proof of soundness, correctness and zero-knowledge.
-		The authors estimate that a good parameter for NSBC problem to be computationally secure are $q = 2, n = 130, k = 257$ using complexity analysis of what they think is the best known cryptanalysis of such problem in \cite{FSS11}.
+		The authors estimate that a good parameter for NSBC problem to be computationally secure are $q = 2, n = 130, k = 257$ using complexity analysis of what they think is the best known cryptanalysis of such problem on a classical computer in \cite{FSS11} with $\mathcal{O}(2^{354})$ operations.
+For the quantum cryptanalysis the estimate that the best attack would be using Grover's algorithm \cite{Grover} with $\mathcal{O}(n^{128})$ operations.
 
@@
 		We present the identification protocol, which serves as a basis for constructing the signature scheme using the Fiat-Shamir transformation.
+
+		\subsection{Keys and signature sizes}
+		The size of the private key is just the size of the seed that is used to generate it which is derived from a PRF.
+So we only need to store the seed which is of size $\lambda$.
+So the size of the private key is $16$ bytes in practice.
+For the public key, we saw that all of it is generated randomly using a PRF and $v_n$ is computed.
+The size of $v_n$ is $2\lambda$ so the is size of public key is $3\lambda$ which in practice is $48$ bytes.
+
+		The signature is composed of the following:
+		\begin{itemize}
+			\item One hash value $h$ corresponding to a global commitment to the $\tau$ round protocol of size $2\lambda$ and one salt of size $\lambda$.
+			\item One punctured PRF key for each round.
+			\item The valued $X_1+t_0(\vec{u}\dot\vec{x}), X_2+t_0(\vec{y}\dot\vec{x}) Y_1+t_0(\vec{v}\dot\vec{y}), Y_2+t_0(\vec{u}\dot\vec{y}) \in \mathbb{F}_{q^k}$ for each round.
+		\end{itemize}
+		The total size of the signature in bits is therefore
+		$$
+			\tau \dot (\lambda log_2(N)) + 10\tau \lambda + 3\lambda = \lambda^2 + 10 \tau \lambda + 3 \lambda \text{ bits},
+		$$
+		where we use the fact that $\lambda = \tau log_2(N)$ in the equation.
+		We remind that we do not need to communicate the auxiliary values $\delta_A, \delta_B, \vec{\delta}_x$ and $\vec{\delta}_y$ if we use a correlated GGM tree.
 

### 2025-07-01 rapport.tex
fini partie MPC

@@
 			\item $w^{\lir}$ is the $i$-th share of w.
-			\item $w^[e]$ is a value used in the $e$-th round of the protocol.
+			\item $w^{[e]}$ is a value used in the $e$-th round of the protocol.
 			\item $\delta_x$ is the offset for the additive sharing of $x$ s.t. $x = \delta_x + \sum_i x^{\lir}$
@@
 		One challenge in this approach is the need to communicate the offset to the verifier so that they can verify the computations.
-However, a recent technique called \emph{Correlated GGM} (cGGM) has been introduced in \cite{GYWZ+22}, which allows the prover to generate random values $x^{\lir}$ using a modified GGM tree, such that $\sum_{i=1}^{N} x^{\lir} = x$, with no computational.
+However, a recent technique called \emph{Correlated GGM} (cGGM) has been introduced in \cite{GYWZ+22}, which allows the prover to generate random values $x^{\lir}$ using a modified GGM tree, such that $\sum_{i=1}^{N} x^{\lir} = x$, with no computational overload.
 		This is achieved by enforcing an invariant over the cGGM tree: at each level, the sum of the nodes remains constant.
@@
 
-		\subsection{Multiparty protocol for the SBC problem}
+		\subsection{Multiparty computation protocol for the SBC problem}
+		A Multi-Party Computation (MPC) protocol is an interactive protocol between $N$ parties $\mathcal{P}_1, \dots, \mathcal{P}_N$, who wish to jointly compute a function $f$ over their private inputs $x^{\lir}$.
+At the end of the protocol, each party outputs its own computed value, denoted $f_i(x) = f(x^{\lir})$.
+
+		Several adversarial models exist for MPC protocols.
+In the \emph{semi-honest} (or passive) model, parties are assumed to follow the protocol honestly but may attempt to infer secret information from the messages they receive.
+In contrast, the \emph{malicious} (or active) model allows adversaries to arbitrarily deviate from the protocol, including corrupting other parties, in order to compromise security.
+
+		In the context of the MPC-in-the-Head (MPCitH) paradigm, we consider only the semi-honest model.
+This is because the parties are simulated \emph{in the head} of the prover, and the only external entity is the verifier, who may attempt to extract the witness from the simulated communications between the parties.
+
+		We recall here the definition of a secure MPC protocol in the semi-host threat model.
+
+		\begin{definition}[def 2.4.5 \cite{feneu2023}]
+			Let $N$ and $t$ be integers such that $1 \leq t \leq N$.
+Let $\Pi_f$ be an MPC protocol with $N$ parties $\mathcal{P}_1,\dots,\mathcal{P}_N$, computing a function $f$.
+The protocol $\Pi_f$ is $t$-private in the semi-honest model if for all $I \subset [N]$ such that $|I| \leq t$, there exists a $PPT$ algorithm $\mathcal{S}$ such that $\mathcal{S}(I, \llbracket x \rrbracket_I, f_I(x))$ is perfectly indistinguishable from the joint distribution of the views of the parties in $I$, where $f_I(x) := {f_i(x), \forall x \in I}$.
+		\end{definition}
+
+		Thanks to the linear sharing, we can easily apply the following operations on the shares:
+
+		\begin{itemize}
+			\item \textbf{Addition}: each party $\mathcal{P}_i$ can locally compute $\llbracket x + y \rrbracket$ by adding their shares:
+			$$
+				(x + y)^{\lir} = x^{\lir} + y^{\lir} \text{ } \forall i \in [N].
+			$$
+			We denote this by $\llbracket x + y \rrbracket = \llbracket x \rrbracket + \llbracket y \rrbracket$.
+			\item \textbf{Adding a constant}: we can easily share the constant $\alpha$ among the parties as $\llbracket \alpha \rrbracket = (\alpha, 0, \dots, 0)$.
+			The parties can then compute $\llbracket x + \alpha \rrbracket = \llbracket x \rrbracket + \llbracket \alpha \rrbracket$. We denote this by $\llbracket x \rrbracket + \alpha$.
+			\item \textbf{Multiplication by a constant}: each party $\mathcal{P}_i$ can locally compute $\llbracket \alpha x \rrbracket$ by multiplying their respective share by $\alpha$:
+			$$
+				(\alpha x)^{\lir} = \alpha x^{\lir} \text{ } \forall i \in [N]. 
+			$$
+			We denote this by $\llbracket \alpha x \rrbracket = \alpha \llbracket x \rrbracket$.
+		\end{itemize}
+
+		Finally, we need a function that we will compute using a MPC protocol to prove that the witness $(\vec{x}, \vec{y}) \in \text{NSBC}(\vec{u}, \vec{v})$.
+
+		Let $X_1, X_2, Y_1, Y_2 \in \mathbb{F}_{q^k}$ be random values. Consider the following polynomial in $t$:
+		\begin{align}
+			F(t) &= (X_1 + t(\vec{u} \cdot \vec{x})) (X_2 + t(\vec{v} \cdot \vec{x})) - (Y_1 	+ t(\vec{v} \cdot \vec{y})) (Y_2 + t(\vec{u} \cdot \vec{y})) \notag \\
+			&= X_1Y_1 - X_2Y_2 \notag \\
+			&+[X_1 (\vec{v} \cdot \vec{y}) + Y_1 (\vec{u} \cdot \vec{x}) - X_2 (\vec{u} \cdot \vec{y}) - Y_2 (\vec{v} \cdot \vec{x})]t \notag \\
+			&+[(\vec{u} \cdot \vec{x}) (\vec{v} \cdot \vec{y}) - (\vec{u} \cdot \vec{y}) (\vec{v} \cdot \vec{x})]t^2. \notag
+		\end{align}
+
+		We notice here that the component of degree $2$ is equal to zero and that $deg(F(t)) < 2$ if and only if $(\vec{x}, \vec{y}) \in \text{NSBC}(\vec{u}, \vec{v})$. In that case, we rewrite $F(t)$ as:
+		$$
+			F(t) = A + Bt
+		$$
+		with $A = X_1Y_1 - X_2Y_2$ and $B = [X_1 (\vec{v} \cdot \vec{y}) + Y_1 (\vec{u} \cdot \vec{x}) - X_2 (\vec{u} \cdot \vec{y}) - Y_2 (\vec{v} \cdot \vec{x})]t$.
+
+		Then we compute the values of $A$ and $B$ and share them among the parties and prove that $F(t) = A + Bt$ which means that $(\vec{x}, \vec{y}) \in \text{NSBC}(\vec{u}, \vec{v})$.
+
+		To do so we apply the following protocol:
+		\begin{enumerate}
+			\item The parties runs an auxiliary protocol to compute a random value $t_0 \in \mathbb{F}_{q^k}$.
+			\item The parties evaluate $F(t_0)$.
+			\item The parties evaluate $A + Bt_0$.
+			\item The parties output \textit{accept} if and only if $F(t_0) = A + Bt_0$ and \textit{reject} otherwise.
+		\end{enumerate}
+
+		The evaluations can be efficiently performed by the parties, thanks to the linear sharing of the coefficients $A$, $B$, and the secret $(\vec{x}, \vec{y})$.
+For a detailed description of the underlying computations, we refer the reader to~\cite{HJ23}.
+
+		One can notice that the evaluations may also coincide if $t_0$ is a root of the polynomial
+		$$
+			(A + Bt) - F(t).
+		$$
+
+		A malicious party $\mathcal{P}_i$ could exploit this by waiting to receive all the shares of $\llbracket t_0 \rrbracket$ from the other parties, and then choosing their own share $t_0^{\lir}$ such that
+		$$
+			t_0 = \sum_{j \neq i} t_0^{\llbracket j \rrbracket} + t_0^{\lir}
+		$$
+		is a root of the polynomial.
+This would allow $\mathcal{P}_i$ to bias the evaluation towards a value that satisfies the equation, which could compromise the protocol by causing it to output \textit{accept} even though $(\vec{x}, \vec{y}) \notin \text{NSBC}(\vec{u}, \vec{v})$.
+
+		To mitigate this issue, parties will not be allowed to choose their share of $\llbracket t_0 \rrbracket$ arbitrarily.
+Instead, each share $t_0^{\lir}$ will be deterministically generated as the output of a hash function applied to a fixed, party-specific string concatenated with a common seed.
+Specifically, the share is computed as:
+		$$
+		t_0^{\lir} = \mathcal{H}(\text{"}t_0\text{ value for party }i\text{"} \, \| \, \text{seed}),
+		$$
+		where $\mathcal{H}$ is a cryptographic hash function and $\|$ denotes concatenation.
+This ensures that the shares are unpredictable yet deterministic, thereby preventing any party from biasing the outcome.
+
+		However, since the value of $t_0$ is truly random, there remains a non-zero probability that it is a root of the polynomial.
+This probability is upper-bounded by $\frac{2}{q^k}$, which is negligible for appropriately chosen parameters.
 
@@
 			\item One punctured PRF key for each round.
-			\item The valued $X_1+t_0(\vec{u}\dot\vec{x}), X_2+t_0(\vec{y}\dot\vec{x}) Y_1+t_0(\vec{v}\dot\vec{y}), Y_2+t_0(\vec{u}\dot\vec{y}) \in \mathbb{F}_{q^k}$ for each round.
+			\item The valued $X_1+t_0(\vec{u}\cdot\vec{x}), X_2+t_0(\vec{y}\cdot\vec{x}) Y_1+t_0(\vec{v}\cdot\vec{y}), Y_2+t_0(\vec{u}\cdot\vec{y}) \in \mathbb{F}_{q^k}$ for each round.
 		\end{itemize}
@@
 		$$
-			\tau \dot (\lambda log_2(N)) + 10\tau \lambda + 3\lambda = \lambda^2 + 10 \tau \lambda + 3 \lambda \text{ bits},
+			\tau \lambda log_2(N) + 10\tau \lambda + 3\lambda = \lambda^2 + 10 \tau \lambda + 3 \lambda \text{ bits},
 		$$
@@
 		We remind that we do not need to communicate the auxiliary values $\delta_A, \delta_B, \vec{\delta}_x$ and $\vec{\delta}_y$ if we use a correlated GGM tree.
+
+		\textbf{Remark:} the actual sizes depend heavily on the machine architecture.
+For instance, on 64-bit architectures, the machine word size is 64 bits, meaning that data structures are typically aligned to multiples of 64 bits.
+As a result, for the parameters $k = 257$, $q = 2$, and $n = 130$, the signature size becomes $4023$ bytes instead of the theoretical $3376$ bytes due to alignment and padding overhead.
 

### 2025-07-02 rapport.tex
fin partie sur SBC

@@
 		\subsection{Multiparty computation protocol for the SBC problem}
-		A Multi-Party Computation (MPC) protocol is an interactive protocol between $N$ parties $\mathcal{P}_1, \dots, \mathcal{P}_N$, who wish to jointly compute a function $f$ over their private inputs $x^{\lir}$.
+		A Multi-Party Computation (MPC) protocol is an interactive protocol between $N$ parties $P_1, \dots, P_N$, who wish to jointly compute a function $f$ over their private inputs $x^{\lir}$.
 At the end of the protocol, each party outputs its own computed value, denoted $f_i(x) = f(x^{\lir})$.
@@
 			Let $N$ and $t$ be integers such that $1 \leq t \leq N$.
-Let $\Pi_f$ be an MPC protocol with $N$ parties $\mathcal{P}_1,\dots,\mathcal{P}_N$, computing a function $f$.
-The protocol $\Pi_f$ is $t$-private in the semi-honest model if for all $I \subset [N]$ such that $|I| \leq t$, there exists a $PPT$ algorithm $\mathcal{S}$ such that $\mathcal{S}(I, \llbracket x \rrbracket_I, f_I(x))$ is perfectly indistinguishable from the joint distribution of the views of the parties in $I$, where $f_I(x) := {f_i(x), \forall x \in I}$.
+Let $\Pi_f$ be an MPC protocol with $N$ parties $P_1,\dots,P_N$, computing a function $f$.
+The protocol $\Pi_f$ is $t$-private in the semi-honest model if for all $I \subset [1, N]$ such that $|I| \leq t$, there exists a $PPT$ algorithm $\mathcal{S}$ such that $\mathcal{S}(I, \llbracket x \rrbracket_I, f_I(x))$ is perfectly indistinguishable from the joint distribution of the views of the parties in $I$, where $f_I(x) := {f_i(x), \forall x \in I}$.
 		\end{definition}
@@
 		\begin{itemize}
-			\item \textbf{Addition}: each party $\mathcal{P}_i$ can locally compute $\llbracket x + y \rrbracket$ by adding their shares:
-			$$
-				(x + y)^{\lir} = x^{\lir} + y^{\lir} \text{ } \forall i \in [N].
+			\item \textbf{Addition}: each party $P_i$ can locally compute $\llbracket x + y \rrbracket$ by adding their shares:
+			$$
+				(x + y)^{\lir} = x^{\lir} + y^{\lir} \text{ } \forall i \in [1, N].
 			$$
@@
 			The parties can then compute $\llbracket x + \alpha \rrbracket = \llbracket x \rrbracket + \llbracket \alpha \rrbracket$. We denote this by $\llbracket x \rrbracket + \alpha$.
-			\item \textbf{Multiplication by a constant}: each party $\mathcal{P}_i$ can locally compute $\llbracket \alpha x \rrbracket$ by multiplying their respective share by $\alpha$:
-			$$
-				(\alpha x)^{\lir} = \alpha x^{\lir} \text{ } \forall i \in [N]. 
+			\item \textbf{Multiplication by a constant}: each party $P_i$ can locally compute $\llbracket \alpha x \rrbracket$ by multiplying their respective share by $\alpha$:
+			$$
+				(\alpha x)^{\lir} = \alpha x^{\lir} \text{ } \forall i \in [1, N]. 
 			$$
@@
 
-		A malicious party $\mathcal{P}_i$ could exploit this by waiting to receive all the shares of $\llbracket t_0 \rrbracket$ from the other parties, and then choosing their own share $t_0^{\lir}$ such that
+		A malicious party $P_i$ could exploit this by waiting to receive all the shares of $\llbracket t_0 \rrbracket$ from the other parties, and then choosing their own share $t_0^{\lir}$ such that
 		$$
@@
 		is a root of the polynomial.
-This would allow $\mathcal{P}_i$ to bias the evaluation towards a value that satisfies the equation, which could compromise the protocol by causing it to output \textit{accept} even though $(\vec{x}, \vec{y}) \notin \text{NSBC}(\vec{u}, \vec{v})$.
+This would allow $P_i$ to bias the evaluation towards a value that satisfies the equation, which could compromise the protocol by causing it to output \textit{accept} even though $(\vec{x}, \vec{y}) \notin \text{NSBC}(\vec{u}, \vec{v})$.
 
 		To mitigate this issue, parties will not be allowed to choose their share of $\llbracket t_0 \rrbracket$ arbitrarily.
-Instead, each share $t_0^{\lir}$ will be deterministically generated as the output of a hash function applied to a fixed, party-specific string concatenated with a common seed.
+Instead, each share $t_0^{\lir}$ will be deterministically generated as the output of a hash function applied to a fixed, party-specific string concatenated with a common seed and something inherit to the party like their shares.
 Specifically, the share is computed as:
 		$$
-		t_0^{\lir} = \mathcal{H}(\text{"}t_0\text{ value for party }i\text{"} \, \| \, \text{seed}),
+		t_0^{\lir} = \mathcal{H}(\text{"}t_0\text{ value for party }i\text{"} \, \| \, \text{seed} | x^{\lir}, y^{\lir}, X_1^{\lir}, X_2^{\lir}, Y_1^{\lir}, Y_2^{\lir}, A^{\lir}, B^{\lir}),
 		$$
@@
 		\subsection{Identification Protocol}
-		We present the identification protocol, which serves as a basis for constructing the signature scheme using the Fiat-Shamir transformation.
+		We present the identification protocol, which serves as the foundation for constructing the signature scheme via the Fiat–Shamir transformation.
+
+		To do so, we transform the MPC protocol into an interactive ZKP of knowledge of the witness $(\vec{x}, \vec{y})$ such that it is a solution of $\text{NSBC}[\vec{u}, \vec{v}]$.
+
+		The prover $\mathcal{P}$ simulates the MPC protocol \emph{in their head} and commits to the views of each party $h = \mathcal{H}_V(View_1, \dots, View_N).$ The view of each party consists of all the messages they send and receive during the protocol.
+In addition, $\mathcal{P}$ commits to the following values:
+		\[
+		X_1 + t_0(\vec{u} \cdot \vec{x}), \quad X_1 + t_0(\vec{v} \cdot \vec{x}), \quad Y_1 + t_0(\vec{v} \cdot \vec{y}), \quad Y_2 + t_0(\vec{u} \cdot \vec{y}), \quad A + Bt_0.
+		\]
+
+		Then, the verifier $\mathcal{V}$ selects a random index $i^* \in [1, N]$. The prover $\mathcal{P}$ opens all the shares of the following values for all $i \in [1, N]$ such that $i \neq i^*$:
+		\[
+		x^{\lir}, \quad y^{\lir}, \quad X_1^{\lir}, \quad X_2^{\lir}, \quad Y_1^{\lir}, \quad Y_2^{\lir}, \quad A^{\lir}, \quad B^{\lir}.
+		\]
+
+		Using this information, $\mathcal{V}$ can reconstruct the shares of $t_0^{\lir}$ as previously described. Moreover, they can compute the corresponding shares of the committed values:
+		\[
+		(X_1 + t_0(\vec{u} \cdot \vec{x}))^{\lir}, \quad (X_1 + t_0(\vec{v} \cdot \vec{x}))^{\lir}, \quad (Y_1 + t_0(\vec{v} \cdot \vec{y}))^{\lir}, \quad (Y_2 + t_0(\vec{u} \cdot \vec{y}))^{\lir}, \quad (A + Bt_0)^{\lir}, \quad \forall i \neq i^*.
+		\]
+
+		Since the sharing is linear, the verifier can efficiently recover the missing shares corresponding to $i^*$. For the detailed description of these computations, we refer the reader to~\cite{HJ23}.
+
+		Finally, the verifier $\mathcal{V}$ can compute $h' = \mathcal{H}_V(View_1, \dots, View_N)$ and compare with $h$. If they match, they \textit{accept} and \textit{reject} otherwise.
+
+		We recall that, to reduce the communication cost, we use correlated GGM (cGGM) trees and transmit only the co-paths of the tree instead of all values.
+Additionally, the HyperCube technique is employed to reduce the computational overhead.
+
+		To convert the identification scheme into a signature scheme, we apply the Fiat-Shamir transformation by replacing the verifier's random choice of $i*$ with the output of a hash function.
+
+		The soundness of the protocol is given by $\frac{1}{N} + \frac{1}{q^k}$.
+This is because the underlying MPC protocol is $(N-1)$-private, so the prover $\mathcal{P}$ can potentially cheat on only one view, and the verifier catches them with probability $\frac{1}{N}$.
+Furthermore, we have shown that a false positive may occur when $t_0$ happens to be a root of a certain polynomial, which occurs with probability $\frac{1}{q^k}$.
+
+		Rather than increasing the number of parties $N$, which would increase computational cost, we repeat the protocol $\tau$ times such that:
+
+		$$
+		\left( \frac{1}{N} + \frac{1}{q^k} \right)^{\tau} < 2^{-\lambda},
+		$$
+		where $\lambda$ is the targeted security level.
 
@@
 So we only need to store the seed which is of size $\lambda$.
-So the size of the private key is $16$ bytes in practice.
+So the size of the private key is $16$ bytes in practice with the proposed parameters.
 For the public key, we saw that all of it is generated randomly using a PRF and $v_n$ is computed.
-The size of $v_n$ is $2\lambda$ so the is size of public key is $3\lambda$ which in practice is $48$ bytes.
+The size of $v_n$ is $2\lambda$ so the is size of public key is $3\lambda$ which in practice with the proposed parameters is $48$ bytes.
 
@@
 			\begin{align*}
-				\text{Succ}(\mathcal{A}) = \mathcal{P}\left[
+				\text{Succ}(\mathcal{A}) = \mathbb{P}\left[
 				\begin{array}{l}
@@
 			$$
-				| \mathcal{P}_{seed \in \{0,1\}^n}[b=0 \mid p \leftarrow PRF(seed)] - \mathcal{P}[b=1 \mid p \overset{{\scriptscriptstyle\$}}{\leftarrow} \{0,1\}^{2n}] | \leq \frac{1}{2}.
+				| \mathbb{P}_{seed \in \{0,1\}^n}[b=0 \mid p \leftarrow PRF(seed)] - \mathbb{P}[b=1 \mid p \overset{{\scriptscriptstyle\$}}{\leftarrow} \{0,1\}^{2n}] | \leq \frac{1}{2}.
 			$$

### 2025-07-02 rapport.tex
ajout benchmarks

@@
 As a result, for the parameters $k = 257$, $q = 2$, and $n = 130$, the signature size becomes $4023$ bytes instead of the theoretical $3376$ bytes due to alignment and padding overhead.
+
+		The authors also provide some practical benchmarks to evaluate the performance of their scheme in realistic settings:
+
+		\begin{table}[h!]
+			\begin{center}
+				\begin{tabular}{|c|c|c||c|c||c|c||c|c|}
+					\hline
+					\multicolumn{3}{|c||}{} &  \multicolumn{2}{c||}{SBC $\text{cGGM}_{\text{SHA}}$} & \multicolumn{2}{c||}{SBC $\text{cGGM}_{\text{AES}}$} & \multicolumn{2}{c|}{$\text{SBC}_{\text{FF}}$ $\text{cGGM}_{\text{AES}}$} \\
+					\hline
+					D & $\tau$ & |sign| & Sign & Verify & Sign & Verify & Sign & Verify \\
+					\hline
+					8 & 16 & 5950 B & 9.34 ms & 9.23 ms & 0.95 ms & 0.85 ms & 0.91 ms & 0.81 ms\\
+					\hline
+					9 & 15 & 5822 B & 16.88 ms & 16.75 ms & 1.15 ms & 1.04 ms & 1.06 ms & 0.96 ms \\
+					\hline
+					10 & 13 & 5260 B & 28.62 ms & 28.58 ms & 1.38 ms & 1.29 ms & 1.22 ms & 1.13 ms\\
+					\hline
+					11 & 12 & 5051 B & 52.33 ms & 52.25 ms & 1.95 ms & 1.86 ms & 1.61 ms & 1.51 ms\\
+					\hline
+					12 & 11 & 4810 B & 95.44 ms & 95.15 ms & 3.03 ms & 2.94 ms & 2.31 ms & 2.22 ms\\
+					\hline
+					13 & 10 & 4537 B & 173.44 ms & 173.51 ms & 5.27 ms & 5.18 ms & 3.64 ms & 3.52 ms\\
+					\hline
+					15 & 9 & 4376 B & 628.41 ms & 627.91 ms & 22.77 ms & 22.54 ms & 11.71 ms & 11.46 ms\\
+					\hline
+					16 & 8 & 4023 B & 1116.88 ms & 1117.77 ms & 40.39 ms & 40.09 ms & 20.94 ms & 20.59 ms\\
+					\hline
+				\end{tabular}
+				\caption{Running times for $\lambda = 128$ using $N = 2^D$ parties and $\tau$ rounds with the parameters $q = 2, k= 257$ and $n = 130$ for basic SBC comparing different version of the cGGM tree using an AMD EPYC 8374F processor running at 3.85 GHz.
+$\text{cGGM}_{\text{AES}}$ denotes the usage of AES as a PRF for the cGGM and $\text{cGGM}_{\text{SHA}}$ denotes the usage of the hash function SHA.}
+				\label{SBC_sig_times}
+			\end{center}
+		\end{table}
+
+		We also ran these benchmarks on a consumer grade processor and the results were approximately the double of the ones by the authors of \cite{HJ23}.
+
+		The \texttt{FF} notation refers to an optimization introduced by the authors in~\cite{HJ23}, known as \emph{Fast Fold}, which is applied during the GGM tree generation.
+While we do not describe this optimization in detail here, we refer the interested reader to~\cite{HJ23} for a comprehensive explanation.
 

### 2025-07-04 rapport.tex
meilleure complexité mémoire crossbred

@@
 				\hline
-				Crossbred & D=12, d=10, k=130 & 159 & 90 \\
+				Crossbred & D=12, d=10, k=130 & 159 & 78 \\
 				\hline
@@
 	\begin{proof}
-		We model the $SBC$ signature scheme \cite{HJ23} as a 5-round protocol (see Protocol~\ref{Protocol_sig}).
-By applying Theorem~\ref{theorem_feneu} from \cite{feneu2023}, and under the assumption that the underlying PRG is a $(t_{\mathcal{A}}, \epsilon_{\mathrm{PRG}})$-secure pseudorandom generator, and as established in Lemma~\ref{UB-KOA} that the $SBC$ function is a $(t_{\mathcal{A}}, \epsilon_{\mathrm{OWF}})$-secure one-way function, we conclude that the $SBC$ signature scheme achieves $EUF$-$CMA$ security.
+		We model the SBC signature scheme \cite{HJ23} as a 5-round protocol (see Protocol~\ref{Protocol_sig}).
+By applying Theorem~\ref{theorem_feneu} from \cite{feneu2023}, and under the assumption that the underlying PRG is a $(t_{\mathcal{A}}, \epsilon_{\mathrm{PRG}})$-secure pseudorandom generator, and as established in Lemma~\ref{UB-KOA} that the SBC function is a $(t_{\mathcal{A}}, \epsilon_{\mathrm{OWF}})$-secure one-way function, we conclude that the SBC signature scheme achieves $EUF$-$CMA$ security.
 	\end{proof}

### 2025-07-09 rapport.tex
ajout petit paragraphe bilinearité

@@
 
+		In our setting, we consider bilinear systems that are slightly overdetermined.
+We argue that when using hybrid methods such as \emph{Crossbred}, which avoid the need to construct Macaulay matrices up to high degrees—as required by algorithms like F5—the only reductions to zero that are likely to occur are those explained by the F5 and Frobenius criteria.
+Indeed, other types of reductions to zero typically appear at degrees $n_x + 1$ and $n_y + 1$ for bilinear systems, which are not reached in our practical use cases.
+Consequently, we consider these systems to behave in a semi-regular manner, although we believe they are, in practice, easier to solve.
+
 	\subsection{Admissible parameters for Crossbred}

### 2025-07-23 rapport.tex
correction de Sorina à faire

@@
 	\newpage
-	\section{MPCitH and SBC signature scheme}\label{sec2}
+	\section{MPC-in-the-Head and SBC signature scheme}\label{sec2}
 
@@
 		The \emph{MPC-in-the-Head} (MPCitH) framework, introduced by \cite{IKOS07}, is a technique for constructing zero-knowledge proofs.
-It enables a prover to convince a verifier that they possess a secret witness $w$ for a given public input $x$, without revealing any information about the witness itself.
-
-		Given $\mathcal{L}$ a language in NP (i.e.
+It enables a prover $\mathcal{P}$ to convince a verifier that they possess a secret witness $w$ for a given public input $x$, without revealing any information about the witness itself.
+
+		Let $\mathcal{L}$ be a language in NP (i.e.
 $x \in \mathcal{L} \Longleftrightarrow \exists w, \mathcal{R}(x, w) = 1$).
+$\mathcal{R}$ is a NP-relation.
 Given a public $x$, $\mathcal{P}$ wants to prove that he knows $w$ such that: $\mathcal{R}(x, w)$.
@@
 
-		To represent the secret witness $w$, a linear secret sharing scheme is used among $N$ simulated parties.
-Each party $i$ receives a random share denoted by $[w]^{\lir}$, and the prover computes an additional correction term defined as:
-		$$
-		\delta_w = w - \sum_{i=1}^{N} [w]^{\lir}.
-		$$
-		This ensures that the full witness can be correctly reconstructed from the shares and the correction term during the simulated execution.
-
-		Next, the prover $\mathcal{P}$ commits to the views of the $N$ fictitious parties, denoted as $View_i$, and sends these commitments to the verifier $\mathcal{V}$.
-Each view contains all messages sent and received by the corresponding party during the simulated execution of the MPC protocol.
-
-		The verifier $\mathcal{V}$ then selects a random index $i \in [1, N]$ and sends it to $\mathcal{P}$.
-In response, $\mathcal{P}$ opens the views of all parties except the $i$-th one, i.e., for all $i' \in [1, N]$ such that $i' \neq i$.
-
-		In order for this protocol to satisfy the zero-knowledge property, the underlying MPC protocol must be $(N-1)$-private, meaning that the joint view of any $N-1$ parties reveals no information about the secret witness $w$.
-
 		We introduce some standard notations in MPCitH protocols. 
@@
 			\item $[w]$, $w$ is a shared value.
-			\item $w^{\lir}$ is the $i$-th share of w.
+			\item $w^{\lir}$ is the $i$-th share of $w$.
 			\item $w^{[e]}$ is a value used in the $e$-th round of the protocol.
-			\item $\delta_x$ is the offset for the additive sharing of $x$ s.t. $x = \delta_x + \sum_i x^{\lir}$
+			\item $\delta_w$ is the offset for the additive sharing of $w$ s.t. $w = \delta_w + \sum_i w^{\lir}$
 		\end{itemize}
+
+		To represent the secret witness $w$, a linear secret sharing scheme is used among $N$ simulated parties.
+Each party $i$ receives a random share denoted by $w^{\lir}$, and the prover computes an additional correction term $\delta_w$.
+
+		This ensures that the full witness can be correctly reconstructed from the shares and the correction term during the simulated execution.
+
+		Next, the prover $\mathcal{P}$ commits to the views of the $N$ fictitious parties, denoted as $View_i$, and sends these commitments to the verifier $\mathcal{V}$.
+Each view contains all messages sent and received by the corresponding party during the simulated execution of the MPC protocol.
+
+		The verifier $\mathcal{V}$ then selects a random index $i \in [1, N]$ and sends it to $\mathcal{P}$.
+In response, $\mathcal{P}$ opens the views of all parties except the $i$-th one, i.e., for all $i' \in [1, N]$ such that $i' \neq i$.
+
+		In order for this protocol to satisfy the zero-knowledge property, the underlying MPC protocol must be $(N-1)$-private, meaning that the joint view of any $N-1$ parties reveals no information about the secret witness $w$.
 
@@
 		This is achieved by enforcing an invariant over the cGGM tree: at each level, the sum of the nodes remains constant.
-More precisely, given a parent node $x$, the left child is defined as $F(x)$ and the right child as $x - F(x)$, where $F$ is a cryptographic psuedorandom function.
+More precisely, given a parent node $x$, the left child is defined as $F(x)$ and the right child as $x - F(x)$, where $F$ is a cryptographic PRF.
 As a result, only approximately half of the tree nodes require PRF evaluations, compared to all nodes in the classical GGM construction.
@@
 
-		\subsection{SBC Problem}
+		\subsection{Subfield Bilinear Collision Problem}
 		The problem is the following: let $q$ be a prime power and two positive integers $k$, $n$.
 		\newline
-		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent over $\mathbb{F}_q$, find two non-colinear vectors $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ such that $$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x})$$
+		Given two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$, which are linearly independent over $\mathbb{F}_q$, find two non-colinear vectors $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ such that 
+		$$(\vec{u} \cdot \vec{x})(\vec{v} \cdot \vec{y}) = (\vec{u} \cdot \vec{y})(\vec{v} \cdot \vec{x}),
+		$$
 		where $(\cdot)$ is the dot product between 2 vectors.
+
 		We denote an instance of the SBC problem given by two vectors $\vec{u}, \vec{v} \in (\mathbb{F}_{q^k})^n$ by $\text{SBC}[\vec{u}, \vec{v}]$.
@@
 
-		\textbf{NSBC variant}.
+		\textbf{Normalized Subfield Bilinear Collision variant}.
 		The authors also introduce a particular version of this problem called normalized SBC (NSBC) where $\vec{x}, \vec{y} \in (\mathbb{F}_q)^n$ and
 		$\vec{x} = (\vec{x'}, 1, 0), \vec{y} = (\vec{y'}, 0, 1)$ to ensure that they are not colinear without needing to regenerate each time they are not.
+
 		We also denote $(\vec{x}, \vec{y}) \in \text{NSBC}[\vec{u}, \vec{v}]$ if $(\vec{x}, \vec{y})$ are solutions to the NSBC problem on $(\vec{u}, \vec{v})$.
-
 		\textbf{Key generation}.
-The private key $(\vec{x}, \vec{y})$ is constructed as such: uniformly at random choose $\vec{x'}, \vec{y'} \in (\mathbb{F}_q)^{n-2}$ and set $\vec{x} = (\vec{x'}, 1, 0)$ and $\vec{y} = (\vec{y'}, 0, 1)$.
+The private key $(\vec{x}, \vec{y})$ is constructed as follows: uniformly at random choose $\vec{x'}, \vec{y'} \in (\mathbb{F}_q)^{n-2}$ and set $\vec{x} = (\vec{x'}, 1, 0)$ and $\vec{y} = (\vec{y'}, 0, 1)$.
 		For the public key, randomly choose $\vec{u} \in (\mathbb{F}_{q^k})^{n}$ and $v_1,\dots,v_{n-1} \in \mathbb{F}_{q^k}$.
@@
 		In our setting, we consider bilinear systems that are slightly overdetermined.
-We argue that when using hybrid methods such as \emph{Crossbred}, which avoid the need to construct Macaulay matrices up to high degrees—as required by algorithms like F5—the only reductions to zero that are likely to occur are those explained by the F5 and Frobenius criteria.
+We argue that when using hybrid methods such as \emph{Crossbred}, which avoid the need to construct Macaulay matrices up to high degrees—as required by algorithms like F5—the only reductions to zero that are likely to occur are those explained by the F5 and Frobenius criterion.
 Indeed, other types of reductions to zero typically appear at degrees $n_x + 1$ and $n_y + 1$ for bilinear systems, which are not reached in our practical use cases.

### 2025-07-23 rapport.tex
better biblio

@@
 \usepackage{mathtools}
-\usepackage[backend=bibtex,style=alphabetic]{biblatex}
+\usepackage[backend=bibtex,style=alphabetic,maxbibnames=9,hyperref=true,abbreviate=false,backref=true]{biblatex}
 \usepackage{csquotes}
@@
 
-		In \cite{Lazard83}, Lazard introduced the connection between linear algebra and Gröbner basis theory via the usage of Macaulay matrices \cite{Mac} which is defined as follows.
+		In \cite{Lazard83}, Lazard introduced the connection between linear algebra and Gröbner basis theory via the usage of Macaulay matrices which is defined as follows.
 
@@
 What is remarkable is that the complexity does not implicate the number of polynomials in the system, $m$.
-There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, \textit{libFES-lite} \cite{Bouilla22} and there have been implementations on FPGA that broke real world cryptography \cite{DDVY20}..
+There is also a software made by one of the authors, Charles Bouillaguet, that is available online and easy to use, \textit{libFES-lite} \cite{Bouilla22} and there have been implementations on FPGA that broke real world cryptography \cite{DDVY20}.
 
@@
 
-		The most straightforward cryptanalysis of this type of systems is to perform a Gröbner basis algorithm like Buchberger's algorithm \cite{Buc}, Faugère's F4 \cite{F99}, F5 \cite{F02} or their multiple variants \cite{EF19}.
+		The most straightforward cryptanalysis of this type of systems is to perform a Gröbner basis algorithm like Buchberger's algorithm \cite{Buc}, Faugère's F4 \cite{F99}, F5 \cite{F02} or their multiple variants \cite{EF17}.
 

### 2025-07-23 rapport.tex
page de garde

@@
 
-\author{Paul Mekhail
-	\thanks{\texttt{paul.mekhail4@pm.me}}}
-\affil{Laboratoire MIS, Université de Picardie Jules Verne, Amiens, France\\Sorbonne Université, Paris, France}
-
-\author{\textbf{Advisor}: Sorina Ionica
-	\thanks{\texttt{sorina.ionica@u-picardie.fr}}}
-\affil{Laboratoire MIS, Université de Picardie Jules Verne, Amiens, France}
-
-\title{Cryptanalysis of SBC based post-quantum signature scheme \\\textit{Graduation internship report}}
+\newcommand{\maketitlepage}{
+	\begin{titlepage}
+		\centering
+		% Logos
+		\includegraphics[width=0.3\textwidth]{LOGO_SU_HORIZ_SEUL_CMJN.png} \hfill
+		\includegraphics[width=0.2\textwidth]{Logo-UPJV-bleu.png} \\[1cm]
+
+		% Noms des établissements
+		{\large \textbf{Sorbonne Université}} \\ 
+		{\large \textbf{Université de Picardie Jules Verne}} \\[0.5cm]
+
+		% Nom du Master
+		{\Large Master of Computer Science} \\
+		{\large Cryptology, High-performance Computing, and Algorithms} \\[1.5cm]
+
+		% Titre entouré de lignes
+		\noindent\rule{\textwidth}{1pt} \\[0.4cm]
+		{\Huge \textbf{Cryptanalysis of SBC-based Post-Quantum Signature Scheme}} \\[0.4cm]
+		\noindent\rule{\textwidth}{1pt} \\[1.5cm]
+
+		% Sous-titre
+		{\Large Graduation Internship Report} \\
+		{\large 2025} \\[2cm]
+
+		% En bas de page, deux colonnes : auteur à gauche, encadrants à droite
+		\vfill
+		\begin{minipage}[t]{0.45\textwidth}
+			\raggedright
+			\textbf{Author:} \\
+			Paul Mekhail \\
+			Sorbonne Université, Université de Picardie Jules Verne \\
+			\texttt{paul.mekhail4@pm.me}
+		\end{minipage}
+		\hfill
+		\begin{minipage}[t]{0.45\textwidth}
+			\raggedleft
+			\textbf{Supervisor:} \\
+			Sorina Ionica \\
+			Laboratoire MIS, Université de Picardie Jules Verne \\
+			\texttt{sorina.ionica@u-picardie.fr} \\[0.5cm]
+			\textbf{Academic Supervisor:} \\
+			Mohab Safey El Din \\
+			LIP6, Sorbonne Université \\
+			\texttt{Mohab.Safey@lip6.fr}
+		\end{minipage}
+	\end{titlepage}
+}
 
 \begin{document}
-	\maketitle
+	\maketitlepage
+
+	\newpage
+	\tableofcontents
+	\newpage
 
@@
 
-		\begin{figure}[H]
-			\begin{center}
-				\includegraphics[scale=0.04]{unilasalle.jpg}
-				\caption{UniLaSalle Engineering School, where the MIS offices are located(for now). Observant readers may notice the unusual UFO-like structure situated in front of the building.}
-			\end{center}	
-		\end{figure}
-
-	\newpage
-	\tableofcontents
-	\newpage
 	\section{MPC-in-the-Head and SBC signature scheme}\label{sec2}
